---
title: "Collating a measles vaccination coverage dataset"
author:
  - name: Rachael Milwid
    email: rachael.milwid@phac-aspc.gc.ca
  - name: Irena Papst
    email: irena.papst@phac-aspc.gc.ca
  - name: "<br>"
  - name: Modelling Hub Division
  - name: Applied Public Health Sciences Directorate
  - name: Science and Policy Integration Branch
  - name: Public Health Agency of Canada
execute:
  echo: false
format: 
  html:
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| output: false

# packages
library(readr)
library(readxl)

library(dplyr)
library(stringr)
library(tidyr)
library(scales)

library(ggplot2)
# common plot theme
theme_set(theme_bw())
th <- theme(legend.position = "bottom")

# Source functions
invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))
```

Our ultimate goal is to compile a data set of measles vaccine coverage
by PT, age, and number of doses in order to implement a more realistic
immunity landscape within Modelling Hub models (*e.g.*, the Agent Based
Model).

We start by collecting the available vaccine coverage data, all of which
was collected for a purpose other than disease modelling (monitoring of
public health milestones, school administrative data, etc). Any
transformations and modelling of data required to use it to inform
models (*e.g.*, age estimates to present day, infer estimates for
missing ages) is done downstream, within the context of each modelling
project, as the most appropriate methodology would likely be model- and
modelling-problem-dependent.

In this report, we:

-   document a [standardized data format](#standardized-data-format) for
    all of the datasets we compile
-   give an overview of the [available national data
    sources](#available-national-data-sources), which are used as a
    fall-back when province-specific data is not readily available
-   show the estimates we've compiled from [provincial
    sources](#provincial) and from [national sources](#cNICS)
-   [summarize](#summary-across-pts) the data across PTs

# Standardized data format {#standardized-data-format}

This data will be standardized into a table with the following columns:

```{r}
#| label: coverage_dict

data_dict <- tibble::tribble(
  ~variable, ~description, ~type,
  "pt", "standard two-letter English abbreviation for province or territory name", "character",
  "location", "name of the location (could be province, public health unit/health authority, or more granular)", "character",
  "year_report", "year for which the vaccine coverage was reported", "numeric",
  "year_birth", "birth year", "numeric",
  "age", "age when coverage was estimated", "character",
  "n_doses", "number of doses used to determine vaccine coverage", "character",
  "value", "vaccine coverage as a proportion", "numeric",
  "source", "data source, be it the province/territory (PT), cNICS, or ANICS)", "character"
)

DT::datatable(data_dict)
```

Either (1) `age` and `year_report` or (2) `year_birth` must be specified
for each coverage estimate.

# Available national data sources {#available-national-data-sources}

## Adult National Immunization Coverage Survey

For adults, there have been a few rounds of the adult coverage survey
([aNICS](https://www.canada.ca/en/public-health/services/immunization-vaccines/vaccination-coverage/adult-national-immunization-coverage-survey-2023-results.html)),
which has some information on 1-dose coverage in people aged 18+ years.
Note that these data are self-reported.

We used the latest version, which is from 2023.

> The 2023 adult National Immunization Coverage Survey (aNICS) asked
> respondents if they have ever been vaccinated for 11 routine vaccines,
> though number of doses received was not available.

> In Canada, 87% of adults have received at least one dose of measles
> vaccine

> Measles-containing vaccine is recommended for susceptible adults born
> in or after 1970. In Canada, adults born prior to 1970 are generally
> presumed to have acquired natural immunity to measles; however, some
> of these individuals may be susceptible. Health care workers,
> travellers to destinations outside of Canada, and military personnel
> should receive MMR vaccine, regardless of year of birth. The aNICS
> collected measles immunization coverage for all respondents regardless
> of age. Some results may be underestimated.

> Immunization coverage for hepatitis B, varicella, HPV, pneumococcal,
> shingles, polio, and meningococcal vaccines was higher among adults
> living in urban settings. Tetanus immunization coverage was higher
> among adults living in rural settings, while **no major differences
> were reported for pertussis, measles, and meningococcal vaccines**.

> Immunization coverage varies by ethnicity and was highest among
> \[...\] Indigenous populations for measles vaccine

```{r}
#| label: vax_aNICS

vax_aNICS <- (read_excel(here::here("data", "measles_raw", "vax_canada_adults_2023.xlsx")) 
 |> filter(Vaccine == "Measles")
 |> mutate(across(everything(), as.character))
 |> select(-Vaccine)
 |> pivot_longer(everything(), names_to = "pt")
 |> mutate(
   pt = case_when(
     pt == "Canada" ~ "CA", 
     pt == "YK" ~ "YT", # recode to standardized abbreviation 
     T ~ pt
    ),
   value = as.numeric(value)
   )
 |> left_join(pt_lookup(), by = join_by(pt))
 |> transmute(
   pt,
   location,
   year_report = 2023,
   age = "18+",
   n_doses = "1+",
   value = as.numeric(value),
   source = "aNICS"
 )
)

DT::datatable(vax_aNICS)
```

# PTs for which provincial data sources are used {#provincial}

## Newfoundland and Labrador

-   a CBC [news
    article](https://www.cbc.ca/player/play/video/1.7112669), in which
    Newfoundland and Labrador's chief medical officer of health, Janice
    Fitzgerald, was interviewed, noted a 95% population-level coverage.
    Note that this estimate is provided for reference, and is not
    necessarily used in the analyses below.\
-   A [government
    report](https://www.gov.nl.ca/releases/2024/health/0314n01/) from
    March 14, 2024 states that:

> The province benefits from one of the highest vaccination rates in
> Canada. In 2022, 94.7 per cent of two-year-olds and 92.6 per cent of
> kindergarteners in Newfoundland and Labrador were fully vaccinated
> with two doses of measles-containing vaccines.

```{r}
#| label: vax_NL

vax_NL <- tibble::tibble(
  pt = "NL", 
  location = "Newfoundland and Labrador",
  year_report = 2022,
  age = c("2", "4"),
  n_doses = rep("2", 2), 
  value = c(0.947, 0.926),
  source = "PT"
)

DT::datatable(vax_NL)
```

```{r}
#| label: NL_plot
#| fig-height: 3

plot_coverage(vax_NL)
```

## Prince Edward Island

The following sources present immunization coverage prior to Grade 1
entry by birth cohort. Note that while there is overlapping data within
the different reports, each report has at least 1 additional birth
cohort compared to the others.

-   [2022-2023 school
    year](https://www.princeedwardisland.ca/sites/default/files/publications/childhood_immunization_year_2022_2023_school_age.pdf)
-   [2023-2024 school
    year](https://www.princeedwardisland.ca/sites/default/files/publications/school_year_2023_2024_immunization_rates_prior_to_grade_1_entry.pdf)
-   [2009
    cohort](https://canvax.ca/sites/default/files/2018-07/PEI_ChildhoodImmunizationInPEI_2017.pdf)

```{r}
#| label: vax_PE

vax_PE <- tibble::tibble(
  pt = "PE",
  location = "Prince Edward Island",
  year_report = c(2024, 2023, 2023, 2023, 2023, 2023, 2017),
  year_birth = c(2017, 2016, 2015, 2014, 2013, 2012, 2009),
  n_doses = "2",
  value = c(0.915,0.893, 0.922, 0.941, 0.93, 0.927, 0.958),
  source = "PT"
)

DT::datatable(vax_PE |> arrange(year_birth))
```

```{r}
#| label: PE_plot
#| fig-height: 3

plot_coverage(vax_PE, x = "year_birth")
```

## Quebec

### Government of Quebec

The [MSSS (Ministère de la Santé et des Services
sociaux)](https://msss.gouv.qc.ca/professionnels/vaccination/piq-impacts-des-programmes-de-vaccination/donnees-de-couverture-vaccinale-au-quebec/),
a government ministry of Quebec, released a report in 2024 documenting
vaccine coverage. In the report, they estimate that in 2021-2022, 88% of
primary school children (6-11 years old) were vaccinated and 92% of the
secondary school children (12-16 years old) were vaccinated[^1]. It is
assumed that these statistics represent the full series (2 doses).

[^1]: School ages taken from [school attendance rules for
    QC](https://www.quebec.ca/en/education/study-quebec/education-system).

### Flash Vigie

The Government of Quebec put out a report, [Flash
vigie](https://publications.msss.gouv.qc.ca/msss/fichiers/flashvigie/FlashVigie_vol19_no2.pdf),
in 2025 detailing measles in QC in 2024. Page 2 contains figures
detailing the regional vaccine uptake. Note that this data has *not*
been incorporated into the dataset below.

### INSPQ

INSPQ puts out a regular biennial report of vaccine coverage for 1 and 2
year olds (7 year olds were added as a cohort in the---2019---report),
so we may be able to leverage older reports to get approximations for
older cohorts, *e.g.*, by using the 2 year old coverage level as the
most recent estimate.

The [2021
report](https://www.inspq.qc.ca/sites/default/files/publications/2776-couverture-vaccinale-enfants-quebecois.pdf)
presents data collected in 2012-2019. The [2024
report](https://www.inspq.qc.ca/sites/default/files/publications/3570-etude-couverture-vaccinale-enfants-quebecois-1-2-7-ans-2021.pdf)
contains additional data collected in 2021.

This study looks at vaccine coverage for 1, 2, and 7 years olds over
time. It looks like it's mostly for all of QC, though there are two
tables for two regions: Capitale-Nationale (Table 17A), Laval (Table
17B).

> En 2019, la proportion des enfants qui ont reçu tous les vaccins du
> programme régulier est de:
>
> -   85 % pour la cohorte 1 an (vaccins reçu avant l'âge de 15 mois);
> -   81 % pour la cohorte 2 ans (vaccins reçu avant l'âge de 24 mois)

The above is a lower bound on the MMR coverage because it's % coverage
of the *complete* childhood series (so if a child has MMR but is missing
something else routine, they wouldn't be counted in this coverage.

> La couverture vaccinale par antigène demeure très élevée pour chacune
> des cohortes. La couverture complète pour le DCaT-VPI-Hib avant 15
> mois est de 96 %, et de 92 % avant 24 mois. Pour le pneumocoque, les
> proportions sont de 94 % et 97 % pour la cohorte 1 an et 2 ans
> respectivement. **Pour le RRO, elles sont de 95 % et 92 %.**

In Annex E, vax coverage for measles by cohort is defined as:

-   1 year old cohort: one dose \>= 12 months
-   2 year old cohort: 2 doses (1re dose âge min. d'adm. = 12 mois,
    Interv. min. de 4 sem. entre ces deux doses)
-   7 year old cohort: 2 doses (1re dose âge min. d'adm. = 12 mois,
    Interv. min. de 4 sem. entre ces deux doses)

RRO = MMR, so 95% of the one-year-old cohort had received one dose and
92% of the two-year-old cohort had received two doses

Table 16: Couverture vaccinale complète par antigène à 15 mois selon le
nombre de doses, de 2012 à 2019, cohorte 1 an

-   Rougeole (1 dose et +), 2019 - Toutes doses %: 97.6 - CV à 15 mois
    %: 94.9
-   We are assuming couverture vaccinale refers to the proportion of
    childeren at the target age who received the recomended number of
    doses, while toutes doses includes children who were vaccinated
    early. For the purposes of this report, we will use the couverture
    vaccinale measurement.

```{r}
#| label: vax_QC

# Coverage at 15months with 1+ doses (Tableau 16) (How to interpret the dose number?)

vax_QC1 <- tibble::tibble(
  pt = "QC",
  location = "Quebec",
  year_report  = c(2012, 2014, 2016, 2019, 2021),
  age = "1.25", # 15 months
  n_doses = "1+", 
  value =  c(0.909, 0.932, 0.927, 0.949, 0.922),
  source = "PT"
)


# Coverage at 24 months with 1+ doses and 2+ doses (Tableau 17) (How to interpret the dose number?)

vax_QC2 <- tibble::tibble(
  pt = "QC", 
  location = "Quebec", 
  year_report  = rep(c(2012, 2014, 2016, 2019, 2021),2),
  age = "2",
  n_doses = c(rep("1+",5), rep("2+", 5)),
  value =  c(0.958, 0.967, 0.963, 0.971, 0.95, 0.839, 0.886, 0.905, 0.919, 0.886),
  source = "PT"
)

# this incorporates data from the MSSS

vax_QC3 <- data.frame(
   pt = "QC", 
  location = "Quebec", 
  year_report  = rep(c(2021),2),
  age = c("6-11", "12-16"),
  n_doses = "2+",
  value =  c(0.88, 0.92),
  source = "PT"
)

vax_QC <- rbind(vax_QC1, vax_QC2, vax_QC3)

DT::datatable(vax_QC)
```

```{r}
#| label: QC_plot
#| fig-height: 3

plot_coverage(vax_QC)
```

## Ontario

Three files are available from Public Health Ontario in their
[Immunization Data
Tool](https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Immunization-Tool):

1.  Immunization coverage for age milestones (coverage by year, PHU, age
    milestone)
2.  Immunization coverage by age (coverage by year, PHU, and age)
3.  Immunization coverage by dose (by year, PHU, and number of doses)

"Immunization coverage" for measles is defined in the Overview tab of
the Immunization Data Tool as:

> the proportion of a population that is appropriately immunized against
> a vaccine preventable disease (VPD) at a specific point in time

[This
doc](https://www.publichealthontario.ca/-/media/Data-Files/immunization-tool-technical-notes.pdf?sc_lang=en&rev=e2858a935e374fdaada91daff5246ee6&hash=0860C5FE20F6FF4F8A310BC953364110),
outlining the methods used for the Immunization Data Tool, defines
"up-to-date coverage" for measles as:

> Must satisfy one of the following criteria: - \>=2 valid doses - Have
> a documented exemption for evidence of immunity

Thus, the immunization coverage by age milestones and by age files
effectively refer to 2-dose vaccine coverage.

### Coverage by age

This analysis was conducted using the second data source above:
"Immunization coverage by age".

```{r}
#| label: vax_ON
vax_ON <- (here::here("data", "measles_raw", "vax_ontario.csv")
  |> read_csv(show_col_types = FALSE)
  |> filter(Antigen == "Measles") 
  |> arrange(Public_health_unit, Age)
  |> transmute(
    pt = "ON",
    location = Public_health_unit,
    year_report = as.numeric(str_extract(Year, "^[[:digit:]]+")),
    age = as.character(Age),
    n_doses = "2",
    value = Coverage/100,
    source = "PT"
  )
)

DT::datatable(vax_ON)
```

```{r}
#| label: ON_prov_df
(vax_ON |> dplyr::filter(location=="Ontario") |> DT::datatable())
```

```{r}
#| label: ON_plot
#| fig-height: 3
plot_coverage(vax_ON |> filter(location=="Ontario"), x = "age")
```

```{r}
#| label: ON_PHU_plot
#| fig-height: 12
#| fig-width: 12
plot_coverage(vax_ON |> filter(location!="Ontario"), ncol = 4) + th
```

### Coverage by dose

This analysis was conducted using the third data source above:
"Immunization coverage by dose".

```{r}
#| label: vax_ON_dose

vax_ON_dose <- (here::here("data", "measles_raw", "vax_ontario_dose.csv")
  |> read_csv(show_col_types = FALSE)
  |> filter(Antigen == "Measles")
  |> transmute(
    pt = "ON",
    location = Public_health_unit,
    year_report = as.numeric(str_extract(Year, "^[[:digit:]]+")), # use first part of year because this is based on school year
    age = as.character(Age),
    n_doses = str_extract(Number_of_Doses, "^[[:digit:]]{1}"),
    value = Proportion/100
  )
  |> mutate(n_doses = if_else(n_doses == "2", "2+", n_doses)) # clarify that '2' doses is actually two or more
  |> arrange(location, age)
  |> dplyr::mutate(source = "PT")
)

DT::datatable(vax_ON_dose)
```

```{r}
#| label: ON_dose_plot
#| fig-height: 3
plot_coverage(vax_ON_dose |> filter(location=="Ontario"))
```

```{r}
#| label: ON_PHU_dose_plot
#| fig-height: 12
#| fig-width: 12
plot_coverage(vax_ON_dose |> filter(location!="Ontario"), ncol = 5) + th
```

## Manitoba

[Manitoba
Health](https://www.gov.mb.ca/health/publichealth/surveillance/immunization/index.html)
published immunization coverage estimates:

> Public Health Information Management System (PHIMS) Immunization
> records for reports from 2015 were extracted from the provincial
> immunization registry contained in the Public Health Information
> Management System, PHIMS.
>
> Below are four downloadable excel file files presenting the
> immunization coverage rates (on December 31 of the reporting year) by
> antigen and the number of immunizations delivered by providers over
> the calendar year in Manitoba and five regions.
>
> As new vaccines become available the numbers of doses required may
> change; therefore the coverage rates may not be comparable over time.
>
> -   Immunization coverage in children - the percentage of vaccinated
>     children that received all recommended doses for an antigen at a
>     specific age.
> -   Immunization coverage in children by continuous/non-continuous
>     status - the percentage of vaccinated children that received all
>     recommended doses for an antigen at a specific age for those who
>     have been continuously registered with Manitoba Health since two
>     months of age or younger (continuous) compared to those who were
>     born in Manitoba, but moved out for a period of time and then
>     returned or who moved to Manitoba at an age older than two months
>     (non-continuous).
> -   Immunization coverage in adults and seniors - the percentage of
>     vaccinated adults that received all recommended doses for an
>     antigen at a specific age.
> -   Immunization providers - the number of doses administered by
>     different types of providers between January 1 and December 31 in
>     the reporting year.

Note that the third file (coverage in adults and seniors) only includes
immunizations that occur in adulthood, so no data on measles.

We use "Immunization coverage in children" here. The most recent edition
is **from 2023**. Vaccine coverage is defined as "the percentage of
vaccinated children that received **all recommended doses** for an
antigen at a specific age". Digging into the
[methods](https://www.gov.mb.ca/health/publichealth/surveillance/immunization/docs/annual_immunization_methods_2023.pdf):

> The number of required doses for each antigen to be complete at a
> specific age was based on Manitoba's Routine Immunization Schedules in
> the reporting year and also in previous years. The routine
> immunization schedules for the reporting year and the number of doses
> required to be complete for each age in the reporting year are
> provided in the Appendix.

The immunization schedule is attached to the table below to determine
the number of doses each coverage estimate represents.

```{r}
#| label: vax_MB

# helper to parse MB's highly structured sheets
parse_sheet <- function(sheet_name){
  skip <- ifelse(sheet_name %in% c("age13", "age17"), 6, 4)
  df <- (
  suppressMessages(read_excel(file_MB, sheet = sheet_name, skip = skip, col_names = FALSE))
  |> select(where(is.character))
)
  
  age <- gsub("age0?", "", sheet_name)
  names(df) <- c("vaccine", locations)
  
  (df
    |> mutate(across(-vaccine, \(x) {
      (gsub("%", "", x)
      |> as.numeric()
      |> (\(y) y/100)()
      )
    }))
    |> pivot_longer(-vaccine, names_to = "location")
    |> mutate(age = {{age}})
    |> relocate(age, .before = value)
  )
}

# get data
file_MB <- here::here("data", "measles_raw", "vax_manitoba_child_coverage_2023.xlsx")
sheets_MB <- readxl::excel_sheets(file_MB)

locations <- ( # the same for all sheets
  suppressMessages(read_excel(file_MB, sheet = sheets_MB[1], skip = 1, n_max = 1, col_names = FALSE))
  |> select(where(is.character))
  |> pivot_longer(everything())
  |> pull(value)
)

schedule <- readxl::read_excel(
  here::here("data", "measles_raw", "MB_immunization_schedules_2023.xlsx"),
  sheet = "childhood_doses",
  skip = 1
) |> transmute(age = as.character(Age), n_doses = as.character(Measles)) |> filter(age != 1)

vax_MB <- lapply(sheets_MB, parse_sheet) |> bind_rows() |> filter(vaccine=="Measles") |> transmute(
  pt = "MB",
  location, 
  year_report = 2023,
  age, value
) |> left_join(schedule, by = join_by(age)) |> relocate(n_doses, .before = value) |> 
  dplyr::mutate(source = "PT")

DT::datatable(vax_MB)
```

```{r}
#| label: MB_plot
#| fig-height: 3
plot_coverage(vax_MB |> filter(location=="Manitoba"))
```

```{r}
#| label: MB_PHU_plot
plot_coverage(vax_MB |> filter(location!="Manitoba")) + th
```

## Alberta

[Alberta
Health](http://www.ahw.gov.ab.ca/IHDA_Retrieval/selectSubCategoryParameters.do)
published vaccine coverage estimates for 2014-2023 with vaccine uptake
by birth cohort, and are dissagregated by zone, subzone, and local area.

> Of note, from January 1st 2021, MMR-Var dose 2 schedule was changed
> from aged 4 years old to 18- months.

LH: From what I can see, the Alberta 1-dose estimates represent 1+. (For example, the 1-dose estimates are slightly larger than the 2-dose estimates and together sum to greater than 1). The only exception would be 2-year-olds prior to year 2021 (i.e. year_birth =< 2019), who would not have received the second vaccine under the old vaccination schedule. I have hence relabelled 1-dose to 1+ in `vax_AB` in all relevant cases, but please revert if there is disagreement. All other provinces seem correctly labelled.

```{r}
#| label: vax_AB

AB_zones <- read_xlsx(here::here("data", "measles_raw", "AB_zones.xlsx")) # lookup table for the zones
vax_AB <- read.csv(here::here("data", "measles_raw", "vax_alberta.csv"))  %>%
  select(-Standard.Error, Standard.Score) %>%
  mutate(year_birth = Birth.Cohort,
         n_doses = word(Immunization.Type, 3)) %>%
  rename(year_report = Year) %>%
  group_by(Geography, n_doses, Birth.Cohort, year_report) %>%
  mutate(n = n(),
         uptake = ifelse(n == 1, Immunization.Percent, last(Immunization.Percent) ),
         zone_type = AB_zones$Calssification[match(Geography, AB_zones$Code)],
         Geography = ifelse(!is.na(zone_type), AB_zones$Name[match(Geography, AB_zones$Code)], Geography)
       ) %>%
  ungroup() %>%
  transmute(
    pt = "AB",
    location = Geography,
    year_birth,
    year_report,
    n_doses,
    value = uptake/100,
    zone_type
  )  

# relabel 1-dose to 1+ where needed
vax_AB <- vax_AB %>% mutate(age = year_report - year_birth) %>%
  mutate(dose_corrected = ifelse(n_doses=="1" & (age>2 | (age==2 & year_birth>2019)), "1+", n_doses)) %>%
  mutate(n_doses = dose_corrected) %>% select(-dose_corrected, -age) |> dplyr::mutate(source = "PT")

#Alberta
DT::datatable(vax_AB |> arrange(year_birth))
```

```{r}
#| label: AB_plot
#| fig-height: 3
#| fig-width: 9

plot_coverage(filter(vax_AB, zone_type == "PT"), x = "year_birth") + th
```

```{r}
#| label: AB_zones_plot
#| fig-width: 9

plot_coverage(filter(vax_AB, zone_type == "Alberta Zone"), x = "year_birth") + th
```

```{r}
#| label: AB_subzones_plot
#| fig-height: 14
#| fig-width: 9

plot_coverage(filter(vax_AB, zone_type == "Subzone"), x = "year_birth", ncol = 5) + th
```

## British Columbia

[This BCCDC
dashboard](http://www.bccdc.ca/health-professionals/data-reports/childhood-immunization-coverage-dashboard)
provides estimates for 2-year olds vaccinated with one dose against
measles ("MMR") and for 7-year olds vaccinated with two dose
("Measles").[^2]

[^2]: The data are trapped in a Tableau dashboard, so they have been
    downloaded as PDF (using the Download icon at the bottom of the
    dashboard), and then the relevant table was pulled out using
    `Microsoft Excel > Data > Get Data > From File > From PDF` (with
    some manual tweaks to fix bad parsing).

Methods for the 2-year old data are
[here](http://www.bccdc.ca/resource-gallery/Documents/Statistics%20and%20Research/Statistics%20and%20Reports/Immunization/Coverage/Data%20Notes_2yo_2023_lowres.pdf).
Methods for the 7-year old data are
[here](http://www.bccdc.ca/resource-gallery/Documents/Statistics%20and%20Research/Statistics%20and%20Reports/Immunization/Coverage/Data%20Notes_7yo_2023_lowres.pdf).

```{r}
#| label: file_BC
file_BC <- here::here("data", "measles_raw", "vax_british-columbia.xlsx")
```

These data are pretty granular on the geographic level; alongside the
province-wide estimates we have estimates for each Health Authority
(HA), and each Health Service Delivery Area (HSDA), a sub-unit of each
HA.

The mapping between HAs and HSDAs is captured here:

```{r}
#| label: zones_BC
zones_BC <- suppressMessages(read_excel(file_BC, sheet = "zones"))

DT::datatable(zones_BC)
```

```{r}
#| label: vax_BC

# helper to parse BC's highly structured sheets
parse_sheet <- function(sheet_name){
  df <- (suppressMessages(read_excel(file_BC, sheet = sheet_name)))
  
  age <- gsub("age", "", sheet_name) |> as.numeric()
  n_doses <- ifelse(age==2, "1", "2")

  (df
    |> fill(`Health Authority`, .direction = "down")
    |> mutate(
        pt = "BC",
        zone_type = case_when(
        `Health Authority`=="BC" ~ "PT",
        `Health Service Delivery Area`=="Total" ~ "Health Authority",
        T ~ "Health Service Delivery Area"
      ),
      location = case_when(
        `Health Authority`=="BC" ~ "British Columbia",
        `Health Service Delivery Area` == "Total" ~ `Health Authority`,
        T ~ `Health Service Delivery Area`
      ))
    |> select(-c(`Health Authority`, `Health Service Delivery Area`))
    |> pivot_longer(cols = where(is.numeric), names_to = "year")
    |> transmute(
      pt,
      location,
      year_report = as.numeric(year),
      age = {{age}},
      n_doses = {{n_doses}},
      value,
      zone_type
    )
    |> arrange(location, age)
    |> mutate(age = as.character(age))
  )
}

# get data
sheets <- readxl::excel_sheets(file_BC)
age_sheets <- sheets[grepl("age", sheets)]
vax_BC <- lapply(age_sheets, parse_sheet) |> bind_rows() |> dplyr::mutate(source = "PT")

DT::datatable(vax_BC)
```

```{r}
#| label: BC_plot
#| fig-height: 4

plot_coverage(vax_BC |> filter(zone_type=="PT")) + th
```

```{r}
#| label: BC_HA_plot
#| fig-height: 10
plot_coverage(vax_BC |> filter(zone_type!="Health Authority"), ncol = 5) + th
```

```{r}
#| label: BC_HSDA_plot
#| fig-height: 10
plot_coverage(vax_BC |> filter(zone_type=="Health Service Delivery Area"), ncol = 5) + th
```

## Northwest Territories

A
[report](https://ournthssa.ca/wp-content/uploads/2024/09/HCP-Alert-Respiratory-season-update-September-17-2024-Final.pdf)
released by the Northwest Territories Health and Social Services in 2024
states that:

> 85% of NWT children aged two had received at least 1 dose of measles

```{r}
#| label: vax_NT
#| 
vax_NT <- tibble::tibble(
  pt = "NT", 
  location = "Northwest Territories", 
  year_report  = 2023,
  age = "2",
  n_doses = "1+" , 
  value =  0.85,
  source = "PT"
)

DT::datatable(vax_NT)
```

```{r}
#| label: NT_plot
#| fig-height: 3

plot_coverage(vax_NT)
```

# cNICS data {#cNICS}

We also pulled estimates from
[cNICS](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5185) for all PTs. For **Nova Scotia, Nunavut, Saskatchewan, New Brunswick, Yukon**, these are the only publicly available estimates for children.

For cNICS:

-   By-PT estimates were obtained from [Statistics Canada Table 13-10-0870-01  Vaccine coverage estimates for recommended vaccines in children and pregnant women](https://doi.org/10.25318/1310087001-eng)
-   cNICS reports a value with a 95% CI. Below we only capture the value.
-   For measles, estimates are available for two and seven year olds, though the latter is only available for 2013. We assume proportions are estimated for *at least* one dose as the number of doses does not seem to be captured in the questionnaire with respect to measles (just whether or not the child has been vaccinated for measles).

```{r}
#| label: vax_cNICS

vax_cNICS <- suppressWarnings(readr::read_csv(here::here("data", "1310087001-eng.csv"), skip = 13, col_names = c("location", "age", "year_report", "value"), na = c("", "NA", ".."), show_col_types = FALSE)) |>
  tidyr::fill(location, age, .direction = "down") |>
  dplyr::mutate(
    age = str_extract(age, "\\d+(?=-year-old)"),
    n_doses = "1+",
    value = value/100,
    source = "cNICS"
  ) |>
  dplyr::left_join(
    pt_lookup(), by = join_by(location)
  ) |>
  dplyr::transmute(
    pt, location, year_report, age, n_doses, value, source
  ) |>
  tidyr::drop_na()
```

```{r}
#| label: cNICS_plot

plot_coverage(vax_cNICS)
```

# Summary across PTs {#summary-across-pts}

We aggregate and export the standardized data across sources:
```{r}
#| label: collect_export_vax
#| eval: true

df_names <- c(
  paste0("vax_", 
    c("NL", "PE", "QC", "ON", "MB", "AB", "BC", "NT", # PT data
      "cNICS", "aNICS")), 
  "vax_ON_dose"
  )
vax <- lapply(df_names, get) |> bind_rows() # all estimates

vax |>
  # enforce variable ordering as in data dictionary
  select(all_of(data_dict$variable)) |>
  readr::write_csv(here::here("data", "measles_vax-coverage-data.csv"))
```

How much data is available from each PT (cNICS and aNICS excluded) can be seen from the following plot:

```{r}
#| label: manipulate_age_data

vax_plot  <- vax %>%
  filter(source == "PT") %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%
  # keep only PT-level data
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>%
  filter(location == location.pt) %>%
  select(-location.pt)
```

```{r}
#| label: overall_summary
#| fig-height: 4
#| fig-width: 9

plot_coverage(vax_plot, ncol = 5) + th
```

If we transform the estimates to **current age** by simply taking the
age at estimate and adding the number of years that have elapsed between
the production of the estimate (`year_report`) and the present year, we
can get a sense of how much vaccine coverage information we have across
PTs by age:

```{r}
#| label: overall_summary_transform_age
#| fig-height: 4
#| fig-width: 9

vax_plot2 <- (vax_plot |>
    numericize(age) |>
    mutate(age_current = age + 2025 - year_report)
)

plot_coverage(vax_plot2, x = "age_current", ncol = 5) + th
```

Note that this calculation is rather rough and could result in an
*underestimate* of vaccine coverage; all of these estimates are at least
a year old, and it's possible that more vaccines have been administered
to these cohorts in the meantime. It's worth noting that
immigration/emigration, which is not captured by this calculation, could
also affect these estimates in either direction (it could increase or
decrease them).

After transforming to current age, there are multiple estimates for a
given age and number of doses in some cases. If we take the **most
recent estimate** available in each case, we get the following coverage
curves (using simple linear interpolation):

```{r}
#| label: layer_line_under

layer_line_under <- function(p){
  p <- p + geom_line(colour = "grey50") 
  
  line_indx <- which(unlist(lapply(p$layers, function(x) "GeomLine" %in% class(x$geom))))
  
  p$layers <- c(p$layers[line_indx], p$layers[-line_indx])

  p
}
```

```{r}
#| label: overall_summary_transform_age2
#| fig-height: 8
#| fig-width: 10

vax_plot2 |>
  group_by(location, age_current, n_doses) |>
  filter(year_report == max(year_report)) |>
  plot_coverage(x = "age_current", ncol = 4) |>
  layer_line_under() + th
```

# Basic data analysis

We wish to understand how many observations we have of a given type to aid with selection of appropriate statistical modelling methods. Number of observations by province, by birth cohort etc. can be extracted from the `vax` dataframe (excluding aNICS):

```{r}
#| label: data_analysis

#' set up data frames
vax_all <- vax |> filter(source != "aNICS") %>% mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%
  numericize(age) %>% mutate(year_birth = ifelse(is.na(year_birth) & !is.na(year_report) & !is.na(age), year_report - round(age), year_birth)) %>% select(-zone_type)  # manipulate age data and year_birth data to fill in blanks

vax_provincial <- vax_all %>% left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>% filter(location == location.pt) %>% select(-location.pt)  # filter out sub-province etc data

#' calculate counts
#' @param total_obs Total number of observations for each province, including sub-province etc data.
#' @param province-level_obs Number of province-level observations for each province.
#' @param birth_obsXXXX Number of province-level observations for each birth cohort.
data_counts <- vax %>% group_by(pt) %>% summarise("total_obs" = n()) %>%
  right_join(vax_provincial %>% group_by(pt) %>% summarise("province-level_obs" = n()), join_by(pt)) %>%
  right_join(vax_provincial %>% group_by(pt, year_birth) %>% summarise("birth-cohort_obs" = n()) %>% pivot_wider(names_from = year_birth, values_from = "birth-cohort_obs", names_prefix = "birth_obs"), join_by(pt))

DT::datatable(data_counts)
# could add a row at the end for Country totals (sum column)
```

## Vaccine uptake rates by cohort over time

It may also be good to gauge where we have multiple observations of the same cohort _by dose_ (since this could help us quantify rate of vaccine uptake to some degree):

```{r}
#| label: uptake_rate

# look at counts separated by dose number
count_by_pt_dose <- vax_provincial |>
  group_by(pt, year_birth, n_doses) |>
  count()

# multiple obs of the same cohort
count_by_pt_dose |> filter(n > 1)
```

It looks like the only PT for which we have a chance of understanding uptake behaviours across cohorts is AB.

```{r}
#| label: uptake_rate_2
count_by_pt_dose |> filter(n > 1, pt == "AB") -> lookup

vax_provincial |> 
  right_join(lookup, by = join_by(pt, year_birth, n_doses)) |> 
  arrange(year_birth, n_doses, year_report) |>
  # calculate first differences of vaccine coverage
  group_by(pt, location, year_birth, n_doses) |> 
  mutate(
    age_range = forcats::as_factor(paste(lag(age), age, sep = "-")),
    value_first_diff = (value-lag(value))/(year_report-lag(year_report))) |>
  filter(!is.na(value_first_diff)) |>
  ggplot(aes(x = year_birth, y = value_first_diff, colour = age_range)) +
  geom_point() +
  facet_wrap(~n_doses, labeller = label_both) +
  scale_x_continuous(labels = scales::label_number(accuracy = 1, big.mark = "")) + 
  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) + 
  labs(
     x = "Birth cohort",
     y = "Annual change in coverage (absolute)",
     title = "Annual rate of uptake for measles vaccine by birth cohort (AB)",
     colour = "Current age range"
  )
```

It looks like all changes are positive, which would indicate overall increasing coverage in the cohort over time (as expected). However, the data overlap only occurs for older cohorts, by which time you would expect a relatively small and stable increase in vaccine uptake over time (which is what is observed for first doses). It seems like there may be a recent decrease in second doses between ages 7-13...

NOTE: are linear first differences appropriate if we believe there to be saturating behaviour? probably not...

Overall, I have to conclude that, in general, this dataset does not include multiple observations of the same cohort (for a given dose number). Where it does (AB), the data are still quite sparse (two observations per cohort), and the rates of change in uptake that we can calculate are for relatively old ages (7-17). I'm not sure this will help us "age" estimates...