---
title: "Code demo: Gaussian Process model for estimating vaccine coverage"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
  - name: Irena Papst
    email: irena.papst@phac-aspc.gc.ca
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-overflow: wrap
    theme: cosmo
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Set up

```{r}
#| label: setup

# Load packages, R functions, and cleaned coverage datasets
source(here::here("results", "load-data.R"))
source(here::here("results", "model-selection.R")) # common grid for model selection + helper function

theme_set(theme_bw())
```

## A note on scales

Model fitting uses Gaussian distributions, whose domain is the entire real line. The vaccine coverage data, however, is restricted to $[0,1]$ (though in practice it never sits exactly on the boundaries). We therefore transform the coverage data with a logit function to map it from $(0,1)$ to $(-\infty, \infty)$, fit the model on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.

# One-dimensional case (current age)

A Gaussian Process model assumes that the independent variable(s) are related. In the one-dimensional case, where current age is the only covariate, we are assuming that ages which are 'close-by' (defined via the shape of the covariance function and the lengthscale parameter `l`) will have similar vaccination coverage. 

The one-dimensional model can be run for any single province/territory (PT) in isolation, _e.g._ Alberta.

```{r}
vax_prov <- vax_clean %>% filter(pt=="AB")
```

## Define covariance function and matrix

We begin with the squared exponential covariance function, `k`, and assume a lengthscale of `l = 2`. Our dataset has 40 inputs (current ages), so the covariance matrix, $K$, is a 40x40 matrix defining the relationship between each pair of current ages.

```{r}
#| label: covariance

# ages over which to predict
first_agecurrent <- 5
last_agecurrent <- 32
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  
npts <- length(xvals)
l <- 2 # lengthscale; in units of age, roughly specifies the size of the neighbourhood for ages, within which vaccine coverages are most strongly related
# evaluate cov function pairwise for current ages using the R/ helper
covmat <- generate_ksqexp_covmat(xvals, xvals, fn=ksqexp, l=l)
```

## Define prior distribution and draw a sample

Define a prior \~ Normal$(0, K)$. This prior describes the relationship between vaccination coverage and current age. It is uniquely defined by mean 0 (we centre the data using the sample mean) and variance K. We take 50 draws (curves) from the prior and plot them:

```{r}
#| label: prior
#| fig-height: 3

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% 
  group_by(draw) %>% 
  ggplot(aes(x=x_i,y=prior,group=factor(draw))) + 
  scale_x_continuous(name="Current age") + 
  geom_point(alpha=0.3) + 
  geom_line(alpha=0.2) + 
  labs(
    title = "Prior draws",
    y = "Vaccine coverage (logit scale)"
  )
```

## Introduce observed data

The mini dataset `vax_prov` has `r nrow(vax_prov)` data points. We expose the model to a subset (8) of them, then see how well it predicts the rest. We use the logistic transform of `y` (`logit(yobs)`), which maps from \[0, 1\] to \[-inf, inf\].

```{r}
#| label: observe_data

# observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30
age_obs <- c(4,9,13,19,20,21,25,30)
xobs <- vax_prov %>% filter(age_current %in% age_obs) %>% pull(age_current)   # can optionally add an error term here
yobs <- vax_prov %>% filter(age_current %in% age_obs) %>% pull(value)

tibble(xobs, yobs, yobs_logit = logit(yobs)) %>%
  knitr::kable(digits = 3, caption = "Observed coverage data (1D)") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```

## Compute posterior distribution

The posterior is uniquely defined by its mean and covariance matrix. We first calculate four block matrices within the full covariance matrix, `kab`, where `a` and `b` are all combinations of `o` (observed, `xobs`) and `u` (unobserved, `xvals`). For example, `kou` is the covariance of observed points `xobs` and unobserved points `xvals`. Then the posterior covariance matrix is given by `kuu - kuo %*% solve(koo) %*% kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

# calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

We take 50 draws from the posterior and plot them against the observed (logit-transformed) data (red points):

```{r}
#| label: plot_posterior
#| fig-height: 3

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% 
  ggplot() + 
  geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + 
  scale_x_continuous(name="Current age") + 
  geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + 
  geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red", size = 2) + 
  labs(
    title = "Posterior draws",
    y = "Vaccine coverage (logit scale)")
```

Inverse-transforming the posterior back to the \[0, 1\] interval gives estimates of missing coverage data in its natural units:

```{r}
#| label: plot_posterior_constrained
#| fig-height: 3

post %>% 
  ggplot() + 
  geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + 
  scale_x_continuous(name="Current age") + 
  scale_y_continuous(
    limits=c(0,1), 
    labels = scales::label_percent(), 
    name = "Vaccine coverage") + 
  geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + 
  geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red", size = 2) +
  labs(title = "Posterior draws")
```

To evaluate model performance, we add in the remaining data that the model has not seen (blue points).

```{r}
#| label: compare_to_data
#| fig-height: 3

post %>% 
  ggplot() + 
  geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + 
  scale_x_continuous(name="Current age") + 
  scale_y_continuous(
    limits=c(0,1), 
    labels = scales::label_percent(), 
    name = "Vaccine coverage") + 
  geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + 
  geom_point(
    data=filter(vax_prov, age_current>=first_agecurrent, age_current<=last_agecurrent), 
    mapping=aes(x=age_current,y=value), 
    col="#208ceaff", size = 2) + 
  geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red", size = 2) + 
  labs(title = "Posterior draws")
```

# Two-dimensional case (current age and PT)

The two-dimensional model adds PT as a second dimension with which to relate vaccine coverage. Here $x_i$ is current age, $y_i$ is a metric representing PT, and the vertical axis ($z_i$) is vaccine coverage. For this demo, we use the Gini index by PT for the $y_i$ values (as reported by [Statistics Canada](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810009601)).

## Covariance and prior

We use the squared exponential covariance function, with a current-age lengthscale `l1` and a *relative* lengthscale of current age to PT, `l2` (see paper methods for more details). If the squared exponential is too strong an assumption, we can increase its variance (_i.e._ the model uncertainty) via the parameter `b`.  As before, we define the prior \~ Normal$(0, K)$ and draw a sample.

```{r}
#| label: prior_2D

# ages over which to predict
first_agecurrent <- 5
last_agecurrent <- 21
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  

province_relation <- readr::read_csv(
  here::here("data", "pt-relation", "9810009601_databaseLoadingData_StatCanGini.csv"), 
  show_col_types = F) %>%
  filter(`Inequality measures (5)` == "Gini index on adjusted household after-tax income") %>% 
  rename(location = GEO, value = VALUE, year = `Year (2)`) %>%
  mutate(value = 100*value) %>% 
  select(location, year, value) %>% 
  group_by(location) %>% 
  summarise(value = mean(value)) %>% 
  arrange(value) # PT Gini data from StatCan
yvals <- province_relation$value
names(yvals) <- province_relation$location

npts <- length(xvals)*length(yvals)
l1 <- 2.5  # x lengthscale
l2 <- 10  # relative lengthscale of x values to y values, discussed in paper methods
yvals <- yvals*l2  # scale y values to the x lengthscale

# covariance matrix in 2D
b <- 0.5 # optional function scale for ksqexp determining output variance
xygrid <- expand_grid(x=xvals, y=yvals)
covmat <- generate_2Dksqexp_covmat(xygrid, xygrid, fn=ksqexp, l=l1, b=b)

# define prior and take 50 draws using a helper function in R/ (returns draw, x_i, y_i, y_label, value)
prov_levels <- (province_relation %>% pull(location))  # useful shortcut for defining PT factors
prior2D <- draw_from_rmnorm(
  n=50, mean=rep(0,npts), varcov=covmat + 1e-6*diag(npts),
  xvals=xvals, yvals=yvals, prov_levels=prov_levels) %>%
  rename(prior2D = value)

# plot prior sample
prior2D %>% 
  ggplot(aes(x=x_i, y=prior2D, group=factor(draw))) + 
  scale_x_continuous(name="Current age") +
  geom_line(alpha=0.3) + 
  facet_wrap(~ y_label) + 
  labs(
    title = "Prior draws",
    y = "Vaccine coverage (logit scale)")
```

## Observed data

We expose the model to a subset of coverages of at least one dose:

```{r}
#| label: data_2D

# observe coverage data in all PTs where data exists (excluding 2-dose)
df_subset <- vax_clean %>% 
  filter(
    between(age_current, first_agecurrent, last_agecurrent),
    n_doses=="1+"
  )
xobs <- df_subset %>% pull(age_current)
yobs <- yvals[df_subset %>% pull(location)]
zobs <- df_subset %>% pull(value)

# apply logistic transform and then centre the data around mean 0
logit_zobs_centred <- logit(zobs) - mean(logit(zobs))

tibble(xobs, yobs, zobs, logit_zobs_centred) |>
  head(10) |>
  knitr::kable(digits = 3, caption = "Observed coverage data (2D)") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```

## Posterior

As in the one-dimensional case, the posterior is uniquely defined by its mean and covariance matrix. Again, we calculate the four block matrices giving the pairwise relationships between the unobserved independent variables `{xvals, yvals}` and the observed independent variables `(xobs, yobs)`, then compute the posterior covariance matrix as `kuu - kuo %*% solve(koo) %*% kou`. In the one-dimensional case the unobserved input was `xvals`; now it is the concatenated grid `{xvals, yvals}` of size `r length(xvals)*10`x`r length(xvals)*10` (`r length(xvals)` ages in each of 10 PTs).

```{r}
#| label: posterior_2D

xygrid <- expand_grid(x=xvals, y=yvals)
xyobs <- tibble(x=xobs,y=yobs)

kuo <- generate_2Dksqexp_covmat(xygrid,xyobs,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between unobserved and observed
kou <- generate_2Dksqexp_covmat(xyobs,xygrid,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between observed and unobserved
koo <- generate_2Dksqexp_covmat(xyobs,xyobs,fn=ksqexp,l=l1,b=b) + 1e-6*diag(nrow(xyobs))  # protect against non-invertibleness
kuu <- generate_2Dksqexp_covmat(xygrid,xygrid,fn=ksqexp,l=l1,b=b)

# calculate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit_zobs_centred)  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

We plot a sample of posterior curves against the observed data (red points), first with vaccine coverage on the logit scale, and then in natural units:

```{r}
#| label: plot_posterior_2D

# draw from the posterior distribution using the R/ helper
post2D <- draw_from_rmnorm(
  n=50, mean=post_mean, varcov=post_covmat + 1e-6*diag(nrow(post_covmat)),
  xvals=xvals, yvals=yvals, prov_levels=(province_relation %>% pull(location))) %>%
  rename(post2D = value) %>%
  mutate(post2D_constrained=invlogit(post2D + mean(logit(zobs))))  # add column to return mean and transform back to [0,1]

# plot posterior sample 
yobs_factors <- factor(names(yobs), levels(post2D$y_label))  # convert yobs to factors for use in facet_wrap
post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=logit_zobs_centred, y_label=yobs_factors), aes(x=x, y=y), col="red", size = 2) +
  scale_x_continuous(name="Current age") +
  facet_wrap(~ y_label) + 
  labs(
    title = "Posterior draws",
    y = "Vaccine coverage (logit scale)")

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red", size = 2) +
  scale_x_continuous(name="Current age") +   scale_y_continuous(
    limits=c(0,1), 
    labels = scales::label_percent(), 
    name = "Vaccine coverage") + 
  facet_wrap(~ y_label) +
  labs(title = "Posterior draws")
```

## Model functions

The full two-dimensional Gaussian Process model can be run using `run_GP()`, including plots of the priors and posteriors. Several PT-relation datasets can be supplied via the `prov_values` argument, including:

- `"Gini"` — Gini index on adjusted household after-tax income from StatCan (mean of 2015 and 2020 values)
- `"GDP"` — log(GDP) data from StatCan
- `"low_income_families"` — 2021 rate of children in low-income families (Market Basket Measure) from the Health Inequalities Data Tool
- `"vaccine_hesitancy"` — 2017 prevalence of parents' vaccine hesitancy (refuse all + hesitant) from cNICS

We also include region relation metrics for the UK:

- `"UK-GDP"` — log(2023 UK regional GDP) from ONS
- `"UK-low_income_families"` — FYE 2023 rate of children in low-income families (DWP Official Statistics)
- `"UK-Gini"` — UK Gini index on total wealth (ONS, April 2016–March 2018)

## Measurement error

By default, our GP model treats the observed data as exact, so posterior draws are forced to pass through every observed data point. We relax this by adding measurement error; the posterior mean and covariance matrix are modified when a `meas_error` value is specified in `run_GP()`:

```{r}
#| label: meas_error

run_GP(vax_clean,prov_values="Gini",l1=1.5,l2=3,b=0.75)  # exact
run_GP(vax_clean,prov_values="Gini",l1=1.5,l2=3,b=0.75,meas_error=0.1,ndrws=50)  # with measurement error
```

## Model selection and validation

We select the "optimal" model (via choices for hyperparameters `k`, `l1`, `l2`, `b`, `meas_error`) through leave-one-out cross-validation. For a given covariance function `k` and lengthscale parameters `l1`, `l2`, `b`:

  1.  Leave one data point out.

  2.  Run the GP model and compute an error metric based on the true vaccine coverage for the missing point.
  
Repeat this procedure for every point in the dataset and sum all errors, then select the set of \{`k`, `l1`, `l2`, `b`\} with the smallest total error.

For the error metric, a simple option would be the mean squared error of the prediction compared to the posterior mean at each left-out age. A option that leverages the full richness of GP models is the *posterior predictive density* (PPD), since this encodes both the predicted mean and variance for predictions around each point. In our work, we equivalently maximize the log PPD (LPPD). 

`compute_lppd()` calculates the full log posterior predictive density analytically. `select_GP()` then does a grid search of hyperparameter space and returns the lppd for each hyperparameter set. The "optimal" model (with max LPPD) is also printed to the console.

```{r}
#| label: model_selection

# compute_lppd() returns the analytic lppd for a single hyperparameter set
compute_lppd(vax_clean, prov_values="Gini", k=ksqexp, l1=1.5, l2=3, b=1, meas_error=0.1)

# select_GP() searches a grid and prints the 'optimal' model (max lppd)
select_GP(vax_clean, prov_values="Gini", k_list=list(ksqexp=ksqexp), l1=c(1.5,2), l2=c(3,10), b=c(1), meas_error=0.1)

# example lppd fit
run_GP(vax_clean,l1=2.5,l2=250,b=0.5,meas_error=0.1,ndrws=50)
```

We can also consider how strong of an optimum this "optimal" hyperparameter set is by plotting heatmaps of the LPPD. 

```{r}
#| label: model_selection_heatmaps

# read the Gini model-selection grid (pre-computed since this computation takes a bit of time)
gp_grid <- read_csv(here::here("results", "GPcombinations_Gini_witherror.csv"))

# plot l2 against l1, colouring by lppd, for several meas_error values
gp_grid %>%
  filter(b == 1, meas_error != 0.01, !l2 %in% c(1.5, 2.5, 3.5)) %>%
  mutate(l1 = factor(l1), l2 = factor(l2),
         lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l2, y = l1, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkgreen", na.value = "white", name = "LPPD (> -100)") +
  facet_grid(rows = vars(k_name), cols = vars(meas_error), scales = "free_x",
             labeller = labeller(meas_error = function(x) paste0("\u03c8 = ", x))) +
  labs(title = stringr::str_wrap("Log Posterior Predictive Density (LPPD) as a function of covariance function parameters", width = 60),
  subtitle = "PT relation: Gini index",
       x = "l2", y = "l1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```