---
title: "Initial exploration of the standardized dataset"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: false
---

For now, ignore any additional data / assumptions from the provincial and national sources. Explore only the standardized data set generated by Irena & Rachael, called "measles_vax-coverage-data.csv".

Load functions and packages.

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions
```

Load the standardized data using a code chunk from "measles_vax-coverage-modelling.qmd".

```{r}
#| label: read_data

#' Read in data and filter to province-level
PT_lookup <- readr::read_csv(here::here("data", "pt_lookup.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(PT_lookup$nm_en))  # filtering out sub-provinces etc
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
```

Explore the data by making use of the `plot_coverage()` function. First fill in age for rows where it is not specified (manipulate_age_data code chunk), then plot. We have removed the 18+ data for now.

```{r}
#| label: manipulate_age_data

vax_plot  <- vax_raw %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%
  # keep only PT-level data
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>%
  filter(location == location.pt) %>%
  select(-location.pt) %>% filter(age!="18+")
```

```{r}
#| label: plot_byprovince

plot_coverage(vax_plot)
```

We can also adapt the `plot_coverage()` function to view all data on the same plot (by turning off the location facet_wrap). We can see how much variability there is in the data by for example age at estimate, year of report, birth year (if present), current age.

```{r}
#| label: plot_coverage_country

#' Plot vaccine coverage from our standardized format, adapting to disaggregate by location
#' 
#' @param df a data frame with four required columns: 
#'  - age: age (character)
#'  - value: vaccine coverage as a proportion (numeric)
#'  - location: geographic location (character)
#'  - n_doses: number of doses use to calculate vaccine coverage (character)
#' @param x the column name of the variable to plot on the x-axis (as a string)
#' @param plot_year_report should the plot include the year reported for each estimate? (logical)
#' @param ncol number of columns for the plot grid
plot_coverage_country <- function(df, x = "age", plot_year_report = TRUE, ncol = 3){
  labs_x <- case_when(x=="age" ~ "Age at estimate", x=="year_birth" ~ "Birth year", x=="age_current" ~ "Current age", T ~ x)  
  
  p <- (df
        |> {\(.) if (!is.numeric(.[[x]])) numericize(., !!sym(x)) else .}()
        |> {\(.) if ("n_doses" %in% colnames(.)) factorize(. , n_doses) else . }()
        |> {\(.) if ("year_report" %in% colnames(.)) factorize(. , year_report) else . }()
        |> ggplot(aes(x = !!sym(x), y = value, shape = n_doses))
        #+ facet_wrap(~ location, ncol = ncol)
        #+ scale_x_continuous(expand = expansion(add = 1))
        + scale_y_continuous(
          limits = c(0,1),
          labels = label_percent()
        )
        + scale_shape_manual(values = c(18, 16, 17, 15, 13))
        + labs(
          x = labs_x,
          y = "Vaccine coverage",
          shape = "Number of doses"
        )
  )
  
  if("year_report" %in% colnames(df) & plot_year_report){
    p <- p + geom_point(aes(colour = year_report), size = 2, stroke = 1.5) + scale_colour_viridis_d(option = "G", direction = -1, end = 0.9) + labs(linetype = "Year reported", colour = "Year reported") 
  } else {
    p <- p + geom_point(size = 2)
  }
  
  p
}
```

```{r}
#| label: plot_country_naive

plot_coverage_country(vax_plot)
plot_coverage_country(vax_plot,x="year_report")
plot_coverage_country(vax_plot,x="year_birth")  # some NAs which haven't been calculated & are currently ignored
vax_plot2 <- (vax_plot %>% numericize(age) %>% mutate(age_current = age + 2025 - year_report))
plot_coverage_country(vax_plot2,x="age_current")
```

Note that Ontario's data represents *exactly* 0, 1, 2 doses, where 0 + 1 + 2 sums to 1. This is currently misleading so remove Ontario data for now (and filter out 0 dose data points).

```{r}
#| label: plot_country_notON
plot_coverage_country(vax_plot %>% filter(n_doses %in% c("1","1+","2","2+")) %>% filter(pt!="ON"))
```
