---
title: "Modelling similarly-sparse UK data"
author: "Irena Papst"
format: 
    html:
        embed-resources: true
---

```{r}
#| label: setup

library(dplyr)
library(ggplot2)

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates

attach_country <- function(df, country){
    if("pt" %in% names(df)) df <- df |> select(-pt)

    df |>
        mutate(country = !!country) |>
        relocate(country, .before = location)
}

# Canadian data
# - - -
# for Gini model
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
# for low income cleaned data for use with low-income and vaccine hesitancy province relation indicators
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT"))) # "low_income_families" has no data for NU, NT, YT

# England data
# - - -
vax_england <- readr::read_csv("data/ukhsa-chart-download-mmr1-regions.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (regions of England)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)

vax_gini <- bind_rows(vax_clean |> attach_country("Canada"), vax_england |> attach_country("England")) |> filter(n_doses == "1+")
vax_LI <- bind_rows(vax_cleanLI |> attach_country("Canada"), vax_england |> attach_country("England"))
```

To strengthen the application of the model to the sparse Canadian data, we want to apply it to a similarly sparse subsample of the very complete UK data.

# Experimental design

The main task here is to emulate the sparsity of the Canadian data to some part of the UK data. The closest comparison is likely to be for England, as it has a similar number of sub-regions as the number of PTs in Canada (exact numbers of PTs depends on the PT-similarity metric as we do not have these metrics for all PTs in our standardized data). The PT-similarity metrics we have for England are: Gini index, percentage of children not in low-income families. (There may be data on vaccine hesitancy, but we have not looked for it yet.)

We will want to run the same experiment (N reps) for each PT similarity metric:

1. Subsample the England data to mimic the CA data sparsity ("training" data)
    - Quantify the CA data as a vector of the number of data points available by PT
    - Sample this vector without replacement 8 times (to match the number of regions in England); call this the "sample size vector".
    - For each region of England, draw V vaccine coverage estimates, where V corresponds to an entry of the sample size vector.
    - Designate the rest of the data as "test" data
2. Select GP model hyperparameters based on training data
3. Train the model on training data with selected hyperparameters
4. Calculate an accuracy score for the test data

Then we can summarize accuracy for each PT similarity metric.

## Calculating accuracy

To calculate an accuracy score based on test data, it seems reasonable to use the posterior predictive density (PPD) in some way. However, I don't think summing the (L)PPD values will result in an intuitively interpretable accuracy score, because it's not giving us values that are readily interpretable. Point densities are not probabilities. These values are useful _in comparisson_, _i.e._, if looking for a model that is more or less predictive of left-out data than another one, but it's not straight forward to interpret. 

How should we define the accuracy, then? I discussed this with David Champredon, and we came to the following idea:

1. For each test point, choose an acceptable level of error, _e.g._, within 5% of the true vaccine coverage (the "target interval").
2. Retrieve parameters for the fitted model's (Normal) posterior predictive density at the test point (distribution mean and variance).
3. Calculate the density's probability mass captured within the target interval (difference between a Normal _cumulative_ distribution function evaluated at the edges of the target interval). This tells us how likely it is that the model would have predicted the true value given our chosen error tolerance.

For instance, say true vaccine coverage is 80% for kids aged 7 in London. We choose an error tolerance of 5%, because epidemiologically, these are "similar enough" values. Then our target interval is 77.5%-82.5%. Suppose 70% of the fitted model's probability mass is in that interval; then our model had a 70% chance of predicting the true value with an acceptable amount of error. The more error we tolerate, the higher the accuracy, but this way the accuracy is informed directly by the real-world application, and so it can be interpretted.

The procedure above could be repeated for each test point, and then we could compute summary statistics on it: an average accuracy score along with a range or confidence interval.

The above is a little simplified because we haven't accounted for measurement error in vaccine coverage estimates. Copilot assures me that this is simply handled by "inflating uncertainty via convolution".

## Implementation via Copilot

IP: The following was generated by Copilot and annotated by me.

We assess the accuracy of probabilistic predictions of a response variable \(z\)
(e.g., population‑level vaccine coverage) as a function of two‑dimensional
predictors \((x,y)\). The method uses an application‑defined error tolerance and
explicitly accounts for measurement error in the observed response values.
It yields an interpretable, probabilistic accuracy metric that reflects both
model uncertainty and data uncertainty.

### Conceptual framework

The available data consist of observed predictors and responses,
denoted \(\underline{x}\) and \(\underline{z}\), respectively.
For each test point \((x_a, y_a, z_a)\), the fitted model produces a full
predictive distribution for an unobserved true response value
\(z_{\text{true},a}\) at predictors \((x_a, y_a)\).

Observed response values \(z_a\) are treated as noisy measurements of the true
quantity rather than as error‑free ground truth. Accuracy is defined in terms of
the probability that the model’s prediction is *sufficiently close* to the
observed response, where “sufficiently close” is determined by a
domain‑specific tolerance.

### Model assumptions

Let:

- \((x_a, y_a)\) denote the predictor values for test point \(a\)
- \(z_{\text{true},a}\) denote the true (latent) response
- \(z_a\) denote the observed response
- \(\mathcal D = \{(\underline{x}, \underline{z})\}\) denote the training data

### Predictive model

The fitted probabilistic model yields a posterior predictive distribution for the
true response at the test predictors:
\[
z_{\text{true},a} \mid x_a, y_a, \mathcal D
\sim
\mathcal N(\mu_a, \sigma_a^2),
\]
where \(\mu_a\) is the posterior predictive mean and \(\sigma_a^2\) represents
model‑based uncertainty about the true response.

IP: Can these values can be retrieved from `compute_posterior2D()`, setting `xvals` and `yvals` to the predictor values for the test points, before the call to `draw_from_rmnorm()`? No, I don't think so. The above is a marginal probability and `compute_posterior2D()` yields parameters for the joint probability. I think we should instead just use the formulas in [Retrieving PPD parameters at a test point](#retrieving-ppd-parameters-at-a-test-point) sequentially on the test points.

### Measurement error model

Observed responses are assumed to satisfy:
\[
z_a = z_{\text{true},a} + e_a,
\qquad
e_a \sim \mathcal N(0, \psi^2),
\]
where \(\psi^2\) denotes the variance of the measurement error in the observed
response.

## Accuracy definition

Let \( \varepsilon > 0 \) denote an application‑defined absolute error tolerance
for the response \(z\) (for example, ±5 percentage points of vaccine coverage).

For the test point \((x_a, y_a, z_a)\), accuracy is defined as:
\[
A_a
=
\Pr\!\left(
\lvert z_{\text{true},a} - z_a \rvert \le \varepsilon
\;\middle|\;
\mathcal D
\right).
\]

Because both the prediction and the observation are uncertain, the difference
\[
z_{\text{true},a} - z_a
\sim
\mathcal N\!\left(
\mu_a - z_a,\;
\sigma_a^2 + \psi^2
\right).
\]

This yields the closed‑form expression:
\[
A_a
=
\Phi\!\left(
\frac{\varepsilon - (\mu_a - z_a)}
{\sqrt{\sigma_a^2 + \psi^2}}
\right)
-
\Phi\!\left(
\frac{-\varepsilon - (\mu_a - z_a)}
{\sqrt{\sigma_a^2 + \psi^2}}
\right),
\]
where \(\Phi(\cdot)\) denotes the standard normal cumulative distribution
function.

## Aggregation

Overall expected accuracy across a test set of size \(m\) is computed as:
\[
\bar A = \frac{1}{m} \sum_{a=1}^m A_a.
\]

Uncertainty in performance may be summarized using the empirical distribution of
the \(A_a\) values or via bootstrap resampling.

## Interpretation

For example, an overall accuracy of \(\bar A = 0.70\) can be interpreted as:

> Accounting for both model uncertainty and measurement error with variance
> \(\psi^2\) in the observed responses \(\underline{z}\), the model assigns, on
> average, a 70% probability that its predictions at predictors \((x,y)\) fall
> within the application‑defined acceptable error tolerance.

## Key properties of this accuracy score

- Uses full posterior predictive uncertainty for \(z \mid (x,y)\)
- Explicitly propagates measurement error with variance \(\psi^2\)
- Defines accuracy using substantively meaningful tolerances
- Produces a smooth, probabilistic accuracy metric
- Complements, but does not replace, calibration diagnostics such as predictive coverage

## Intended use

This metric is well suited to population‑level public‑health modeling—such as
predicting vaccine coverage across age–location predictor spaces—where observed
responses are noisy and decision relevance depends on whether predictions are
*close enough* rather than exact.

# Data exploration and prep

```{r}
#| label: n_obs_gini

n_obs <- vax_gini |> group_by(country) |> select(location) |> distinct() |> count()
```

We start with an analysis of the Gini index model, which has data for `r n_obs |> filter(country == "Canada") |> pull(n)` Canadian PTs and `r n_obs |> filter(country == "England") |> pull(n)` regions of England.

The age ranges in Canadian data do not exactly conincide with those found in the England data:

```{r}
vax_gini |>
    group_by(country, location) |>
    summarize(age_min = min(age_current), age_max = max(age_current))
```

If we filter both down to the largest intersection in age for a fair comparison, and we are left with the following ranges by location: 

```{r}
#| label: filter_data
vax <- vax_gini |>
    group_by(country) |>
    mutate(age_min = min(age_current), age_max = max(age_current)) |>
    ungroup() |>
    mutate(age_min = max(age_min), age_max = min(age_max)) |>
    filter(between(age_current, age_min, age_max))
```

```{r}
vax |>
    group_by(country) |>
    group_split() |>
    purrr::map(\(df){
        df |>
            group_by(location) |>
            summarize(
                age_min = min(age_current), age_max = max(age_current)
            )
    })
```

After filtering, we now only have 6 PTs in the Canadian data, so when testing against the England data, we will only take 6 regions at a time to fit.

## Sample train and test set

We start by getting the sample size of the (filtered) PT data:

```{r}
sample_size <- vax |>
    filter(country == "Canada") |>
    group_by(location) |>
    summarize(n = n())
```

Pull a subset of the England data according to the PT sample sizes:

```{r}
vax_england <- filter(vax, country == "England")

sample_by_location <- tibble(
    location = sample(
        unique(vax_england$location), 
        size = nrow(sample_size), 
        replace = FALSE
    ), 
    n = sample(sample_size$n)
)

vax_england_subset <- vax_england |>
  right_join( # filtering join to subsample regions
    sample_by_location,
    by = "location"
  )

vax_train <- vax_england_subset |>
  group_split(location) |>
  purrr::map(\(df) slice_sample(df, n = df$n[1]) |> select(-n)) |>
  purrr::list_rbind()

vax_test <- anti_join(vax_england_subset, vax_train)
```

# Retrieving PPD parameters at a test point

From [Liza's notes](https://github.com/phac-modelling-hub/immunity-landscape/issues/4), the PPD (with measurement error) at a test point $x_a$ with training points $(\underline{x}, \underline{z})$ is given by

$$
\mathcal{P} (z_{a} | x_{a}, \underline{x}, \underline{z}) \sim \mathcal{N}(\mu_{a}, \sigma_{a}^{2})
$$

where 

$$
\mu_{a} = \underline{k}(x_{a}, \underline{x})^{\top} (K + \psi^2I)^{-1}\underline{z},
$$

$$
\sigma_{a}^{2} = k(x_{a}, x_{a}) - \underline{k}(x_{a}, \underline{x})^{T} (K + \psi^2I)^{-1} \underline{k}(x_{a},\underline{x}).
$$

Here, $K$ is the covariance matrix of **training** data (computed with $K_{ij} = k(x_{i},x_{j})$) for a covariance function $k$, and $\psi$ is the measurement error. We test a few covariance functions in selecting model hyperparameters, but one example is the squared exponential:

$$
k(x_{i}, x_{j}) = b \exp \left( -\frac{(x_{i} - x_{j})^2}{2l^2} \right)
$$

This comes from Rasmussen and Williams (2005), equations 2.25 and 2.26, which we can work out for a specific test point as follows:

```{r}
# \mu_{a} = \underline{k}(x_{a}, \underline{x})^{\top} (K + \psi^2I)^{-1}\underline{z},
# \sigma_{a}^{2} = k(x_{a}, x_{a}) - \underline{k}(x_{a}, \underline{x})^{T} (K + \psi^2I)^{-1} \underline{k}(x_{a},\underline{x}).

# set hyperparameters
k <- ksqexp
l1 <- 2.5
l2 <- 2
b <- 0.5
meas_error <- 0.05

# observed values
xobs <- vax_train$age_current
prov_values <- extract_province_relation("UK-Gini", vax_dataset = vax_train)
prov_values <- prov_values * l2
yobs <- prov_values[vax_train %>% dplyr::pull(location)]
x_underline_x <- xobs # to match notation in equation above
x_underline_y <- yobs # to match notation in equation above
zobs <- vax_train$value
centring_term <- mean(LaplacesDemon::logit(zobs))
logit_zobs_centred <- LaplacesDemon::logit(zobs) - centring_term
z_underline <- logit_zobs_centred

# test values
x_a_x <- vax_test$age_current[1] # x component of x_a (actually 2D coordinate)
prov_values <- extract_province_relation("UK-Gini", vax_dataset = vax_test)
prov_values <- prov_values * l2
y_test <- prov_values[vax_test %>% dplyr::pull(location)]
x_a_y <- y_test[[1]]
# test interval
# error_tol <- 0.02
# z_a <- LaplacesDemon::logit(vax_test$value[1]) - centring_term
# z_lwr <- z_a - error_tol/2
# z_upr <- z_a + error_tol/2

# mean's first term: \underline{k}(x_{a}, \underline{x})^{\top}
first_term <- numeric(nrow(vax_train))
for (i in seq_len(nrow(vax_train))){
    r <- sqrt((x_a_x - x_underline_x[i])^2 + (x_a_y - x_underline_y[i])^2)
    first_term[i] <- k(r, l1, b) # l2 included in y scaling
}

# mean's second term: (K + \psi^2I)^{-1}
xyobs <- tibble::tibble(x = xobs, y = yobs)
K <- generate_2Dksqexp_covmat(xyobs, xyobs, fn = k, l = l1, b = b) + 1e-12 * diag(nrow(xyobs))
second_term <- solve(K + diag(meas_error^2, nrow(K)))

# mean's third term: \underline{z} 
third_term <- z_underline

# PPD at x_a's
ppd_mean_a <- (first_term%*%second_term%*%third_term)[1,1]

# PPD at x_a's variance
#\sigma_{a}^{2} = k(x_{a}, x_{a}) - \underline{k}(x_{a}, \underline{x})^{T} (K + \psi^2I)^{-1} \underline{k}(x_{a},\underline{x}).
r_a <- 0 # the distance between x_a and x_a
ppd_var_a <- k(0, l1, b) - (first_term%*%second_term%*%first_term)[1,1]
```

We could compare this to the bulk computation of means and covariances in `compute_posterior2D()` to check that whether they match:

```{r}
# test point
xygrid <- tibble::tibble(
    x = x_a_x,
    y = x_a_y
)

fit <- fit_GP(xygrid = xygrid, xobs = x_underline_x, yobs = x_underline_y, zobs = vax_train$value, k = k, l1 = l1, b = b, meas_error = meas_error)

waldo::compare(fit$post_mean[1,1], ppd_mean_a)
waldo::compare(fit$post_covmat[1,1], ppd_var_a)
```

Very close! Difference is likley from the perturbation on `koo` to protect against non-invertibleness... This should go away once the Cholesky decomposition method of inverting `koo` is implemented.

Compare to new implementation in `compute_accuracy()`:

```{r}
ppd_params <- compute_accuracy(
    vax_train,
    vax_test,
    prov_values = "UK-Gini",
    k = k,
    l1 = l1,
    l2 = l2,
    b = b,
    meas_error = meas_error
)

waldo::compare(ppd_params$ppd_means[1], ppd_mean_a)
waldo::compare(ppd_params$ppd_vars[1], ppd_var_a)
```

# Experiment implementation

Fit the model to the subsampled data:

```{r}
#| label: fit_england_gini

rerun <- FALSE
if(rerun){
    fit <- select_GP(
        vax_train, 
        prov_values="UK-Gini", 
        k_list=list(ksqexp=ksqexp, kexp=kexp),  
        l1=c(1,1.25,1.5,1.75,2,2.25,2.5), 
        l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), 
        b=c(0.5,1,1.5), 
        meas_error=c(0.5,0.1,0.05,0.01,0)
    )

    # temp save since the above takes a while to run
    if(!dir.exists(here::here("results", "tmp"))) dir.create(here::here("results", "tmp"))

    saveRDS(fit, here::here("results", "tmp", "fit_england_gini.rds"))
} else {
    fit <- readRDS(here::here("results", "tmp", "fit_england_gini.rds"))
}

# retreive optimal hyperparameters
fit_opt <- fit |> arrange(desc(lppd_exact)) |> slice(1)
```

## Prototype

IP: everything below needs to be updated, starting with the definition of `compute_accuracy()` based on the algorithm suggested by Copilot to incorporate measurement error in all response variables (including the "true" test values).

```{r}
#| label: compute_accuracy

error_tolerance <- 0.05 # 5%
last_agecurrent <- 21
prov_values <- "UK-Gini"
k <- get(fit_opt$k_name)
l1 <- fit_opt$l1
l2 <- fit_opt$l2
b <- fit_opt$b
meas_error <- fit_opt$meas_error

# to use compute_lppd() equations, need to feed in vax_train _and_ vax_test, then only do the for loop stuff on the indices corresponding with vax_test (add those indices above!)
acc <- compute_accuracy(
    train = vax_train |> transmute(x = age_current, y = ), 
    test = vax_test,
    error_tolerance = error_tolerance,
    last_agecurrent = last_agecurrent,
    prov_values = prov_values,
    k = k,
    l1 = l1,
    l2 = l2,
    b = b,
    meas_error = meas_error
)
```

```{r}
#| label: plot_accuracy

ggplot(data = tibble::tibble(acc = acc), aes(x = acc)) +
    geom_histogram() +
    labs(
        title = "Distribution of accuracy across test points",
         x = "Accuracy",
         y = "Count"
    )
```

Let's look at true value vs accuracy to see if there is a correlation (so perhaps the model only correctly predicts values within a certain range):

```{r}
pp_acc <- ggplot(data = tibble::tibble(true = vax_test$value, acc = acc), aes(x = true, y = acc, colour = acc)) +
    geom_point() +
    scale_x_continuous(limits = range(vax_england$value)) +
    labs(
        title = "True Value vs Accuracy",
        x = "True Value",
        y = "Accuracy"
    )

pp_acc
```

Doesn't seem to be the case... How hard of a prediction problem is this? Here are distributions of the test and train data sampled here under the accuracy scores:

```{r}
pp_dist <- dplyr::bind_rows(
    dplyr::mutate(vax_test, type = "test"),
    dplyr::mutate(vax_train, type = "train")
) |>
    ggplot(aes(x = value, fill = type)) +
    geom_histogram() +
    scale_x_continuous(limits = range(vax_england$value)) +
    facet_grid(rows = vars(type)) +
    labs(
        title = "Distribution of vaccine coverage values in England",
        x = "Vaccine Coverage",
        y = "Count"
    )

library(patchwork)
pp_acc / pp_dist
```

There is high accuracy for test points in the range that overlaps with the bulk of the training data, but there is also reasonably high accuracy outside of this range. The model seems to be doing something in extrapolating from the training data...

Another way to visualize would be to create a plot of posterior draws by age and region, include the train obs as black points, and then test obs coloured by accuracy (on a continuous scale between 0 and 1).[^1] This would help us see which points have been easily predicted and which have not. Though this is already, to some extent, eye-ballable when test points are plotted overtop of posterior draws because that already gives an impression of prediction mean/variance under the test point... so maybe not worth spending time on?

[^1]: I think this could be done relatively easily, but it would be good to repurpose existing code to plot posterior draws for this (Liza suggests using `plot_posterior2D()`)?

Note, the above is just for one test set, so we would want to repeat this process multiple times (with different random subsamples of the England data). That would give us multiple sets of accuracy scores, and we could run summary stats on this. We could then also run a sensitivity analysis of these scores to the error tolerance to show just how robust the framework is. 

The above is also only for the Gini model. We could repeat for the low-income model and then compare accuracy values between them to see if one PT similarity metric is potentially more predictive of vaccine coverage than the other.

This is getting to be a bit of a data heavy experiment... may want to consider moving to a `{targets}` pipeline.