---
title: "Full Gaussian Process model for estimating vaccine coverage"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

# Set Up

Load functions and packages. Read the standardized cleaned data from the 'take 2' preliminary modelling approach (aiming to collapse different dose estimates to 1+). The data set `vax_clean` contains province-level estimates. We aim to work with current age (effectively collapsing year_birth and age dimensions into one).

```{r}
#| label: initial_setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(stringi)
library(scales)
library(purrr)
library(tibble)
library(mnormt) # for multivariate normal
library(LaplacesDemon) # for logistic transform

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read cleaned data from 'take 2' approach, from data processing qmd file
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
vax_cNICS <- readr::read_csv(here::here("data", "measles_CNICS.csv"), show_col_types = FALSE) %>%
  filter(!(pt %in% c("SK","YT","NB","CA"))) %>% mutate(age_current=current_year - year_birth)

#' cleaned data for use with low-income and vaccine hesitancy province relation indicators
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT")))  # "low_income_families" has no data for NU, NT, YT
vax_cleanVH <- vax_clean %>% filter(n_doses!="2+") %>% mutate(location = if_else(location %in% c("Newfoundland and Labrador", "Prince Edward Island", "Nova Scotia", "New Brunswick"), "Atlantic region", location)) %>% mutate(location = if_else(location %in% c("Yukon", "Northwest Territories", "Nunavut"), "Northern region", location)) %>%  mutate(pt = if_else(location %in% c("Atlantic region"), "AR", pt)) %>% mutate(pt = if_else(location %in% c("Northern region"), "NR", pt)) %>% group_by(across(-value)) %>% summarise(value = mean(value), .groups = "drop")  # "vaccine_hesitancy" groups Atlantic PTs and Northen PTs
```

## A note on scales

Model fitting is done using Gaussian distributions, whose domain is the entire real line. However, the vaccine coverage data is restricted to $[0,1]$ (though in practice, they're not on the boundaries of this interval). Thus, we transform the vaccine coverage data using a logit function to map it from $(0,1)$ to $(-\infty, \infty)$, perform the model fitting on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.

## A note on assumptions

A Gaussian Process model assumes relation of the independent variable(s), i.e. that ages which are 'related / close-by' (defined however we wish with lengthscale parameter l) will have similar vaccination coverage. Equally, we assume that 'related' provinces will also have similar coverage. Some tests of this assumption are in the 2D model extensions section.

# 1D case (current age)

The 1D model can be run for any single province, e.g. Alberta.

```{r}
vax_prov <- vax_clean %>% filter(pt=="AB") #%>% filter(n_doses!="2+")
```

## Define covariance function / covariance matrix

Use the squared exponential covariance function, `k`, to begin with and assume lengthscale `l=2`. Our dataset has 40 inputs/ independent variables (current age), so our covariance matrix is a 40x40 matrix defining the 'relationships' between each pair of current ages.

```{r}
#| label: covariance

# ksqexp <- function(r,l,b=1) {  # this is also defined in ksqexp.R
#   b*exp(-r^2 / (2*l^2))
# }

first_agecurrent <- 5  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  
npts <- length(xvals)
l <- 2
#' evaluate cov function pairwise for current ages x1_i and x2_i (i=1,..,n)
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>% 
  pull(k) %>% matrix(nrow=npts)  # could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). This prior describes the relationship between vaccination coverage and current age. It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (current ages). Vertical axis is vaccination coverage.

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") + geom_point(alpha=0.3) + geom_line(alpha=0.2)
```

## Introduce observed data

For this mini dataset `vax_prov`, we have a dataset of n data points. First expose the model to \~8 data points (to see how well it can predict the remainder). We will use the logistic transform of y (`logit(yobs)`) which maps from \[0,1\] to \[-inf,inf\].

```{r}
#| label: observe_data

#' observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30
xobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(age_current)   # can optionally add an error term here
yobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(value)
# xobs <- vax_prov %>% pull(age_current)   # can optionally add an error term here
# yobs <- vax_prov %>% pull(value)

tibble(xobs, yobs, logit(yobs))
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved current ages `xvals` and observed current ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

#' calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed logit-transformed data. The posterior function can be used to provide estimates of our missing coverage data if inverse-transformed back to a \[0,1\] interval (second plot).

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red")

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```

We can add in the remaining true data (which the model has not seen) to evaluate its performance. The true coverage values are marked as red (seen) and green (not seen) points.

```{r}
#| label: compare_to_data

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(name="current age x_i", breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=(vax_prov %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% pull(age_current)), y=(vax_prov %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% pull(value))), aes(x=x,y=y), col="green") + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")

#ggsave(paste0("plots/ABpost_l",l,"_ksqexp",".pdf"), height=4.51, width=7.29)
```

## Extensions

To improve this model, the next step would be to experiment with choice of covariance function `k` and lengthscale `l`, and to add in province as a second dimension. It is also worth considering how to treat dose data - currently the model struggles with data points for 2 dose. It may be worth fitting two fairly separate GPs (1+ and 2 dose), or using the 2 dose estimates as minimums for the prior in some way.

# 2D case (current age and province)

The 2D model adds province as a second dimension. In this model, x_i is current age, y_i is province, and vertical axis (z_i) is always estimated vaccine coverage. Note: more information is needed to inform relationship between provinces.

## Covariance and prior

We use the squared exponential covariance function to begin with, with current age lengthscale, `l1`, and *relative* lengthscale of current age to province, `l2`. If the squared exponential is too strong an assumption, we can also increase its variance (i.e. the model uncertainty) with parameter `b`. Relation of provinces currently uses 2024 GDP from StatCan (https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610040201) -- this should be updated when a better method is found. Define the prior \~ Normal(0,cov) as before and draw a sample.

Note that we currently assume a mean of 0. We could specify a non-zero mean here in the prior, or equivalently, we could centre the observed data to mean 0, run the Gaussian Process as normal and then rescale the data back at the end. We will do the latter as it simplifies calculation.

```{r}
#| label: prior_2D
#run_GP(vax_clean, prov_values = "Gini", k=ksqexp, l1=2.5, l2=10, b=0.5, meas_error=0.1)
#' setup
#' now both x and y are inputs, with output z
first_agecurrent <- 5  # first age for the GP model
last_agecurrent <- 21  #age_pre1970-1  # last age for the GP model
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced) 

province_relation <- province_relation_data <- readr::read_csv(here::here("data", "9810009601_databaseLoadingData_StatCanGini.csv"), show_col_types = F) %>%
        filter(`Inequality measures (5)` == "Gini index on adjusted household after-tax income") %>% rename(location = GEO, value = VALUE, year = `Year (2)`) %>%
        mutate(value = 100*value) %>% select(location, year, value) %>% group_by(location) %>% summarise(value = mean(value)) %>% arrange(value) # province Gini data from StatCan
yvals <- log(province_relation %>% pull(value))  # a vector of provinces -- currently log(GDP)
names(yvals) <- province_relation %>% pull(location)

npts <- length(xvals)*length(yvals)
l1 <- 2.5  # x lengthscale
l2 <- 10  # relative lengthscale of x values to y values
yvals <- yvals*l2  # scale y values to the x lengthscale

#' covariance matrix in 2D
b <- 0.5 # optional function scale for ksqexp determining output variance
covmat <- expand_grid(x1=xvals, y1=yvals, x2=xvals, y2=yvals) %>% mutate(k=ksqexp(sqrt((x1-x2)^2 + (y1-y2)^2), l1, b=b)) %>%
  pull(k) %>% matrix(nrow=npts)

#' define prior and take 50 draws
xygrid <- expand_grid(x=xvals, y=yvals)
prior2D <- tibble() 
for(i in 1:50){  # 50 realisations from a 2D gaussian process, with both x, y as independent variables
  prior2D <- bind_rows(prior2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, prior2D=rmnorm(1, rep(0,npts), covmat + 1e-6*diag(npts))))
}

#' plot prior sample
prov_levels <- (province_relation %>% pull(location))  # useful shortcut for defining province factors
prior2D <- prior2D %>% mutate(y_label = names(yvals)[match(y_i, yvals)]) %>% 
  mutate(y_label = factor(y_label, levels=prov_levels))
prior2D %>% ggplot(aes(x=x_i, y=prior2D, group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") +
  geom_line(alpha=0.3) + facet_wrap(~ y_label)

# prior2D %>% ggplot(aes(x=x_i, y=invlogit(prior2D + mean(logit(zobs))), group=factor(draw))) +  # transformed centred plot for talks
#   scale_x_continuous(name="current age x_i", limits=c(1,32), breaks=xvals) + ylab("prior") + ylim(c(0,1)) +
#   geom_line(alpha=0.3) + facet_wrap(~ y_label) + theme(axis.text.x = element_text(size = 6))
```

## Observed data

Expose the model to the 1 dose data points. Exclude 2-dose data. We will use the logistic transform of z (`logit(zobs)`) which maps from \[0,1\] to \[-inf,inf\]. Our GP uses a prior mean of 0 so we centre the data around 0 after transforming.

Extension: A possible model extension is to add 2-dose data as minimums at this point. We are currently considering how to do this...

```{r}
#| label: data_2D

#' observe coverage data in all provinces where data exists (excluding 2-dose)
xobs <- vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% filter(n_doses!="2+") %>% pull(age_current)
yobs <- yvals[vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% filter(n_doses!="2+") %>% pull(location)]
zobs <- vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% filter(n_doses!="2+") %>% pull(value)

#' optional extension: add 2+-dose data as minimums?
# vax_obs <- vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2+")  # unclear how to incorporate as a minimum

#' apply logistic transform and then centre the data around mean 0
logit_zobs_centred <- logit(zobs) - mean(logit(zobs))

tibble(xobs, yobs, zobs, logit_zobs_centred)
```

## Posterior

As in the 1D case, the posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved independent variables `{xvals, yvals}` and observed independent variables `(xobs, yobs)`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`. In the 1D case, our unobserved was `xvals`, now it is the concatenated vector grid `{xvals, yvals}` of size 320x320 (32 ages in each of 10 provinces).

Plot a sample of posterior curves against the observed data (red points).

```{r}
#| label: posterior_2D

xygrid <- expand_grid(x=xvals, y=yvals)
xyobs <- tibble(x=xobs,y=yobs)

kuo <- generate_2Dksqexp_covmat(xygrid,xyobs,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between unobserved and observed
kou <- generate_2Dksqexp_covmat(xyobs,xygrid,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between observed and unobserved
koo <- generate_2Dksqexp_covmat(xyobs,xyobs,fn=ksqexp,l=l1,b=b) + 1e-6*diag(nrow(xyobs))  # protect against non-invertibleness
kuu <- generate_2Dksqexp_covmat(xygrid,xygrid,fn=ksqexp,l=l1,b=b)

#' calculate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit_zobs_centred)  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance

#' optional: examine koo
max(koo)
mean(koo)
median(koo)
min(koo)
```

```{r}
#| label: plot_posterior_2D

#' draw from the posterior distribution and record in dataframe post2D
post2D <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post2D <- bind_rows(post2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, post2D=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post2D <- post2D %>% mutate(post2D_constrained=invlogit(post2D + mean(logit(zobs)))) %>%  # add column to return mean and transform back to [0,1]
  mutate(y_label = names(yvals)[match(y_i, yvals)]) %>%  # add column to return province names
  mutate(y_label = factor(y_label, levels=(province_relation %>% pull(location)))) %>%  # order provinces
  select(draw, x_i, y_i, y_label, post2D, post2D_constrained)  # reorder columns

#' plot posterior sample 
yobs_factors <- factor(names(yobs), levels(post2D$y_label))  # convert yobs to factors for use in facet_wrap
post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=logit_zobs_centred, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") +
  facet_wrap(~ y_label)

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)


#' to plot only the data, as facets
post2D %>% ggplot() +
  geom_point(data=tibble(x=xobs, y=zobs*100, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(limits=c(first_agecurrent, last_agecurrent), breaks=seq(first_agecurrent, last_agecurrent, by=5), name="current age") +
  scale_y_continuous(limits=c(0,100), name="vaccine coverage (%)") +
  facet_wrap(~ y_label) +
  theme(axis.text=element_text(size=12), axis.title=element_text(size=20), plot.title = element_text(size=25), strip.text.x = element_text(size=9))
ggsave(paste0("plots/1+dose_data_plot",".jpeg"), height=4.51, width=7.29)

# #' to plot only the data, as separate image files
# province_levels <- levels(yobs_factors)
# walk(province_levels, function(prov) {
#   # subset observations for this province
#   idx <- which(yobs_factors == prov)
#   df_obs <- tibble(x = xobs[idx], y = zobs[idx]*100, y_label = prov)
#   p <- ggplot() +
#     geom_point(data=df_obs, aes(x=x, y=y), col="red") +
#     scale_x_continuous(limits=c(first_agecurrent, last_agecurrent), breaks=seq(first_agecurrent, last_agecurrent, by=5), name="current age") +
#     scale_y_continuous(limits=c(0,100), name="vaccine coverage (%)") +
#     ggtitle(paste("Province:", prov))
#   ggsave(filename = paste0("plots/1+dose_data_plot_", prov, ".jpeg"), plot=p, height=4.51, width=7.29)
# })
```

As before, we can compare performance by adding back in data points that the model has not seen in green (including for example 2-dose data):

```{r}
#| label: compare2D

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) + 
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) + 
  geom_point(data=tibble(x=(vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% pull(age_current)),  # adding in full data from vax_clean
                         y=(vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% pull(value)),
                         y_label=factor((vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent) %>% pull(location)), levels=prov_levels)),
                         aes(x=x,y=y), col="green") + 
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)

#ggsave(paste0("plots/post2D_l1",l1,"_l2",l2,"_ksqexp","_alldata",".pdf"), height=4.51, width=7.29)
```

Optional: We can also drop any posterior draws that do not satisfy the 2-dose data points as a minimum. Due to the underlying mathematics, this is (*we think*) mathematically equivalent to constraining the prior and proportionally reweighting the remaining valid draws. This turns out to be too strong a constraint so the code snippet is skipped.

```{r}
#| label: set_minimums2D
#| eval: false

# extract 2+-dose data from vax_clean, matching column names with post2D
data_2dose <- tibble(x_i=(vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent, n_doses=="2+") %>% pull(age_current)),  # adding in full data from vax_clean
                     y_label=factor((vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent, n_doses=="2+") %>% pull(location)), levels=prov_levels),
                     value_2dose=(vax_clean %>% filter(age_current>=first_agecurrent, age_current<=last_agecurrent, n_doses=="2+") %>% pull(value)))
# filter out posterior draws that do not meet the criterion, using all()
post2D_new <- post2D %>% ungroup() %>% left_join(data_2dose, by=c("x_i","y_label")) %>%
  mutate(value_2dose=ifelse(is.na(value_2dose), 0, value_2dose)) %>%  ##UP TO HERE
  mutate(row_valid=(post2D_constrained >= value_2dose)) %>%
  group_by(draw) %>% filter(all(row_valid))
```

# 2D case model extensions

## Model functions

We now wish to improve our basic model by adding information to the prior and experimenting with choice of covariance function and related parameters. First, define helper functions to enable easy running of the 2D model (saved in "R/"):

-   `compute_prior2D()`
-   `plot_prior2D()`
-   `compute_posterior2D()`
-   `plot_posterior2D()`.

The full 2D Gaussian Process model can then be run with `run_GP(vax_dataset)`. All functions work with multiple province relation datasets.

```{r}
#| label: auto2D

run_GP(vax_clean)
```

From a quick by-eye model selection with GDP province relation, current best GP model parameter values are approximately `l1=1.5, l2=3, b=0.75` for `k=ksqexp`. Plausible range (by-eye) was previously `1 <= l1 < 2`, `l1 < l2 < 3*l1`, and `0.5 <= b <= 1.5` (GDP, provincial data, no measurement error). This hyperparameter space can be searched with`select_GP()` below.

```{r}
#| label: fit_byeye

# for (l1 in seq(1,3,1)) {  # l1>=2 gives meaningless outputs; l=1.5 (or 1) seems reasonable.
#   for (l2 in seq(1,5,0.5)) {  # l2<=2.5 is too strong a relation; l>=3 seems reasonable for log(GBP) data.
#     for (b in seq(0.5,1,0.25)) {  # b>1 gives too much variance; b=0.75 or 1 seems reasonable at present.
#       run_GP(vax_clean, l1=l1, l2=l2, b=b)
#     }
#   }
# }

run_GP(vax_clean,l1=1.5,l2=3,b=0.75)
```

## Running with cNICS data

Note that the provincial estimates collected in `vax_clean` are from multiple sources with varying data collection methodology. As an alternative, the national cNICS study, although perhaps less trustworthy, does have consistent methodology across provinces. We make use of Irena's cleaned, standardised cNICS data and run_GP(). The final plot compares model fit against the full (1+ and 2 dose) provincial dataset `vax_clean`.

Note: Nunavut youngest estimate is described as "treat with caution" in the cNICS online dataset.

Plausible range (by-eye) was previously `1 <= l1 <= 4`, `l1 < l2 <= 50*l1`, and `0.5 <= b <= 1.5` (GDP, cNICS data, no measurement error). This hyperparameter space can be searched with`select_GP()` below.

```{r}
#| label: run_cNICS

run_GP(vax_cNICS, l1=3.5, l2=4, b=1)
```

## Measurement error

Our GP model currently treats the observed data as exact estimates - posterior draws are forced to cross through all observed data points. We hence add measurement error. For this, the posterior mean and covariance matrix are modified when a parameter `meas_error` is specified in `run_GP()`:

```{r}
#| label: meas_error

run_GP(vax_clean,l1=1.5,l2=3,b=0.75)  # exact
run_GP(vax_clean,l1=1.5,l2=3,b=0.75,meas_error=0.01,ndrws=50)  # with measurement error
run_GP(vax_clean,l1=1.5,l2=3,b=0.75,meas_error=0.1,ndrws=50)  # with measurement error
run_GP(vax_clean,l1=1.25,l2=1,b=1.5,meas_error=0.01,ndrws=50)  # with measurement error

```

See the [GitHub Issue #3](https://github.com/phac-modelling-hub/immunity-landscape/issues/3) for a discussion of methodology.

## Model selection and validation

We aim to find the optimal choice of model (hyperparameter values `k`, `l1`, `l2`, `b`, and measurement error `meas_error`), and validate with take-one-out cross-validation. An example algorithm is as follows:

-   For a given covariance function k and lengthscale parameters l1, l2, b:

    1.  Pull out one data point.

    2.  Run GP model and compute e.g. sum squared error between posterior mean and the missing data point.

    3.  Repeat for every data point (e.g. 137 model runs for the 1+ provincial data). Sum errors.

-   Select the optimal k/ l1/ l2/ b combination.

One option is to work with the posterior mean at each missing age as described. A better option is to work with a richer goodness-of-fit metric such as the (log) *posterior predictive density* at each missing age instead. This makes use of the full richness of our Gaussian Process model. We hence choose the latter metric.

`estimate_lppd()` estimates the posterior predictive density from a sample, in a similar way to the algorithm described above - NOW OBSELETE. `compute_lppd()` calculates the full log posterior predictive density analytically. See the [GitHub Issue #4](https://github.com/phac-modelling-hub/immunity-landscape/issues/4) for a discussion of methodology. `select_GP()` then does a search of hyperparameter space and returns the lppd for each hyperparameter set. The 'optimal' model (with max lppd) is also printed to the console.

```{r}
#| label: model_selection_GDP_witherror

#example3 <- compute_lppd(vax_clean, prov_values="GDP", k=ksqexp, l1=1.5, l2=3, b=1, meas_error=0.1)
#example4 <- select_GP(vax_clean, prov_values="GDP", k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1), meas_error=0.1)

#' exhaustive search of reasonable hyperparameter space (provincial dataset, GDP) with error
GPcombinations_GDP2 <- select_GP(vax_clean, prov_values="GDP", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_GDP2 %>% readr::write_csv(here::here("results", "GPcombinations_GDP_witherror.csv"))

#' example lppd fit
run_GP(vax_clean,l1=2.5,l2=250,b=0.5,meas_error=0.1,ndrws=50)
```

We can also explore if there are local maximums in the model selection, rather than only considering the model with largest lppd.

```{r}
#| label: model_selection_heatmaps

my_tibble_GDP2 <- read_csv(here::here("results", "GPcombinations_GDP_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_LI2 <- read_csv(here::here("results", "GPcombinations_lowincome_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_VH2 <- read_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_VH2exp <- read_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv")) %>% filter(k_name == "kexp" & b == 1)
my_tibble_GI <- read_csv(here::here("results", "GPcombinations_Gini_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)

#' plot l1 against l2, colouring by lppd, for several meas_error values
my_tibble_GDP2 %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -200, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -200") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for GDP model, varying measurement error",
       x = "l1", y = "l2")

my_tibble_LI2 %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -200, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -200") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for Low-Income model, varying measurement error",
       x = "l1", y = "l2")

my_tibble_VH2 %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -200, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -200") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for Vaccine Hesitancy model (k = sqexp), varying measurement error",
       x = "l1", y = "l2")

my_tibble_VH2exp %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -200, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -200") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for Vaccine Hesitancy model (k = exp), varying measurement error",
       x = "l1", y = "l2")

my_tibble_GI %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -200, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -200") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for Gini model, varying measurement error",
       x = "l1", y = "l2")
```

## Relation of provinces

Relation of provinces is encoded in `prov_values` and can be extracted from the function `extract_province_relation()`. This function has a number of socioecomonic datasets preloaded in a standardised format for use in the GP model. Examples include "GDP" and "low_income_families". \*Note that for "low_income_families" there is currently no data for NU, NT, and YT.

```{r}
#| label: province_relation_lowincome

prov_valuesLI <- extract_province_relation("low_income_families", vax_dataset=vax_cleanLI)  # vac_cleanLI is defined in set_up at beginning of file

example <- select_GP(vax_cleanLI, prov_values="low_income_families", k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1))

#' exhaustive search with error of reasonable hyperparameter space (provincial dataset, children in low-income families province relation metric)
GPcombinations_LI2 <- select_GP(vax_cleanLI, prov_values="low_income_families", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_LI2 %>% readr::write_csv(here::here("results", "GPcombinations_lowincome_witherror.csv"))

# #' plots of final model (using best parameter values)
# run_GP(vax_clean, prov_values="GDP", l1=2.5, l2=250, b=0.5, meas_error=0.1)
# ggsave(paste0("plots/bestGPmodel_GDP",".jpeg"), height=4.51, width=7.29)
# run_GP(vax_cleanLI, prov_values="low_income_families", l1=2.5, l2=25, b=0.5, meas_error=0.1)
# ggsave(paste0("plots/bestGPmodel_lowincome",".jpeg"), height=4.51, width=7.29)
# run_GP(vax_cleanVH, prov_values="vaccine_hesitancy", l1=2.5, l2=6.0, b=0.5, meas_error=0.1)
# ggsave(paste0("plots/bestGPmodel_vaccinehesitancy",".jpeg"), height=4.51, width=7.29)
```

A new dataset on vaccine hesitancy requires some manipulation to use (condensing coverage data for Atlantic provinces and Northern provinces):

```{r}
#| label: province_relation_hesitancy

#' use vax_cleanVH which combines smaller provinces into regions (from set_up code snippet at beginning of file)
head(vax_cleanVH)

#' run with vax_cleanVH - examples
run_GP(vax_cleanVH, prov_values="vaccine_hesitancy", l1=2.5, l2=4, b=0.5, meas_error=0.1)
run_GP(vax_cleanVH, prov_values="vaccine_hesitancy", l1=2.5, l2=0.5, b=0.5, meas_error=0)

#' exhaustive search with error of reasonable hyperparameter space (provincial dataset, vaccine hesitancy province relation metric)
GPcombinations_VH2 <- select_GP(vax_cleanVH, prov_values="vaccine_hesitancy", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_VH2 %>% readr::write_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv"))
```

We also wish to add a new indicator: Gini index. This runs on all provinces.

```{r}
#| label: province_relation_gini

#' example model
run_GP(vax_clean, prov_values="Gini", l1=2.5, l2=10, b=0.5, meas_error=0.1)

#' exhaustive search of reasonable hyperparameter space with error
GPcombinations_GI <- select_GP(vax_clean, prov_values="Gini", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_GI %>% readr::write_csv(here::here("results", "GPcombinations_Gini_witherror.csv"))
```

## Assumptions tests

A Gaussian Process model assumes relation of the independent variable(s), i.e. that 'related' ages (defined however we wish with lengthscale parameter l1) will have similar vaccination coverage and that 'related' provinces will also have similar coverage.

To visualise relation of provinces, we can plot `province_relation_data` on an axis. One could compare to for example the mean cNICS coverage estimate for each province. A range of plots are shown.

```{r}
#| label: assumption_test_provinces

#' load province relation values as tibbles
prov_valuesGDP <- enframe(extract_province_relation("GDP", vax_dataset=vax_clean), name="location", value="prov_valuesGDP")  # GDP
prov_valuesLI <- enframe(extract_province_relation("low_income_families", vax_dataset=vax_cleanLI), name="location", value="prov_valuesLI")  # low_income
prov_valuesVH <- enframe(extract_province_relation("vaccine_hesitancy", vax_dataset=vax_cleanVH), name="location", value="prov_valuesVH")  # vaccine_hesitancy
prov_valuesGI <- enframe(extract_province_relation("Gini", vax_dataset=vax_clean), name="location", value="prov_valuesGI")  # Gini
ARval <- prov_valuesVH %>% filter(location=="Atlantic region") %>% pull(prov_valuesVH)
NRval <- prov_valuesVH %>% filter(location=="Northern region") %>% pull(prov_valuesVH)
prov_valuesVH <- prov_valuesVH %>% rbind(c("Newfoundland and Labrador", ARval)) %>% rbind(c("Prince Edward Island", ARval)) %>%  # this is very crude code!
  rbind(c("Nova Scotia", ARval)) %>% rbind(c("Nunavut", NRval)) %>% rbind(c("Northwest Territories", NRval)) %>% mutate(prov_valuesVH=as.numeric(prov_valuesVH))

#' load cNICS as tibbles
cNICS_2021 <- vax_cNICS %>% filter(year_report==2021) %>% select(location, "cNICS_2021"=value) %>% filter(location!="Nunavut") %>%
  mutate(cNICS_2021=cNICS_2021*100)  # removing low NU point
cNICS_2017 <- vax_cNICS %>% filter(year_report==2017) %>% select(location, "cNICS_2017"=value) %>% mutate(cNICS_2017=cNICS_2017*100)

#' combine into one
provinces_test <- prov_valuesGDP %>% left_join(prov_valuesLI, by="location") %>% left_join(prov_valuesVH, by="location") %>% left_join(prov_valuesGI, by="location") %>%
  left_join(cNICS_2021, by="location") %>% left_join(cNICS_2017, by="location") %>% 
  mutate(cNICS_mean=(cNICS_2021 + cNICS_2017)/2)

#' plot
library(ggrepel)
ggplot(provinces_test, aes(x = prov_valuesGDP, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  ylim(85,100) + labs(title="Comparing province relation datasets to coverage", x="log(GDP)", y = "vaccine coverage, cNICS 2021 (%)")
ggplot(provinces_test, aes(x = prov_valuesLI, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  xlim(85,100) + ylim(85,100) + labs(title="Comparing province relation datasets to coverage", x="% of children not in low-income families", y = "vaccine coverage, cNICS 2021 (%)")
ggplot(provinces_test, aes(x = prov_valuesVH, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  xlim(70,100) + ylim(85,100) + labs(title="Comparing province relation datasets to coverage", x="% of parents who are not vaccine hesitant", y = "vaccine coverage, cNICS 2021 (%)")
ggplot(provinces_test, aes(x = prov_valuesGI, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  ylim(85,100) + labs(title="Comparing province relation datasets to coverage", x="Gini index", y = "vaccine coverage, cNICS 2021 (%)")

#' instead consider provincial data mean over cNICS
vax_means_noaNICS <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned_noadults.csv"), show_col_types = FALSE) %>% 
  filter(n_doses %in% c("1", "1+")) %>% group_by(location) %>% summarise(average = mean(value)) %>% select(location, "vax_means_noaNICS"=average)
provinces_test <- provinces_test %>% left_join(vax_means_noaNICS, by="location") %>% mutate(vax_means_noaNICS=vax_means_noaNICS*100)
ggplot(provinces_test, aes(x = prov_valuesGDP, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  ylim(79,100) + labs(title="Comparing province relation datasets to coverage", x="log(GDP)", y = "vaccine coverage, provincial data mean (%)")
ggplot(provinces_test, aes(x = prov_valuesLI, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  xlim(79,100) + ylim(79,100) + labs(title="Comparing province relation datasets to coverage", x="% of children not in low-income families", y = "vaccine coverage, provincial data mean (%)")
ggplot(provinces_test, aes(x = prov_valuesVH, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  xlim(70,100) + ylim(70,100) + labs(title="Comparing province relation datasets to coverage", x="% of parents who are not vaccine hesitant", y = "vaccine coverage, provincial data mean (%)")
ggplot(provinces_test, aes(x = prov_valuesGI, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4) +
  ylim(79,100) + labs(title="Comparing province relation datasets to coverage", x="Gini index", y = "vaccine coverage, provincial data mean (%)")

#' for fun, plot provincial data against cNICS
#"C:/Users/lh667/AppData/Local/Packages/MicrosoftWindows.Client.Core_cw5n1h2txyewy/TempState/ScreenClip/{FC770BFB-663B-4B30-A70C-5B5DC4D78C9F}.png"ggplot(provinces_test, aes(x = vax_means_noaNICS, y = cNICS_mean)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
```

To test relation of ages, we can compute pairwise differences in absolute coverage for all data points and observe if pairs of adjacent ages have a lower difference on average:

```{r}
#| label: assumption_test_ages

first_agecurrent <- 5
last_agecurrent <- 21

#' 1+ dose provincial data, no aNICS
#' read in provincial data with aNICS removed
vax_dataset1 <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned_noadults.csv"), show_col_types = FALSE) %>%
  filter(n_doses %in% c("1", "1+")) %>% filter(age_current >= first_agecurrent, age_current <= last_agecurrent)

#' define a function to compute all pairwise differences
compute_pairwise_diffs <- function(df) {
  df <- df %>% arrange(age_current)
  
  expand_grid(age1 = df$age_current, age2 = df$age_current) %>% filter(age1 < age2) %>%
    left_join(df %>% select(age1 = age_current, cov1 = value), by = "age1") %>%
    left_join(df %>% select(age2 = age_current, cov2 = value), by = "age2") %>%
    mutate(diff = abs(cov1 - cov2), adjacent = (age2 - age1 == 1),
           age_gap = age2 - age1, gap_group = case_when(age_gap == 1 ~ "1 year apart",
                                                        age_gap == 2 ~ "2 years apart",
                                                        age_gap == 3 ~ "3 years apart",
                                                        age_gap > 3 ~ ">3 years apart"))
}

#' compute pairwise for e.g. Alberta
alberta_pairs <- vax_dataset1 %>% filter(location=="Alberta") %>% compute_pairwise_diffs()

#' plot histogram for e.g. Alberta
ggplot(alberta_pairs, aes(x = diff*100, fill = gap_group)) + geom_histogram() +
  labs(title = "Pairwise Coverage Differences — Alberta (1+ dose)",
    x = "Absolute Difference in Coverage (%)",
    y = "Count",
    fill = "Adjacent ages")

#' compute pairwise and plot histogram for remaining provinces
province_pairs <- vax_dataset1 %>% group_by(location) %>% group_modify(~ compute_pairwise_diffs(.x))
ggplot(province_pairs, aes(x = diff*100, fill = gap_group)) + geom_histogram() +  # use fill=adjacent for only adjacent ages OR fill=gap_group otherwise
  facet_wrap(~ location) + labs(title = "Pairwise Coverage Differences — By Province (1+ dose)",
                                x = "Absolute Difference in Coverage (%)", y = "Count", fill = "Age gap")
ggsave(paste0("plots/testing_age_assumption1",".pdf"), height=4.51, width=7.29)  # example code to save plot

#' 2 dose provincial data
#' repeat steps as above
vax_dataset2 <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned_noadults.csv"), show_col_types = FALSE) %>%
  filter(n_doses %in% c("2", "2+"))
province_pairs2 <- vax_dataset2 %>% group_by(location) %>% group_modify(~ compute_pairwise_diffs(.x))
ggplot(province_pairs2, aes(x = diff*100, fill = gap_group)) + geom_histogram() + 
  facet_wrap(~ location) + labs(title = "Pairwise Coverage Differences — By Province (2+ dose)",
                                x = "Absolute Difference in Coverage (%)", y = "Count", fill = "Age gap")
ggsave(paste0("plots/testing_age_assumption2",".pdf"), height=4.51, width=7.29)  # example code to save plot

#' Aggregated provincial data
vax_dataset3 <- vax_dataset1 %>% mutate(id = row_number()) %>% select(-location)  # remove location label
country_pairs <- expand_grid(id1 = vax_dataset3$id, id2 = vax_dataset3$id) %>%
  filter(id1 < id2) %>%
  left_join(vax_dataset3 %>% select(id1 = id, age1 = age_current, cov1 = value), by = "id1") %>%
  left_join(vax_dataset3 %>% select(id2 = id, age2 = age_current, cov2 = value), by = "id2") %>%
  mutate(diff = abs(cov1 - cov2), adjacent = (abs(age2 - age1) == 1),
         age_gap = age2 - age1, gap_group = case_when(age_gap == 1 ~ "1 year apart",
                                                      age_gap == 2 ~ "2 years apart",
                                                      age_gap == 3 ~ "3 years apart",
                                                      age_gap > 3 ~ ">3 years apart",
                                                      age_gap == 0 ~ "0 years apart"))
ggplot(country_pairs, aes(x = diff*100, fill = gap_group)) + geom_histogram() + 
  labs(title = "Pairwise Coverage Differences: All provincial data",
  x = "Absolute Difference in Coverage (%)", y = "Count", fill = "Age gap") +
  theme(axis.text=element_text(size=18), axis.title=element_text(size=20), plot.title = element_text(size=25))
ggsave(paste0("plots/testing_age_assumption3",".jpeg"), height=4.51, width=7.29)  # example code to save plot


vax_dataset4 <- vax_dataset2 %>% mutate(id = row_number()) %>% select(-location)
country_pairs2 <- expand_grid(id1 = vax_dataset4$id, id2 = vax_dataset4$id) %>%
  filter(id1 < id2) %>%
  left_join(vax_dataset4 %>% select(id1 = id, age1 = age_current, cov1 = value), by = "id1") %>%
  left_join(vax_dataset4 %>% select(id2 = id, age2 = age_current, cov2 = value), by = "id2") %>%
  mutate(diff = abs(cov1 - cov2), adjacent = (abs(age2 - age1) == 1),
          age_gap = age2 - age1, gap_group = case_when(age_gap == 1 ~ "1 year apart",
                                                       age_gap == 2 ~ "2 years apart",
                                                       age_gap == 3 ~ "3 years apart",
                                                       age_gap > 3 ~ ">3 years apart",
                                                       age_gap == 0 ~ "0 years apart"))
ggplot(country_pairs2, aes(x = diff*100, fill = gap_group)) + geom_histogram() + 
  labs(title = "Pairwise Coverage Differences — All provincial data (2 dose)")
```

We see a good relationship between coverage at adjacent ages, both in the 1+ and 2+ dose data.

## UK comparison

We can plot UK vaccine coverage data, and run through the model, as a comparison. Data are for 2013-2025, at least one measles dose, aged 5 years. `vax_uk` gives vaccine coverage data for the four nations of the UK (England, Wales, Scotland, Northern Ireland). `vax_england` gives the 10 regions of England (e.g. London, South West, South East etc.). All vaccine coverage data is taken from the UKHSA dashboard: https://ukhsa-dashboard.data.gov.uk/cover/childhood-vaccinations.

```{r}
#| label: uk

#' UK data
vax_UK <- read_csv("data/ukhsa-chart-download-mmr1-nations.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (England, Wales, Scotland, NI)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)
vax_england <- read_csv("data/ukhsa-chart-download-mmr1-regions.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (regions of England)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)

vax_UK %>% ggplot() + geom_point(aes(x=age_current, y=value), col="red") +
  scale_x_continuous(name="current age x_i", limits=c(5,21)) + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ location)
vax_england %>% ggplot() + geom_point(aes(x=age_current, y=value), col="red") +
  scale_x_continuous(name="current age x_i", limits=c(5,21)) + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ location)

#' UK model
run_GP(vax_UK, prov_values = "UK-GDP")
run_GP(vax_UK, prov_values = "UK-low_income_families")
run_GP(vax_england, prov_values = "UK-GDP")
run_GP(vax_england, prov_values = "UK-low_income_families")
run_GP(vax_england, prov_values = "UK-Gini")

#' UK hyperparams
GPcombinations_UKGDP <- select_GP(vax_UK, prov_values="UK-GDP", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_UKGDP %>% readr::write_csv(here::here("results", "GPcombinations_UKGDP_witherror.csv"))

GPcombinations_UKLI <- select_GP(vax_UK, prov_values="UK-low_income_families", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_UKLI %>% readr::write_csv(here::here("results", "GPcombinations_UKLI_witherror.csv"))

GPcombinations_englandGDP <- select_GP(vax_england, prov_values="UK-GDP", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_englandGDP %>% readr::write_csv(here::here("results", "GPcombinations_englandGDP_witherror.csv"))

GPcombinations_englandLI <- select_GP(vax_england, prov_values="UK-low_income_families", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_englandLI %>% readr::write_csv(here::here("results", "GPcombinations_englandLI_witherror.csv"))

GPcombinations_englandGini <- select_GP(vax_england, prov_values="UK-Gini", k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_englandGini %>% readr::write_csv(here::here("results", "GPcombinations_englandGini_witherror.csv"))

#' UK lppd heatmaps
my_tibble_a <- read_csv(here::here("results", "GPcombinations_UKGDP_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_b <- read_csv(here::here("results", "GPcombinations_UKLI_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_c <- read_csv(here::here("results", "GPcombinations_englandGDP_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_d <- read_csv(here::here("results", "GPcombinations_englandLI_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_e <- read_csv(here::here("results", "GPcombinations_englandGini_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)

my_tibble_a %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%  # ignoring poor model fits
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -100") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for UK GDP model, varying measurement error", x = "l1", y = "l2")
ggsave(paste0("plots/lppdheatmap_UKGDP",".jpeg"), height=4.51, width=7.29)

my_tibble_b %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -100") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for UK Low-Income model, varying measurement error", x = "l1", y = "l2")
ggsave(paste0("plots/lppdheatmap_UKLI",".jpeg"), height=4.51, width=7.29)

my_tibble_c %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -100") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for England GDP model, varying measurement error", x = "l1", y = "l2")
ggsave(paste0("plots/lppdheatmap_englandGDP",".jpeg"), height=4.51, width=7.29)

my_tibble_d %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -100") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for England Low-Income model, varying measurement error", x = "l1", y = "l2")
ggsave(paste0("plots/lppdheatmap_englandLI",".jpeg"), height=4.51, width=7.29)

my_tibble_e %>%
  mutate(l1 = factor(l1), l2 = factor(l2)) %>%
  mutate(lppd_capped = if_else(lppd_exact < -100, NA_real_, lppd_exact)) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_capped)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red", na.value="white", name = "lppd > -100") +
  facet_wrap(~ meas_error) +
    #ylim(0,101) +
  labs(title = "lppd heatmaps for England Gini model, varying measurement error", x = "l1", y = "l2")
ggsave(paste0("plots/lppdheatmap_englandGini",".jpeg"), height=4.51, width=7.29)

#' Best model outputs
run_GP(vax_england, prov_values = "UK-Gini", k=ksqexp, l1=2.25, l2=8, b=0.5, meas_error=0.05)
run_GP(vax_england, prov_values = "UK-low_income_families", k=ksqexp, l1=2.25, l2=3.5, b=0.5, meas_error=0.05)
```

## Sensitivity analysis 1 - prior means

One sensitivity analysis consideration is in exploring the choice of prior mean. Our main model uses a prior mean equivalent to the mean of the observed multi-province dataset (\~90%). \[Technically our main model assumes a global prior mean of zero; each data point is (transformed and) centred by subtracting the mean of the observed data (in logit space)\]. We wish to instead set a different prior mean for each province.

```{r}
#| label: sens_anal_mean

```

## Leave-one-province-out (LOPO-CV)

The function `compute_lppd_lopo` computes lppd for one set of model params from leave-one-province-out cross-validation (same as LOO-CV but leaving out the data points for one whole province at a time). Model selection (`select_GP()`) also computes this for each set of model params and stores the results as a column named `lppd_LOPO`.

```{r}
#| label: lopo

#' example with diagnostics plot
example <- compute_lppd_lopo(vax_clean, prov_values="GDP", k=ksqexp, l1=1.5, l2=0.2, b=1)
example
prov_values <- extract_province_relation("GDP", vax_clean)
plot(x=prov_values[names(example$mean_lppd_by_province)], y=example$mean_lppd_by_province, xlab = "province value e.g. GDP", ylab = "Mean LOPO lppd")
```

The above diagnostic plot allows us to explore the relative contribution of each province to the total log posterior predictive density (after scaling by \# of observations). For the GDP model framework, we see that for poor models, there is a large contribution from the ?tail-end? provinces to total lppd, whereas with our best models (largest LOO lppd / large l2), the diagnostic plot is fairly flat indicating that each province contributes equally (after scaling by \# of observations) and the model essentially collapses to an age model.

```{r}
#| label: lopo_investigation1

#' GDP model framework
#' explore province predictability (LOPO lppd) for one model e.g. best model from model selection
model_2087 <- compute_lppd_lopo(vax_clean, prov_values="GDP", k=ksqexp, l1=2.5, l2=250, b=0.5, meas_error=0.1)
prov_values <- extract_province_relation("GDP", vax_dataset=vax_clean)
plot(x=prov_values[names(model_2087$mean_lppd_by_province)], y=model_2087$mean_lppd_by_province, xlab = "province, by e.g. GDP value", ylab = "Mean LOPO lppd", main = "Province predictability for model_2087 (best)")
max(model_2087$mean_lppd_by_province) - min(model_2087$mean_lppd_by_province)  # small difference == age model
#' compare to a poor model
model_1515 <- compute_lppd_lopo(vax_clean, prov_values="GDP", k=ksqexp, l1=2.25, l2=0.2, b=1.5, meas_error=0)
plot(x=prov_values[names(model_1515$mean_lppd_by_province)], y=model_1515$mean_lppd_by_province, xlab = "province, by e.g. GDP value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1515 (worst)")
max(model_1515$mean_lppd_by_province) - min(model_1515$mean_lppd_by_province)  # large difference == provinces contribute differently == relation between provinces


#' LI model framework
#' explore province predictability (LOPO lppd) for one model e.g. best model from model selection
model_1982 <- compute_lppd_lopo(vax_cleanLI, prov_values="low_income_families", k=ksqexp, l1=2.5, l2=25, b=0.5, meas_error=0.1)
prov_values <- extract_province_relation("low_income_families", vax_dataset=vax_cleanLI)
plot(x=prov_values[names(model_1982$mean_lppd_by_province)], y=model_1982$mean_lppd_by_province, xlab = "province, by e.g. LI value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1982 (best)")
max(model_1982$mean_lppd_by_province) - min(model_1982$mean_lppd_by_province)  # small difference == age model
#' compare to a poor model
model_1515 <- compute_lppd_lopo(vax_cleanLI, prov_values="low_income_families", k=ksqexp, l1=2.25, l2=0.2, b=1.5, meas_error=0)
plot(x=prov_values[names(model_1515$mean_lppd_by_province)], y=model_1515$mean_lppd_by_province, xlab = "province, by e.g. LI value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1515 (worst)")
max(model_1515$mean_lppd_by_province) - min(model_1515$mean_lppd_by_province)  # large difference == provinces contribute differently == relation between provinces


#' VH model framework
#' explore province predictability (LOPO lppd) for one model e.g. best model from model selection
model_1937 <- compute_lppd_lopo(vax_cleanVH, prov_values="vaccine_hesitancy", k=ksqexp, l1=2.5, l2=6, b=0.5, meas_error=0.1)
prov_values <- extract_province_relation("vaccine_hesitancy", vax_dataset=vax_cleanVH)
plot(x=prov_values[names(model_1937$mean_lppd_by_province)], y=model_1937$mean_lppd_by_province, xlab = "province, by e.g. VH value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1937 (best)")
max(model_1937$mean_lppd_by_province) - min(model_1937$mean_lppd_by_province)  # small difference == age model
#' compare to a poor model
model_1215 <- compute_lppd_lopo(vax_cleanVH, prov_values="vaccine_hesitancy", k=ksqexp, l1=2, l2=0.1, b=1.5, meas_error=0)
plot(x=prov_values[names(model_1215$mean_lppd_by_province)], y=model_1215$mean_lppd_by_province, xlab = "province, by e.g. VH value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1215 (worst)")
max(model_1215$mean_lppd_by_province) - min(model_1215$mean_lppd_by_province)  # large difference == provinces contribute differently == relation between provinces


#' Gini model framework
#' explore province predictability (LOPO lppd) for one model e.g. best model from model selection
model_2072 <- compute_lppd_lopo(vax_clean, prov_values="Gini", k=ksqexp, l1=2.5, l2=10, b=0.5, meas_error=0.1)
prov_values <- extract_province_relation("Gini", vax_dataset=vax_clean)
plot(x=prov_values[names(model_2072$mean_lppd_by_province)], y=model_2072$mean_lppd_by_province, xlab = "province, by e.g. Gini value", ylab = "Mean LOPO lppd", main = "Province predictability for model_2072 (best)")
max(model_2072$mean_lppd_by_province) - min(model_2072$mean_lppd_by_province)  # small difference == age model
#' compare to a poor model
model_1895 <- compute_lppd_lopo(vax_clean, prov_values="Gini", k=ksqexp, l1=2.5, l2=0.1, b=0.5, meas_error=0)
plot(x=prov_values[names(model_1895$mean_lppd_by_province)], y=model_1895$mean_lppd_by_province, xlab = "province, by e.g. Gini value", ylab = "Mean LOPO lppd", main = "Province predictability for model_1895 (worst)")
max(model_1895$mean_lppd_by_province) - min(model_1895$mean_lppd_by_province)  # large difference == provinces contribute differently == relation between provinces
```

We can also explore the difference `delta` between LOO lppd and LOPO lppd. Similar values indicate weak province predictability ie. that provinces are independent and our model is essentially behaving as an age model. From the plots, we see that this happens with large l2.

```{r}
#| label: lopo_investigation2

#' GDP model framework
#' load model selection results and plot difference between LOPO and LOO lppd's
delta_plot <- read_csv(here::here("results", "GPcombinations_GDP_witherror.csv"))  # load a model selection csv file
delta_plot <- delta_plot %>% arrange(l2) %>%  # sort by increasing l2
  mutate(delta=lppd_LOPO - lppd_exact)  # compute difference between LOPO lppd and LOO lppd
delta_plot %>% ggplot(aes(x = l2, y = delta)) + geom_point() + ylim(-5000, 5000) + ylab("LOPO - LOO") + # plot delta for increasing l2
 xlim(0,250)


#' LI model framework
#' load model selection results and plot difference between LOPO and LOO lppd's
delta_plot2 <- read_csv(here::here("results", "GPcombinations_lowincome_witherror.csv"))  # load a model selection csv file
delta_plot2 <- delta_plot2 %>% arrange(l2) %>%  # sort by increasing l2
  mutate(delta=lppd_LOPO - lppd_exact)  # compute difference between LOPO lppd and LOO lppd
delta_plot2 %>% ggplot(aes(x = l2, y = delta)) + geom_point() + ylim(-5000, 5000) + ylab("LOPO - LOO") + # plot delta for increasing l2
 xlim(0,250)


#' VH model framework
#' load model selection results and plot difference between LOPO and LOO lppd's
delta_plot3 <- read_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv"))  # load a model selection csv file
delta_plot3 <- delta_plot3 %>% arrange(l2) %>%  # sort by increasing l2
  mutate(delta=lppd_LOPO - lppd_exact)  # compute difference between LOPO lppd and LOO lppd
delta_plot3 %>% ggplot(aes(x = l2, y = delta)) + geom_point() + ylim(-5000, 5000) + ylab("LOPO - LOO") + # plot delta for increasing l2
  xlim(0,250)

#' Gini model framework
#' load model selection results and plot difference between LOPO and LOO lppd's
delta_plot4 <- read_csv(here::here("results", "GPcombinations_Gini_witherror.csv"))  # load a model selection csv file
delta_plot4 <- delta_plot4 %>% arrange(l2) %>%  # sort by increasing l2
  mutate(delta=lppd_LOPO - lppd_exact)  # compute difference between LOPO lppd and LOO lppd
delta_plot4 %>% ggplot(aes(x = l2, y = delta)) + geom_point() + ylim(-5000, 5000) + ylab("LOPO - LOO") + # plot delta for increasing l2
  xlim(0,250)


#' An alternative visualisation is to look at the difference between LOO and LOPO for each best model:
best_gdp <- delta_plot %>% filter(lppd_exact==max(lppd_exact))
best_li <- delta_plot2 %>% filter(lppd_exact==max(lppd_exact))
best_vh <- delta_plot3 %>% filter(lppd_exact==max(lppd_exact))
best_gini <- delta_plot4 %>% filter(lppd_exact==max(lppd_exact))
```

This leads to the question: do any of our modelling frameworks utilise relation between provinces/regions? Consider the England LI model. From model selection, this framework does not favour largest l2 (best LOO lppd currently gives l2=3.5, compared to the plausible upper limit of 25). Repeat the LOPO analysis for England LI. Take code from uk code snippet above.

```{r}
#| label: lopo_investigation3

#' England LI model framework
#' load England vaccine coverage data
vax_england <- read_csv("data/ukhsa-chart-download-mmr1-regions.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (regions of England)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)

#' explore province predictability (LOPO lppd) for one model e.g. best model from model selection
model_1698 <- compute_lppd_lopo(vax_england, prov_values="UK-low_income_families", k=ksqexp, l1=2.25, l2=3.5, b=0.5, meas_error=0.05)
prov_values <- extract_province_relation("UK-low_income_families", vax_dataset=vax_england)
plot(x=prov_values[names(model_1698$mean_lppd_by_province)], y=model_1698$mean_lppd_by_province, xlab = "region of England, by e.g. LI value", ylab = "Mean LOPO lppd", main = "Region predictability for model_1698 (best)")
max(model_1698$mean_lppd_by_province) - min(model_1698$mean_lppd_by_province)  # small difference == age model
#' compare to a poor model
model_950 <- compute_lppd_lopo(vax_england, prov_values="UK-low_income_families", k=ksqexp, l1=1.75, l2=0.1, b=0.5, meas_error=0)
plot(x=prov_values[names(model_950$mean_lppd_by_province)], y=model_950$mean_lppd_by_province, xlab = "region of England, by e.g. LI value", ylab = "Mean LOPO lppd", main = "Region predictability for model_950 (worst)")
max(model_950$mean_lppd_by_province) - min(model_950$mean_lppd_by_province)  # large difference == provinces contribute differently == relation between provinces

#' load model selection results and plot difference between LOPO and LOO lppd's
delta_plot4 <- read_csv(here::here("results", "GPcombinations_englandLI_witherror.csv"))  # load a model selection csv file
delta_plot4 <- delta_plot4 %>% arrange(l2) %>%  # sort by increasing l2
  mutate(delta=lppd_LOPO - lppd_exact)  # compute difference between LOPO lppd and LOO lppd
delta_plot4 %>% ggplot(aes(x = l2, y = delta)) + geom_point() + ylim(-5000, 5000) + ylab("LOPO - LOO") + # plot delta for increasing l2
 scale_x_continuous(breaks = seq(0, max(delta_plot4$l2), by=50))

```

We then had further discussions into these LOPO investigations and came up with a fourth more-interpretive analysis: Really we are interested in how each province compares to being left out of the age+province model, compared to just an age model. Do provinces matter?

For this, we wish to compare lppd for one province in the best age+province model case (best l2) to just the age model (e.g. l2=500). The function `compute_lppd_lopo` already outputs the lppd for each province so we utilise this.

```{r}
#| label: lopo_investigation4

# Consider each model framework in turn
# model configurations
l2_lim <- list(Gini = c(0.1, 50), LI = c(0.2, 25), VH = c(0.1,  8))
model_configs <- list(Gini = list(csv="GPcombinations_Gini_witherror.csv", prov_values="Gini", l2_lim=l2_lim$Gini, title_lab="Gini",
                                  vax_dataset=vax_clean),
                      LI = list(csv="GPcombinations_lowincome_witherror.csv", prov_values="low_income_families", l2_lim=l2_lim$LI, title_lab="LI",
                                vax_dataset=vax_cleanLI),
                      VH = list(csv="GPcombinations_vaccinehesitancy_witherror.csv", prov_values="vaccine_hesitancy", l2_lim=l2_lim$VH, 
                                title_lab="VH", vax_dataset=vax_cleanVH))

lopo_comparisons <- list()
prop_summaries   <- list()
p_diffs          <- list()

# For each model framework:
for (model_name in names(model_configs)) {
  cfg <- model_configs[[model_name]]
  
  # extract parameter values for best model
  best_model <- read_csv(here::here("results", cfg$csv)) %>% filter(l2 <= cfg$l2_lim[2], l2 >= cfg$l2_lim[1]) %>% slice_max(lppd_exact, n=1)
  k         <- get(best_model$k_name)
  l1        <- best_model$l1
  l2        <- best_model$l2
  b         <- best_model$b
  meas_error <- best_model$meas_error
  
  # compute lppd lopo by province for best model and age-only model
  best_model_lopo <- compute_lppd_lopo(cfg$vax_dataset, prov_values=cfg$prov_values, k=k, l1=l1, l2=l2, b=b, meas_error=meas_error)
  age_model_lopo <- compute_lppd_lopo(cfg$vax_dataset, prov_values=cfg$prov_values, k=k, l1=l1, l2=500, b=b, meas_error=meas_error)
  
  # compare for each province
  lopo_comparisons[[model_name]] <- tibble(province = names(best_model_lopo$mean_lppd_by_province), best_model = best_model_lopo$mean_lppd_by_province,
    age_model  = age_model_lopo$mean_lppd_by_province, diff = best_model_lopo$mean_lppd_by_province - age_model_lopo$mean_lppd_by_province) %>%
    left_join(pt_lookup(), by = c("province" = "location"))
  
  # proportion positive vs negative
  prop_summaries[[model_name]] <- lopo_comparisons[[model_name]] %>% summarise(prop_best_wins = mean(diff > 0), prop_age_wins = mean(diff < 0))
  print(model_name); print(prop_summaries[[model_name]])
  
  # plot difference (best model − age model) by province
  p_diffs[[model_name]] <- ggplot(lopo_comparisons[[model_name]], aes(x = pt, y = diff, fill = diff > 0)) + geom_col() +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
    scale_fill_manual(values = c("TRUE" = "#2166ac", "FALSE" = "#d73027"), labels = c("TRUE" = "Full model preferred", "FALSE" = "Age model preferred"),
                      name = NULL) +
    labs(x = "Province", y = "Difference in mean LPPD per province", title = paste0("Relative LOPO performance (", cfg$title_lab, " model)")) +
    ylim(-0.75,0.75) + theme_minimal()
}

# view plots
p_diffs$Gini
p_diffs$LI
p_diffs$VH
```

## Simulation-based calibration (SBC)

For each iteration i, the below code generates a synthetic data set with known hyperparameters and then runs the model selection process to see how well the true hyperparameters can be recovered. The data generation process (Step 1) is as follows: choose values for k, l1, l2, b, and meas_error; compute a GP for this parameter combination and calculate the posteriors; save one random posterior draw and treat this as the synthetic data.

```{r}
#| label: sbc

SBC_results_list <- vector("list", 100)
k_select <- function(a) {  # quick function to randomly choose ksqexp or kexp
  if (a==1) {
    list(fn=ksqexp, name="ksqexp")
  } else if (a==0) {
    list(fn=kexp, name="kexp")
  }
}

for (i in 1:50) {
  #' Step 1: Create synthetic data by randomly drawing one posterior from a GP with known hyperparams
  #' define hyperparams for this iteration
  k_object <- k_select(sample(0:1, 1))
  k <- k_object$fn
  l1 <- sample(c(1,1.25,1.5,1.75,2,2.25,2.5), 1)  #runif(1, 1, 2.5)
  l2 <- sample(c(0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25), 1)  #runif(1, 0.2, 25)
  b <- sample(c(0.5,1,1.5), 1)  #runif(1, 0.5, 1.5)
  meas_error <- sample(c(0.5,0.1,0.05,0.01,0), 1)  #runif(1, 0, 0.5)
  params <- list(k=k, k_name=k_object$name, l1=l1, l2=l2, b=b, meas_error=meas_error)  # save param values
  #' run model
  post <- run_GP(vax_cleanLI, prov_values="low_income_families", k=kexp, l1=l1, l2=l2, 
                 b=b, meas_error=meas_error, show_plots=F)
  #' randomly save one posterior draw as the data
  a <- sample(1:50, 1)
  synthetic_data <- post %>% filter(draw == a) %>% select(-draw, -post2D, -y_i)
  #' reformat
  synthetic_data <- synthetic_data %>% mutate(location=as.character(y_label)) %>%
    rename(age_current=x_i, value=post2D_constrained) %>% select(-y_label) %>%
    mutate(n_doses="1+")
  # remove x% missing data
  n <- as.numeric(count(vax_cleanLI %>% filter((age_current>=5) & (age_current<=21) & n_doses!="2+")))  # vax_cleanLI has 34/168 data points
  synthetic_data_partial <- synthetic_data %>% mutate(id=row_number()) %>%
    filter(age_current > 4) %>%  # no data for under 5 year olds
    filter(id %in% sample(min(id):max(id), n, replace=F))  # randomly remove rows leaving n points

  #' Step 2: Run model selection process with the synthetic data
  synthetic_modelselection_LI <- select_GP(synthetic_data_partial, prov_values="low_income_families",
                                           k_list=list(ksqexp=ksqexp, kexp=kexp),  l1=c(1,1.25,1.5,1.75,2,2.25,2.5),
                                           l2=c(0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25), b=c(0.5,1,1.5),
                                           meas_error=c(0.5,0.1,0.05,0.01,0))
  synthetic_modelselection_LI %>% 
    readr::write_csv(here::here("results", paste0("Synthetic_modelselection_LI_",i,".csv")))
  
  #' Step 3: Identify how well the true model performs and store results
  true_model <- synthetic_modelselection_LI %>% select(-model_no, -lppd_LOPO) %>%
    filter(k_name==params$k_name & l1==params$l1 & l2==params$l2 & b==params$b & meas_error==params$meas_error)
  min_lppd <- synthetic_modelselection_LI %>% filter(lppd_exact==min(lppd_exact)) %>% pull(lppd_exact)
  max_lppd <- synthetic_modelselection_LI %>% filter(lppd_exact==max(lppd_exact)) %>% pull(lppd_exact)
  
  SBC_results_list[[i]] <- tibble(i=i, true_model=true_model, min_lppd=min_lppd, max_lppd=max_lppd)
}
SBC_results <- bind_rows(SBC_results_list)
SBC_results %>% saveRDS(here::here("results", "SBC_results_LI.rds"))
```
