---
title: "Full Gaussian Process model for estimating vaccine coverage"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

# Set Up

Load functions and packages. Read the standardized cleaned data from the 'take 2' approach (aiming to collapse different dose estimates to 1+). The data set `vax_clean` contains province-level estimates. We aim to work with current age (effectively collapsing year_birth and age dimensions into one).

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(mnormt) # for multivariate normal
library(LaplacesDemon) # for logistic transform

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read cleaned data from 'take 2' approach
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
```

## A note on scales

Model fitting is done using Gaussian distributions, whose domain is the entire real line. However, the vaccine coverage data is restricted to $[0,1]$ (though in practice, they're not on the boundaries of this interval). Thus, we transform the vaccine coverage data using a logit function to map it from $(0,1)$ to $(-\infty, \infty)$, perform the model fitting on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.


# 1D case (current age)
The 1D model can be run for any single province, e.g. Alberta.
```{r}
vax_prov <- vax_clean %>% filter(pt=="AB") #%>% filter(n_doses!="2")
```

## Define covariance function / covariance matrix

Use the squared exponential covariance function, `k`, to begin with and assume lengthscale `l=2`. Our dataset has 40 inputs/ independent variables (current age), so our covariance matrix is a 40x40 matrix defining the 'relationships' between each pair of current ages.

```{r}
#| label: covariance

ksqexp <- function(r,l) {  # this is also defined in ksqexp.R
  exp(-r^2 / (2*l^2))
}

first_agecurrent <- 1  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model 
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  
npts <- length(xvals)
l <- 2
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>%  # evaluation of cov function pairwise for current ages x1_i and x2_i (i=1,..,n)
  pull(k) %>% matrix(nrow=npts)
# could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (current ages).

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals) + geom_point(alpha=0.3) + geom_line(alpha=0.2)
```

## Introduce observed data

For this mini dataset `vax_prov`, we have a dataset of n data points. First expose the model to ~10 data points (to see how well it can predict the remainder). We will use the logistic transform of y (`logit(yobs)`) which maps from \[0,1\] to \[-inf,inf\].

```{r}
#| label: observe_data

# observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30
xobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(age_current)   # can optionally add an error term here
yobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(value)
# xobs <- vax_prov %>% pull(age_current)   # can optionally add an error term here
# yobs <- vax_prov %>% pull(value)

tibble(xobs, yobs, logit(yobs))
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved current ages `xvals` and observed current ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

# calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed logit-transformed data. The posterior function can be used to provide estimates of our missing coverage data if inverse-transformed back to a \[0,1\] interval (second plot).

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals) + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red")

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```

We can add in the remaining true data (which the model has not seen) to evaluate its performance. The true coverage values are marked as red (seen) and green (not seen) points.

```{r}
#| label: compare_to_data
post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(name="current age x_i", breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(age_current)), y=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(value))), aes(x=x,y=y), col="green") + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")

#ggsave(paste0("plots/ABpost_l",l,"_ksqexp",".pdf"), height=4.51, width=7.29)
```
## Extensions
To improve this model, the next step would be to experiment with choice of covariance function `k` and lengthscale `l`, and to add in province as a second dimension. It is also worth considering how to treat dose data - currently the model struggles with data points for 2 dose. It may be worth fitting two fairly separate GPs (1+ and 2 dose), or using the 2 dose estimates as minimums for the prior in some way.


# 2D case (current age and province)
The 2D model adds province as a second dimension (only prior and observed data shown do far). Note: more information is needed to inform relationship between provinces - currently an arbitrary alphabetical relation is assumed.

## Covariance and prior
We use the squared exponential covariance function to begin with, with current age lengthscale, `l1`, and _relative_ lengthscale of current age to province, `l2`. Define the prior ~ Normal(0,cov) as before and draw a sample.
```{r}
#| label: prior_2D

# setup
# now both x and y are inputs, with output z
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced) 
yvals <- 1:length(unique(vax_clean$location))  # a vector of provinces -- currently arbitrary relation (alphabetical)
names(yvals) <- sort(unique(vax_clean$location))
# xvals <- vax_clean %>% mutate(age = pull_first(age)) %>% pull(age) %>% unique() %>% sort()  # a vector of ages
# yvals <- vax_clean %>% pull(year_report) %>% unique() %>% sort()  # a vector of year_report
npts <- length(xvals)*length(yvals)
l1 <- 2  # x lengthscale
l2 <- 1  # relative lengthscale of x values to y values
yvals <- yvals*l2  # scale y values to the x lengthscale

# covariance matrix in 2D
covmat <- expand_grid(x1=xvals, y1=yvals, x2=xvals, y2=yvals) %>% mutate(k=ksqexp(sqrt((x1-x2)^2 + (y1-y2)^2), l1)) %>%
  pull(k) %>% matrix(nrow=npts)

# define prior and take 50 draws
xygrid <- expand_grid(x=xvals, y=yvals)
prior2D <- tibble() 
for(i in 1:50){  # 50 realisations from a 2D gaussian process, with both x, y axes independent variables
  prior2D <- bind_rows(prior2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, prior2D=rmnorm(1, rep(0,npts), covmat + 1e-6*diag(npts))))
}

# plot prior sample
prior2D <- prior2D %>% mutate(y_label = names(yvals)[match(y_i, yvals)])
prior2D %>% ggplot(aes(x=x_i, y=prior2D, group=factor(draw))) + scale_x_continuous(breaks=xvals,name="current age x_i") + geom_line(alpha=0.3) + facet_wrap(~ y_label)
```
## Observed data
Expose the model to ~10 data points (to see how well it can predict the remainder). We will use the logistic transform of z (logit(zobs)) which maps from [0,1] to [-inf,inf].
```{r}
#| label: data_2D

# observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30, in all provinces where data exists
xobs <- vax_clean %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(age_current)   # can optionally add an error term here
yobs <- yvals[vax_clean %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(location)]  # pulling province values from yvals by name for now
zobs <- vax_clean %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(value)
# xobs <- vax_clean %>% pull(age_current)   # can optionally add an error term here
# yobs <- vax_clean %>% pull(location)
# zobs <- vax_clean %>% pull(value)

yobs <- yobs*l2  # scaling y data by relative lengthscale l2

tibble(xobs, yobs, zobs, logit(zobs))
```

