---
title: "Modelling similarly-sparse UK data"
author: "Irena Papst"
format: 
    html:
        embed-resources: true
---

```{r}
#| label: setup

library(dplyr)
library(ggplot2)

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates

attach_country <- function(df, country){
    if("pt" %in% names(df)) df <- df |> select(-pt)

    df |>
        mutate(country = !!country) |>
        relocate(country, .before = location)
}

# Canadian data
# - - -
# for Gini model
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
# for low income cleaned data for use with low-income and vaccine hesitancy province relation indicators
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT"))) # "low_income_families" has no data for NU, NT, YT

# England data
# - - -
vax_england <- readr::read_csv("data/ukhsa-chart-download-mmr1-regions.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (regions of England)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)

vax_gini <- bind_rows(vax_clean |> attach_country("Canada"), vax_england |> attach_country("England")) |> filter(n_doses == "1+")
vax_LI <- bind_rows(vax_cleanLI |> attach_country("Canada"), vax_england |> attach_country("England"))
```

To strengthen the application of the model to the sparse Canadian data, we want to apply it to a similarly sparse subsample of the very complete UK data.

# Experimental design

The main task here is to emulate the sparsity of the Canadian data to some part of the UK data. The closest comparison is likely to be for England, as it has a similar number of sub-regions as the number of PTs in Canada (exact numbers of PTs depends on the PT-similarity metric as we do not have these metrics for all PTs in our standardized data). The PT-similarity metrics we have for England are: Gini index, percentage of children not in low-income families. (There may be data on vaccine hesitancy, but we have not looked for it yet.)

We will want to run the same experiment (N reps) for each PT similarity metric:

1. Subsample the England data to mimic the CA data sparsity
    - Quantify the CA data as a vector of the number of data points available by PT
    - Sample this vector without replacement 8 times (to match the number of regions in England); call this the "sample size vector".
    - For each region of England, draw V vaccine coverage estimates, where V corresponds to an entry of the sample size vector.
2. Fit the GP model to the subsampled data (code from `gp-model.qmd`)
3. Calculate an accuracy based on left-out data (full set; code from `gp-model.qmd`)
    - Liza mentioned that this falls out of the model fitting process already? I can't remember the details right now...

Then we can summarize accuracy for each PT similarity metric.

# Data exploration and prep

```{r}
#| label: n_obs_gini

n_obs <- vax_gini |> group_by(country) |> select(location) |> distinct() |> count()
```

We start with an analysis of the Gini index model, which has data for `r n_obs |> filter(country == "Canada") |> pull(n)` Canadian PTs and `r n_obs |> filter(country == "England") |> pull(n)` regions of England.

The age ranges in Canadian data do not exactly conincide with those found in the England data:

```{r}
vax_gini |>
    group_by(country, location) |>
    summarize(age_min = min(age_current), age_max = max(age_current))
```

If we filter both down to the largest intersection in age for a fair comparison, and we are left with the following ranges by location: 

```{r}
#| label: filter_data
vax <- vax_gini |>
    group_by(country) |>
    mutate(age_min = min(age_current), age_max = max(age_current)) |>
    ungroup() |>
    mutate(age_min = max(age_min), age_max = min(age_max)) |>
    filter(between(age_current, age_min, age_max))
```

```{r}
vax |>
    group_by(country) |>
    group_split() |>
    purrr::map(\(df){
        df |>
            group_by(location) |>
            summarize(
                age_min = min(age_current), age_max = max(age_current)
            )
    })
```

After filtering, we now only have 6 PTs in the Canadian data, so when testing against the England data, we will only take 6 regions at a time to fit.

# Implementation

We start by getting the sample size of the (filtered) PT data:

```{r}
sample_size <- vax |>
    filter(country == "Canada") |>
    group_by(location) |>
    summarize(n = n())
```

Pull a subset of the England data according to the PT sample sizes:

```{r}
vax_england <- filter(vax, country == "England")

sample_by_location <- tibble(
    location = sample(
        unique(vax_england$location), 
        size = nrow(sample_size), 
        replace = FALSE
    ), 
    n = sample(sample_size$n)
)

vax_england_subset <- vax_england |>
  right_join( # filtering join to subsample regions
    sample_by_location,
    by = "location"
  )

vax_train <- vax_england_subset |>
  group_split(location) |>
  purrr::map(\(df) slice_sample(df, n = df$n[1]) |> select(-n)) |>
  purrr::list_rbind()

vax_test <- anti_join(vax_england_subset, vax_train)
```

Fit the model to the subsampled data:

```{r}
#| label: fit_england_gini

rerun <- FALSE
if(rerun){
    fit <- select_GP(
        vax_train, 
        prov_values="UK-Gini", 
        k_list=list(ksqexp=ksqexp, kexp=kexp),  
        l1=c(1,1.25,1.5,1.75,2,2.25,2.5), 
        l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), 
        b=c(0.5,1,1.5), 
        meas_error=c(0.5,0.1,0.05,0.01,0)
    )

    # temp save since the above takes a while to run
    if(!dir.exists(here::here("results", "tmp"))) dir.create(here::here("results", "tmp"))

    saveRDS(fit, here::here("results", "tmp", "fit_england_gini.rds"))
} else {
    fit <- readRDS(here::here("results", "tmp", "fit_england_gini.rds"))
}

# retreive optimal hyperparameters
fit_opt <- fit |> arrange(desc(lppd_exact)) |> slice(1)
```

To calculate an accuracy score based on the left-out (test) data, it seems reasonable to use the posterior predictive density (PPD) in some way. However, I don't think summing the (L)PPD values will result in an intuitively interpretable accuracy score, because it's not giving us values that are readily interpretable. Point densities are not probabilities. Of course, these values are useful _in comparisson_, _i.e._, if looking for a model that is more or less predictive of left-out data than another one, but it's not straight forward to interpret. 

How should we define the accuracy, then? I discussed this with David Champredon, and we came to the following idea:

1. For each test point, choose an acceptable level of error, _e.g._, within 5% of the true vaccine coverage (the "target interval").
2. Retrieve parameters for the fitted model's (normal) posterior predictive distribution at the test point (distribution mean and variance).
3. Calculate the density's probability mass captured within the target interval (difference between a normal cumulative distribution function evaluated at the edges of the target interval). This tells us how likely it is that the model would have predicted the true value given our chosen error tolerance.

For instance, say true vaccine coverage is 80% for kids aged 7 in London. We choose an error tolerance of 5%, because epidemiologically, these are "similar enough" values. Then our target interval is 77.5%-82.5%. Suppose 70% of the fitted model's probability mass is in that interval; then our model had a 70% chance of predicting the true value with an acceptable amount of error. The more error we tolerate, the higher the accuracy, but this way the accuracy is informed directly by the real-world application, and so it can be interpretted. 

The proceedure above could be repeated for each test point, and then we could compute summary statistics on it: an average accuracy score along with a range or confidence interval.

## Prototype

`compute_lppd()` computes required inputs for the above accuracy calculation (mean and variance of the posterior predictive density at each test point). Notes on the math can be found in [this issue](https://github.com/phac-modelling-hub/immunity-landscape/issues/4#issuecomment-3410698100).

```{r}
#| label: compute_accuracy

error_tolerance <- 0.02 # 2%
last_agecurrent <- 21
prov_values <- "UK-Gini"
k <- get(fit_opt$k_name)
l1 <- fit_opt$l1
l2 <- fit_opt$l2
b <- fit_opt$b
meas_error <- fit_opt$meas_error

# to use compute_lppd() equations, need to feed in vax_train _and_ vax_test, then only do the for loop stuff on the indices corresponding with vax_test (add those indices above!)
acc <- compute_accuracy(
    train = vax_train, 
    test = vax_test,
    error_tolerance = error_tolerance,
    last_agecurrent = last_agecurrent,
    prov_values = prov_values,
    k = k,
    l1 = l1,
    l2 = l2,
    b = b,
    meas_error = meas_error
)
```

```{r}
#| label: plot_accuracy

ggplot(data = tibble::tibble(acc = acc), aes(x = acc)) +
    geom_histogram() +
    labs(
        title = "Distribution of accuracy across test points",
         x = "Accuracy",
         y = "Count"
    )
```

The accuracy values look pretty good! Not all perfect but pretty good considering the error tolerance is only 2%. 

Another way to visualize would be to create a plot of posterior draws by age and region, include the train obs as black points, and then test obs coloured by accuracy (on a continuous scale between 0 and 1).[^1] This would help us see which points have been easily predicted and which have not. Though this is already, to some extent, eye-ballable when test points are plotted overtop of posterior draws because that already gives an impression of prediction mean/variance under the test point... so maybe not worth spending time on?

[^1]: I think this could be done relatively easily, but it would be good to repurpose existing code to plot posterior draws for this (perhaps from the `plot_posterior` chunk within `gp-model.qmd`)?

Note, the above is just for one test set, so we would want to repeat this process multiple times (with different random subsamples of the England data). That would give us multiple sets of accuracy scores, and we could run summary stats on this. We could then also run a sensitivity analysis of these scores to the error tolerance to show just how robust the framework is. 

The above is also only for the Gini model. We could repeat for the low-income model and then compare accuracy values between them to see if one PT similarity metric is potentially more predictive of vaccine coverage than the other.

This is getting to be a bit of a data heavy experiment... may want to consider moving to a `{targets}` pipeline.