---
title: "Full Gaussian Process model for estimating vaccine coverage"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

# Set Up

Load functions and packages. Read the standardized cleaned data from the 'take 2' preliminary modelling approach (aiming to collapse different dose estimates to 1+). The data set `vax_clean` contains province-level estimates. We aim to work with current age (effectively collapsing year_birth and age dimensions into one).

```{r}
#| label: initial_setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(stringi)
library(scales)
library(purrr)
library(tibble)
library(mnormt) # for multivariate normal
library(LaplacesDemon) # for logistic transform

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read cleaned data from 'take 2' approach
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
vax_cNICS <- readr::read_csv(here::here("data", "measles_CNICS.csv"), show_col_types = FALSE) %>%
  filter(!(pt %in% c("SK","YT","NB","CA"))) %>% mutate(age_current=current_year - year_birth)
```

## A note on scales

Model fitting is done using Gaussian distributions, whose domain is the entire real line. However, the vaccine coverage data is restricted to $[0,1]$ (though in practice, they're not on the boundaries of this interval). Thus, we transform the vaccine coverage data using a logit function to map it from $(0,1)$ to $(-\infty, \infty)$, perform the model fitting on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.

## A note on assumptions

A Gaussian Process model assumes relation of the independent variable(s), i.e. that ages which are 'related / close-by' (defined however we wish with lengthscale parameter l) will have similar vaccination coverage. Equally, we assume that 'related' provinces will also have similar coverage. Some tests of this assumption are the 2D model extensions section.

# 1D case (current age)

The 1D model can be run for any single province, e.g. Alberta.

```{r}
vax_prov <- vax_clean %>% filter(pt=="AB") #%>% filter(n_doses!="2")
```

## Define covariance function / covariance matrix

Use the squared exponential covariance function, `k`, to begin with and assume lengthscale `l=2`. Our dataset has 40 inputs/ independent variables (current age), so our covariance matrix is a 40x40 matrix defining the 'relationships' between each pair of current ages.

```{r}
#| label: covariance

# ksqexp <- function(r,l,b=1) {  # this is also defined in ksqexp.R
#   b*exp(-r^2 / (2*l^2))
# }

first_agecurrent <- 1  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model 
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  
npts <- length(xvals)
l <- 2
#' evaluate cov function pairwise for current ages x1_i and x2_i (i=1,..,n)
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>% 
  pull(k) %>% matrix(nrow=npts)  # could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). This prior describes the relationship between vaccination coverage and current age. It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (current ages). Vertical axis is vaccination coverage.

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") + geom_point(alpha=0.3) + geom_line(alpha=0.2)
```

## Introduce observed data

For this mini dataset `vax_prov`, we have a dataset of n data points. First expose the model to \~8 data points (to see how well it can predict the remainder). We will use the logistic transform of y (`logit(yobs)`) which maps from \[0,1\] to \[-inf,inf\].

```{r}
#| label: observe_data

#' observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30
xobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(age_current)   # can optionally add an error term here
yobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(value)
# xobs <- vax_prov %>% pull(age_current)   # can optionally add an error term here
# yobs <- vax_prov %>% pull(value)

tibble(xobs, yobs, logit(yobs))
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved current ages `xvals` and observed current ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

#' calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed logit-transformed data. The posterior function can be used to provide estimates of our missing coverage data if inverse-transformed back to a \[0,1\] interval (second plot).

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red")

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```

We can add in the remaining true data (which the model has not seen) to evaluate its performance. The true coverage values are marked as red (seen) and green (not seen) points.

```{r}
#| label: compare_to_data

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(name="current age x_i", breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(age_current)), y=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(value))), aes(x=x,y=y), col="green") + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")

#ggsave(paste0("plots/ABpost_l",l,"_ksqexp",".pdf"), height=4.51, width=7.29)
```

## Extensions

To improve this model, the next step would be to experiment with choice of covariance function `k` and lengthscale `l`, and to add in province as a second dimension. It is also worth considering how to treat dose data - currently the model struggles with data points for 2 dose. It may be worth fitting two fairly separate GPs (1+ and 2 dose), or using the 2 dose estimates as minimums for the prior in some way.

# 2D case (current age and province)

The 2D model adds province as a second dimension. In this model, x_i is current age, y_i is province, and vertical axis (z_i) is always estimated vaccine coverage. Note: more information is needed to inform relationship between provinces.

## Covariance and prior

We use the squared exponential covariance function to begin with, with current age lengthscale, `l1`, and *relative* lengthscale of current age to province, `l2`. If the squared exponential is too strong an assumption, we can also increase its variance (i.e. the model uncertainty) with parameter `b`. Relation of provinces currently uses GDP from StatCan (https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610040201) -- this should be updated when a better method is found. Define the prior \~ Normal(0,cov) as before and draw a sample.

Note that we currently assume a mean of 0. We could specify a non-zero mean here in the prior, or equivalently, we could centre the observed data to mean 0, run the Gaussian Process as normal and then rescale the data back at the end. We will do the latter as it simplifies calculation.

```{r}
#| label: prior_2D

#' setup
#' now both x and y are inputs, with output z
first_agecurrent <- 1  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model 
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced) 

province_relation <- readr::read_csv(here::here("data", "3610040201-noSymbol.csv"), show_col_types = F) %>%
  filter(location %in% unique(vax_clean$location))  # province GDP data from StatCan
yvals <- log(province_relation %>% pull(Dollars))  # a vector of provinces -- currently log(GDP)
names(yvals) <- province_relation %>% pull(location)

npts <- length(xvals)*length(yvals)
l1 <- 1.5  # x lengthscale
l2 <- 3  # relative lengthscale of x values to y values
yvals <- yvals*l2  # scale y values to the x lengthscale

#' covariance matrix in 2D
b <- 1 # optional function scale for ksqexp determining output variance
covmat <- expand_grid(x1=xvals, y1=yvals, x2=xvals, y2=yvals) %>% mutate(k=ksqexp(sqrt((x1-x2)^2 + (y1-y2)^2), l1, b=b)) %>%
  pull(k) %>% matrix(nrow=npts)

#' define prior and take 50 draws
xygrid <- expand_grid(x=xvals, y=yvals)
prior2D <- tibble() 
for(i in 1:50){  # 50 realisations from a 2D gaussian process, with both x, y as independent variables
  prior2D <- bind_rows(prior2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, prior2D=rmnorm(1, rep(0,npts), covmat + 1e-6*diag(npts))))
}

#' plot prior sample
prov_levels <- (province_relation %>% pull(location))  # useful shortcut for defining province factors
prior2D <- prior2D %>% mutate(y_label = names(yvals)[match(y_i, yvals)]) %>% 
  mutate(y_label = factor(y_label, levels=prov_levels))
prior2D %>% ggplot(aes(x=x_i, y=prior2D, group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") +
  geom_line(alpha=0.3) + facet_wrap(~ y_label)

# prior2D %>% ggplot(aes(x=x_i, y=invlogit(prior2D + mean(logit(zobs))), group=factor(draw))) +  # transformed centred plot for talks
#   scale_x_continuous(name="current age x_i", limits=c(1,32), breaks=xvals) + ylab("prior") + ylim(c(0,1)) +
#   geom_line(alpha=0.3) + facet_wrap(~ y_label) + theme(axis.text.x = element_text(size = 6))
```

## Observed data

Expose the model to \~50 data points (to see how well it can predict the remainder). Exclude 2-dose data. We will use the logistic transform of z (`logit(zobs)`) which maps from \[0,1\] to \[-inf,inf\]. Our GP uses a prior mean of 0 so we centre the data around 0 after transforming.

To expose the model to the full vaccine coverage dataset, uncomment the three commented lines below.

Extension: A possible model extension is to add 2-dose data as minimums at this point. We are currently considering how to do this...

```{r}
#| label: data_2D

#' observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30 in all provinces where data exists
xobs <- vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(age_current)   # can optionally add error term here
yobs <- yvals[vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(location)]  # Alberta==25.5 etc as in yvals (log(GDP))  **already scaled by l2
zobs <- vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(value)

#' alternatively uncomment to observe all available data (excluding 2-dose)
xobs <- vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(age_current)
yobs <- yvals[vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(location)]
zobs <- vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(value)

#' optional extension: add 2-dose data as minimums?
# vax_obs <- vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2")  # unclear how to incorporate as a minimum

#' apply logistic transform and then centre the data around mean 0
logit_zobs_centred <- logit(zobs) - mean(logit(zobs))

tibble(xobs, yobs, zobs, logit_zobs_centred)
```

## Posterior

As in the 1D case, the posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved independent variables `{xvals, yvals}` and observed independent variables `(xobs, yobs)`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`. In the 1D case, our unobserved was `xvals`, now it is the concatenated vector grid `{xvals, yvals}` of size 320x320 (32 ages in each of 10 provinces).

Plot a sample of posterior curves against the observed data (red points).

```{r}
#| label: posterior_2D

xygrid <- expand_grid(x=xvals, y=yvals)
xyobs <- tibble(x=xobs,y=yobs)

kuo <- generate_2Dksqexp_covmat(xygrid,xyobs,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between unobserved and observed
kou <- generate_2Dksqexp_covmat(xyobs,xygrid,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between observed and unobserved
koo <- generate_2Dksqexp_covmat(xyobs,xyobs,fn=ksqexp,l=l1,b=b) + 1e-6*diag(nrow(xyobs))  # protect against non-invertibleness
kuu <- generate_2Dksqexp_covmat(xygrid,xygrid,fn=ksqexp,l=l1,b=b)

#' calculate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit_zobs_centred)  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

```{r}
#| label: plot_posterior_2D

#' draw from the posterior distribution and record in dataframe post2D
post2D <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post2D <- bind_rows(post2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, post2D=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post2D <- post2D %>% mutate(post2D_constrained=invlogit(post2D + mean(logit(zobs)))) %>%  # add column to return mean and transform back to [0,1]
  mutate(y_label = names(yvals)[match(y_i, yvals)]) %>%  # add column to return province names
  mutate(y_label = factor(y_label, levels=(province_relation %>% pull(location)))) %>%  # order provinces
  select(draw, x_i, y_i, y_label, post2D, post2D_constrained)  # reorder columns

#' plot posterior sample 
yobs_factors <- factor(names(yobs), levels(post2D$y_label))  # convert yobs to factors for use in facet_wrap
post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=logit_zobs_centred, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") +
  facet_wrap(~ y_label)

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)
```

As before, we can compare performance by adding back in data points that the model has not seen in green (including for example 2-dose data):

```{r}
#| label: compare2D

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) + 
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) + 
  geom_point(data=tibble(x=(vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(age_current)),  # adding in full data from vax_clean
                         y=(vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(value)), 
                         y_label=factor((vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(location)), levels=prov_levels)),
                         aes(x=x,y=y), col="green") + 
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)

#ggsave(paste0("plots/post2D_l1",l1,"_l2",l2,"_ksqexp","_alldata",".pdf"), height=4.51, width=7.29)
```

Optional: We can also drop any posterior draws that do not satisfy the 2-dose data points as a minimum. Due to the underlying mathematics, this is (*we think*) mathematically equivalent to constraining the prior and proportionally reweighting the remaining valid draws. This turns out to be too strong a constraint so the code snippet is skipped.

```{r}
#| label: set_minimums2D
#| eval: false

# extract 2-dose data from vax_clean, matching column names with post2D
data_2dose <- tibble(x_i=(vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(age_current)),  # adding in full data from vax_clean
                     y_label=factor((vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(location)), levels=prov_levels),
                     value_2dose=(vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(value)))
# filter out posterior draws that do not meet the criterion, using all()
post2D_new <- post2D %>% ungroup() %>% left_join(data_2dose, by=c("x_i","y_label")) %>%
  mutate(value_2dose=ifelse(is.na(value_2dose), 0, value_2dose)) %>%  ##UP TO HERE
  mutate(row_valid=(post2D_constrained >= value_2dose)) %>%
  group_by(draw) %>% filter(all(row_valid))
```

# 2D case model extensions

## Model functions

We now wish to improve our basic model by adding information to the prior and experimenting with choice of covariance function and related parameters. First, define helper functions to enable easy running of the 2D model (saved in "R/"):

-   `compute_prior2D()`
-   `plot_prior2D()`
-   `compute_posterior2D()`
-   `plot_posterior2D()`.

The full 2D Gaussian Process model can then be run with `run_GP(vax_dataset)`. All functions work with multiple province relation datasets.

```{r}
#| label: auto2D

run_GP(vax_clean)
```

From a quick by-eye model selection with GDP province relation, current best GP model parameter values are approximately `l1=1.5, l2=3, b=0.75` for `k=ksqexp`. Plausible range (by-eye) is `1 <= l1 < 2`, `l1 < l2 < 3*l1`, and `0.5 <= b <= 1.5` (GDP, provincial data, no measurement error). This hyperparameter space can be searched with`select_GP()` below.

```{r}
#| label: fit_byeye

# for (l1 in seq(1,3,1)) {  # l1>=2 gives meaningless outputs; l=1.5 (or 1) seems reasonable.
#   for (l2 in seq(1,5,0.5)) {  # l2<=2.5 is too strong a relation; l>=3 seems reasonable for log(GBP) data.
#     for (b in seq(0.5,1,0.25)) {  # b>1 gives too much variance; b=0.75 or 1 seems reasonable at present.
#       run_GP(vax_clean, l1=l1, l2=l2, b=b)
#     }
#   }
# }

run_GP(vax_clean,l1=1.5,l2=3,b=0.75)
```

## Running with cNICS data

Note that the provincial estimates collected in `vax_clean` are from multiple sources with varying data collection methodology. As an alternative, the national cNICS study, although perhaps less trustworthy, does have consistent methodology across provinces. We make use of Irena's cleaned, standardised cNICS data and run_GP(). The final plot compares model fit against the full (1+ and 2 dose) provincial dataset `vax_clean`.

Note: Nunavut youngest estimate is described as "treat with caution" in the cNICS online dataset.

Plausible range (by-eye) is `1 <= l1 <= 4`, `l1 < l2 <= 50*l1`, and `0.5 <= b <= 1.5` (GDP, cNICS data, no measurement error). This hyperparameter space can be searched with`select_GP()` below.

```{r}
#| label: run_cNICS

run_GP(vax_cNICS, l1=3.5, l2=4, b=1)
```

## Measurement error

Our GP model currently treats the observed data as exact estimates - posterior draws are forced to cross through all observed data points. We hence add measurement error. For this, the posterior mean and covariance matrix are modified when a parameter `meas_error` is specified in `run_GP()`:

```{r}
#| label: meas_error

run_GP(vax_clean,l1=1.5,l2=3,b=0.75)  # exact
run_GP(vax_clean,l1=1.5,l2=3,b=0.75,meas_error=0.01,ndrws=50)  # with measurement error
run_GP(vax_clean,l1=1.5,l2=3,b=0.75,meas_error=0.1,ndrws=50)  # with measurement error
run_GP(vax_clean,l1=1.25,l2=1,b=1.5,meas_error=0.01,ndrws=50)  # with measurement error

```

See the [GitHub Issue #3](https://github.com/phac-modelling-hub/immunity-landscape/issues/3) for a discussion of methodology.

## Model selection and validation

We aim to find the optimal choice of model (hyperparameter values `k`, `l1`, `l2`, `b`), and validate with take-one-out cross-validation. First assume no measurement error. An example algorithm is as follows:

-   For a given covariance function k and lengthscale parameters l1, l2, b:

    1.  Pull out one data point.

    2.  Run GP model and compute e.g. sum squared error between posterior mean and the missing data point.

    3.  Repeat for every data point (e.g. 137 model runs for the 1+ provincial data). Sum errors.

-   Select the optimal k/ l1/ l2/ b combination.

One option is to work with the posterior mean at each missing age as described. A better option is to work with a richer goodness-of-fit metric such as the (log) *posterior predictive density* at each missing age instead. This makes use of the full richness of our Gaussian Process model. We hence choose the latter metric.

`estimate_lppd()` estimates the posterior predictive density from a sample, in a similar way to the algorithm described above. `compute_lppd()` calculates the full log posterior predictive density analytically. See the [GitHub Issue #4](https://github.com/phac-modelling-hub/immunity-landscape/issues/4) for a discussion of methodology. `select_GP()` then does a search of hyperparameter space and returns the lppd for each hyperparameter set. The 'optimal' model (with max lppd) is also printed to the console.

```{r}
#| label: model_selection_noerror

#' define province relation (prov_values)
prov_values <- extract_province_relation("GDP", vax_dataset=vax_clean)

#' examples of estimating/ calculating lppd
#example1 <- estimate_lppd(vax_clean, prov_values=prov_values, k=ksqexp, l1=1.5, l2=3, b=1)  # run time: ~15 seconds
#example2 <- compute_lppd(vax_clean, prov_values=prov_values, k=ksqexp, l1=1.5, l2=3, b=1)  # run time: <1 second

#' example model selection
example <- select_GP(vax_clean, prov_values=prov_values, k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1))  # prov_values defined above

#' exhaustive search of reasonable hyperparameter space (provinical dataset)
GPcombinations_GDP <- select_GP(vax_clean, prov_values="GDP", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1.0,1.5,2,2.5,3.0,3.5,4.0), b=c(0.5,1,1.5))
GPcombinations_GDP %>% readr::write_csv(here::here("results", "GPcombinations_GDP.csv"))

#' exhaustive search of reasonable hyperparameter space (cNICS dataset)
GPcombinations_GDP_cNICS <- select_GP(vax_cNICS, prov_values="GDP", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.5,2,2.5,3,3.25,3.5,4), l2=c(1.5,2,3,3.5,4,4.5,5,6,10,50,100), b=c(0.5,1,1.5))
GPcombinations_GDP_cNICS %>% readr::write_csv(here::here("results", "GPcombinations_GDP_cNICS.csv"))
```

This 'best' hyperparameters result in spiky and overfitted estimates. Instead re-run model selection and validation with measurement error, to relax the model.

```{r}
#| label: model_selection_witherror

example3 <- compute_lppd(vax_clean, prov_values="GDP", k=ksqexp, l1=1.5, l2=3, b=1, meas_error=0.1)
example4 <- select_GP(vax_clean, prov_values="GDP", k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1), meas_error=0.1)

#' exhaustive search of reasonable hyperparameter space (provinical dataset) with error
GPcombinations_GDP2 <- select_GP(vax_clean, prov_values="GDP", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1,1.5,2,2.5,3.0,3.5,4.0), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_GDP2 %>% readr::write_csv(here::here("results", "GPcombinations_GDP_witherror.csv"))

#' exhaustive search of reasonable hyperparameter space (cNICS dataset) with error
GPcombinations_GDP_cNICS2 <- select_GP(vax_cNICS, prov_values="GDP", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.5,2,2.5,3,3.25,3.5,4), l2=c(1.5,2,3,3.5,4,4.5,5,6,10,50,100), b=c(0.5,1,1.5), meas_error=c(0.1,0.05,0.01))
GPcombinations_GDP_cNICS2 %>% readr::write_csv(here::here("results", "GPcombinations_GDP_cNICS_witherror.csv"))

run_GP(vax_clean,l1=1.25,l2=1,b=1,meas_error=0.01,ndrws=50)
run_GP(vax_clean,l1=1,l2=4,b=1,meas_error=0.01,ndrws=50)
```

We can also explore if there are local maximums in the model selection, rather than only considering the model with largest lppd.

```{r}
#| label: model_selection_heatmaps

my_tibble_GDP2 <- read_csv(here::here("results", "GPcombinations_GDP_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_LI2 <- read_csv(here::here("results", "GPcombinations_lowincome_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)
my_tibble_VH2 <- read_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv")) %>% filter(k_name == "ksqexp" & b == 1)

#' plot l1 against l2, colouring by lppd, for several meas_error values
my_tibble_GDP2 %>%
  mutate(lppd_bin = cut(lppd_exact, breaks = seq(-500, 100, length.out = 7))) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_bin)) +
  geom_tile() +
  scale_fill_brewer(name = "lppd > -500") +
  facet_wrap(~ meas_error) +
  labs(title = "lppd heatmaps for GDP model, varying measurement error",
       x = "l1", y = "l2", fill = "lppd > -500")

my_tibble_LI2 %>%
  mutate(lppd_bin = cut(lppd_exact, breaks = seq(-500, 100, length.out = 7))) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_bin)) +
  geom_tile() +
  scale_fill_brewer(name = "lppd > -500") +
  facet_wrap(~ meas_error) +
  labs(title = "lppd heatmaps for Low-Income model, varying measurement error",
       x = "l1", y = "l2", fill = "lppd > -500")

my_tibble_VH2 %>% 
  mutate(lppd_bin = cut(lppd_exact, breaks = seq(-500, 100, length.out = 7))) %>%
  ggplot(aes(x = l1, y = l2, fill = lppd_bin)) +
  geom_tile() +
  scale_fill_brewer(name = "lppd > -500") +
  facet_wrap(~ meas_error) +
  labs(title = "lppd heatmaps for Vaccine Hesitancy model, varying measurement error",
       x = "l1", y = "l2", fill = "lppd > -500")
```



## Relation of provinces

Relation of provinces is encoded in `prov_values` and can be extracted from the function `extract_province_relation()`. This function has a number of socioecomonic datasets preloaded in a standardised format for use in the GP model. Examples include "GDP" and "low_income_families". \*Note that for "low_income_families" there is currently no data for NU, NT, and YT.

```{r}
#| label: province_relation_examples

prov_values <- extract_province_relation("GDP", vax_dataset=vax_clean)
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT")))  # "low_income_families" has no data for NU, NT, YT
prov_values <- extract_province_relation("low_income_families", vax_dataset=vax_cleanLI)

example <- select_GP(vax_cleanLI, prov_values=prov_values, k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1))
example2 <- select_GP(vax_cleanLI, prov_values="low_income_families", k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1))

#' exhaustive search of reasonable hyperparameter space (provincial dataset, children in low-income families province relation metric)
# GPcombinations_lowincome<- select_GP(vax_cleanLI, prov_values="low_income_families", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1,1.5,2,2.5,3,3.5,4), b=c(0.5,1,1.5))
# GPcombinations_lowincome %>% readr::write_csv(here::here("results", "GPcombinations_lowincome.csv"))

#' with error
GPcombinations_LI2 <- select_GP(vax_cleanLI, prov_values="low_income_families", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1,1.5,2,2.5,3,3.5,4), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_LI2 %>% readr::write_csv(here::here("results", "GPcombinations_lowincome_witherror.csv"))

#' plots of final model (using best parameter values)
# run_GP(vax_clean, prov_values="GDP", l1=1.25, l2=1.5, b=0.5, meas_error=0.01)
# ggsave(paste0("plots/bestGPmodel_GDP",".pdf"), height=4.51, width=7.29)
# run_GP(vax_cNICS, prov_values="GDP", l1=2.5, l2=1.5, b=1.5, meas_error=0.01)
# ggsave(paste0("plots/bestGPmodel_GDP_cNICS",".pdf"), height=4.51, width=7.29)
# run_GP(vax_cleanLI, prov_values="low_income_families", l1=1.5, l2=1, b=0.5, meas_error=0.01)
# run_GP(vax_cleanLI, prov_values="low_income_families", l1=1.75, l2=2, b=0.5, meas_error=0.01)
# ggsave(paste0("plots/bestGPmodel_lowincome",".pdf"), height=4.51, width=7.29)
```

A new dataset on vaccine hesitancy requires some manipulation to use (condensing coverage data for Atlantic provinces and Northern provinces):

```{r}
#| label: province_relation_hesitancy

#' combine smaller provinces into regions
vax_regional <- vax_clean %>% filter(n_doses!="2") %>%
  mutate(location = if_else(location %in% c("Newfoundland and Labrador", "Prince Edward Island", "Nova Scotia", "New Brunswick"), "Atlantic region", location)) %>%
  mutate(location = if_else(location %in% c("Yukon", "Northwest Territories", "Nunavut"), "Northern region", location)) %>%
  mutate(pt = if_else(location %in% c("Atlantic region"), "AR", pt)) %>%
  mutate(pt = if_else(location %in% c("Northern region"), "NR", pt)) %>%
  group_by(across(-value)) %>% summarise(value = mean(value), .groups = "drop")

#' run with vax_regional
run_GP(vax_regional, prov_values="vaccine_hesitancy", l1=2, l2=1, b=1)  # l1>l2 looks ok here, but l1>4*l2 is spiky. And true l2 (l1*l2) meaningful when >1?

#' exhaustive search of reasonable hyperparameter space (provincial dataset, vaccine hesitancy province relation metric)
# GPcombinations_VH <- select_GP(vax_regional, prov_values="vaccine_hesitancy", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1,1.5,2,2.5,3,3.5,4), b=c(0.5,1,1.5))
# GPcombinations_VH %>% readr::write_csv(here::here("results", "GPcombinations_vaccinehesitancy.csv"))

#' with error
GPcombinations_VH2 <- select_GP(vax_regional, prov_values="vaccine_hesitancy", k_list=list(ksqexp=ksqexp, kexp=kexp), l1=c(1,1.25,1.5,1.75,2,2.25,2.5), l2=c(0.5,1,1.5,2,2.5,3,3.5,4), b=c(0.5,1,1.5), meas_error=c(0.5,0.1,0.05,0.01,0))
GPcombinations_VH2 %>% readr::write_csv(here::here("results", "GPcombinations_vaccinehesitancy_witherror.csv"))
```

## Assumptions tests

A Gaussian Process model assumes relation of the independent variable(s), i.e. that 'related' ages (defined however we wish with lengthscale parameter l1) will have similar vaccination coverage and that 'related' provinces will also have similar coverage.

To visualise relation of provinces, we can plot `province_relation_data` on an axis. One could compare to for example the mean cNICS coverage estimate for each province. Inspecting `provinces_test` (or looking at the example plots), we see that these are terrible -- suggesting either that all of our province relation data are poor OR that the cNICS data is poor? We also consider aNICS data but unfortunately this is no better. A better dataset is needed for testing the assumption of relation of provinces.

```{r}
#| label: assumption_test_provinces

#' load province relation values as tibbles
prov_valuesGDP <- enframe(extract_province_relation("GDP", vax_dataset=vax_clean), name="location", value="prov_valuesGDP")  # GDP
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT")))
prov_valuesLI <- enframe(extract_province_relation("low_income_families", vax_dataset=vax_cleanLI), name="location", value="prov_valuesLI")  # low_income
prov_valuesVH <- enframe(extract_province_relation("vaccine_hesitancy", vax_dataset=vax_regional), name="location", value="prov_valuesVH")  # vaccine_hesitancy (vax_regional defined above)
ARval <- prov_valuesVH %>% filter(location=="Atlantic region") %>% pull(prov_valuesVH)
NRval <- prov_valuesVH %>% filter(location=="Northern region") %>% pull(prov_valuesVH)
prov_valuesVH <- prov_valuesVH %>% rbind(c("Newfoundland and Labrador", ARval)) %>% rbind(c("Prince Edward Island", ARval)) %>%  # this is very crude code!
  rbind(c("Nova Scotia", ARval)) %>% rbind(c("Nunavut", NRval)) %>% rbind(c("Northwest Territories", NRval)) %>% mutate(prov_valuesVH=as.numeric(prov_valuesVH))

#' load cNICS as tibbles
cNICS_2021 <- vax_cNICS %>% filter(year_report==2021) %>% select(location, "cNICS_2021"=value)
cNICS_2017 <- vax_cNICS %>% filter(year_report==2017) %>% select(location, "cNICS_2017"=value)
cNICS_2013a <- vax_cNICS %>% filter(year_report==2013 & age==2) %>% select(location, "cNICS_2013a"=value)
cNICS_2013b <- vax_cNICS %>% filter(year_report==2013 & age==7) %>% select(location, "cNICS_2013b"=value)
cNICS_2013c <- vax_cNICS %>% filter(year_report==2013 & age==17) %>% select(location, "cNICS_2013c"=value)

#' combine into one
provinces_test <- prov_valuesGDP %>% left_join(prov_valuesLI, by="location") %>% left_join(prov_valuesVH, by="location") %>%
  left_join(cNICS_2021, by="location") %>% left_join(cNICS_2017, by="location") %>% left_join(cNICS_2013a, by="location") %>% 
  left_join(cNICS_2013b, by="location") %>% left_join(cNICS_2013c, by="location") %>% 
  mutate(cNICS_mean=(cNICS_2021 + cNICS_2017 + cNICS_2013a + cNICS_2013b + cNICS_2013c)/5)

#' plot
library(ggrepel)
ggplot(provinces_test, aes(x = prov_valuesLI, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = prov_valuesVH, y = cNICS_2021)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = prov_valuesVH, y = cNICS_mean)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)

#' instead consider aNICS data over cNICS
vax_aNICS <- vax_clean %>% filter(age_current == max(age_current)) %>% select(location, "aNICS"=value) # simple hack to extract an adult estimate
provinces_test <- provinces_test %>% left_join(vax_aNICS, by="location")
ggplot(provinces_test, aes(x = prov_valuesLI, y = aNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggsave(paste0("plots/bestprovrelation",".pdf"), height=4.51, width=7.29)  # example code to save best plot

#' instead consider provincial data mean over cNICS
vax_means_withaNICS <- vax_clean %>% filter(n_doses %in% c("1", "1+")) %>% group_by(location) %>% summarise(average = mean(value)) %>% 
  select(location, "vax_means_withaNICS"=average)
vax_means_noaNICS <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2-noadults.csv"), show_col_types = FALSE) %>% 
  filter(n_doses %in% c("1", "1+")) %>% group_by(location) %>% summarise(average = mean(value)) %>% select(location, "vax_means_noaNICS"=average)
provinces_test <- provinces_test %>% left_join(vax_means_withaNICS, by="location") %>% left_join(vax_means_noaNICS, by="location")
ggplot(provinces_test, aes(x = prov_valuesLI, y = vax_means_withaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = prov_valuesLI, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = prov_valuesGDP, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = prov_valuesVH, y = vax_means_noaNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)

#' for fun, plot provincial data against cNICS and aNICS
ggplot(provinces_test, aes(x = vax_means_noaNICS, y = cNICS_mean)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
ggplot(provinces_test, aes(x = vax_means_noaNICS, y = aNICS)) + geom_point(color = "blue", size = 3) + geom_text(aes(label = location), size = 4)
```

To test relation of ages, we can compute pairwise differences in absolute coverage for all data points and observe if pairs of adjacent ages have a lower difference on average:

```{r}
#| label: assumption_test_ages

#' 1+ dose provincial data, no aNICS
# read in provincial data with aNICS removed
vax_dataset1 <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2-noadults.csv"), show_col_types = FALSE) %>%
  filter(n_doses %in% c("1", "1+"))

#' define a function to compute all pairwise differences
compute_pairwise_diffs <- function(df) {
  df <- df %>% arrange(age_current)
  
  expand_grid(age1 = df$age_current, age2 = df$age_current) %>% filter(age1 < age2) %>%
    left_join(df %>% select(age1 = age_current, cov1 = value), by = "age1") %>%
    left_join(df %>% select(age2 = age_current, cov2 = value), by = "age2") %>%
    mutate(diff = abs(cov1 - cov2), adjacent = (age2 - age1 == 1))
}

#' compute pairwise for e.g. Alberta
alberta_pairs <- vax_dataset1 %>% filter(location=="Alberta") %>% compute_pairwise_diffs()

#' plot histogram for e.g. Alberta
ggplot(alberta_pairs, aes(x = diff, fill = adjacent)) + geom_histogram() +
  labs(title = "Pairwise Coverage Differences — Alberta (1+ dose)",
    x = "Absolute Difference in Coverage (%)",
    y = "Count",
    fill = "Adjacent ages")

#' compute pairwise and plot histogram for remaining provinces
province_pairs <- vax_dataset1 %>% group_by(location) %>% group_modify(~ compute_pairwise_diffs(.x))
ggplot(province_pairs, aes(x = diff, fill = adjacent)) + geom_histogram() + 
  facet_wrap(~ location) + labs(title = "Pairwise Coverage Differences — By Province (1+ dose)",
                                x = "Absolute Difference in Coverage (%)", y = "Count", fill = "Adjacent ages")
ggsave(paste0("plots/testing_age_assumption1",".pdf"), height=4.51, width=7.29)  # example code to save plot

#' 2 dose provincial data
#' repeat steps as above
vax_dataset2 <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2-noadults.csv"), show_col_types = FALSE) %>%
  filter(n_doses %in% c("2", "2+"))
province_pairs2 <- vax_dataset2 %>% group_by(location) %>% group_modify(~ compute_pairwise_diffs(.x))
ggplot(province_pairs2, aes(x = diff, fill = adjacent)) + geom_histogram() + 
  facet_wrap(~ location) + labs(title = "Pairwise Coverage Differences — By Province (2 dose)",
                                x = "Absolute Difference in Coverage (%)", y = "Count", fill = "Adjacent ages")
ggsave(paste0("plots/testing_age_assumption2",".pdf"), height=4.51, width=7.29)  # example code to save plot

#' Aggregated provincial data
vax_dataset3 <- vax_dataset1 %>% mutate(id = row_number()) %>% select(-location)  # remove location label
country_pairs <- expand_grid(id1 = vax_dataset3$id, id2 = vax_dataset3$id) %>%
  filter(id1 < id2) %>%
  left_join(vax_dataset3 %>% select(id1 = id, age1 = age_current, cov1 = value), by = "id1") %>%
  left_join(vax_dataset3 %>% select(id2 = id, age2 = age_current, cov2 = value), by = "id2") %>%
  mutate(diff = abs(cov1 - cov2), adjacent = (abs(age2 - age1) == 1))
ggplot(country_pairs, aes(x = diff, fill = adjacent)) + geom_histogram() + 
  labs(title = "Pairwise Coverage Differences — All provincial data (1+ dose)")

vax_dataset4 <- vax_dataset2 %>% mutate(id = row_number()) %>% select(-location)
country_pairs2 <- expand_grid(id1 = vax_dataset4$id, id2 = vax_dataset4$id) %>%
  filter(id1 < id2) %>%
  left_join(vax_dataset4 %>% select(id1 = id, age1 = age_current, cov1 = value), by = "id1") %>%
  left_join(vax_dataset4 %>% select(id2 = id, age2 = age_current, cov2 = value), by = "id2") %>%
  mutate(diff = abs(cov1 - cov2), adjacent = (abs(age2 - age1) == 1))
ggplot(country_pairs2, aes(x = diff, fill = adjacent)) + geom_histogram() + 
  labs(title = "Pairwise Coverage Differences — All provincial data (2 dose)")
```

We see a good relationship between coverage at adjacent ages, both in the 1+ and 2 dose data.
