---
title: "Full Gaussian Process model for estimating vaccine coverage"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

# Set Up

Load functions and packages. Read the standardized cleaned data from the 'take 2' preliminary modelling approach (aiming to collapse different dose estimates to 1+). The data set `vax_clean` contains province-level estimates. We aim to work with current age (effectively collapsing year_birth and age dimensions into one).

```{r}
#| label: initial_setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(purrr)
library(mnormt) # for multivariate normal
library(LaplacesDemon) # for logistic transform

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read cleaned data from 'take 2' approach
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
```

## A note on scales

Model fitting is done using Gaussian distributions, whose domain is the entire real line. However, the vaccine coverage data is restricted to $[0,1]$ (though in practice, they're not on the boundaries of this interval). Thus, we transform the vaccine coverage data using a logit function to map it from $(0,1)$ to $(-\infty, \infty)$, perform the model fitting on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.

## A note on assumptions

A Gaussian Process model assumes relation of the independent variable(s), i.e. that ages which are 'close-by' (defined however we wish with parameter l) will have similar vaccination coverage. We can test if this is a plausible assumption for our dataset. ...

# 1D case (current age)

The 1D model can be run for any single province, e.g. Alberta.

```{r}
vax_prov <- vax_clean %>% filter(pt=="AB") #%>% filter(n_doses!="2")
```

## Define covariance function / covariance matrix

Use the squared exponential covariance function, `k`, to begin with and assume lengthscale `l=2`. Our dataset has 40 inputs/ independent variables (current age), so our covariance matrix is a 40x40 matrix defining the 'relationships' between each pair of current ages.

```{r}
#| label: covariance

# ksqexp <- function(r,l,b=1) {  # this is also defined in ksqexp.R
#   b*exp(-r^2 / (2*l^2))
# }

first_agecurrent <- 1  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model 
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced)  
npts <- length(xvals)
l <- 2
#' evaluate cov function pairwise for current ages x1_i and x2_i (i=1,..,n)
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>% 
  pull(k) %>% matrix(nrow=npts)  # could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). This prior describes the relationship between vaccination coverage and current age. It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (current ages). Vertical axis is vaccination coverage.

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") + geom_point(alpha=0.3) + geom_line(alpha=0.2)
```

## Introduce observed data

For this mini dataset `vax_prov`, we have a dataset of n data points. First expose the model to \~8 data points (to see how well it can predict the remainder). We will use the logistic transform of y (`logit(yobs)`) which maps from \[0,1\] to \[-inf,inf\].

```{r}
#| label: observe_data

#' observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30
xobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(age_current)   # can optionally add an error term here
yobs <- vax_prov %>% filter(age_current %in% c(4,9,13,19,20,21,25,30)) %>% pull(value)
# xobs <- vax_prov %>% pull(age_current)   # can optionally add an error term here
# yobs <- vax_prov %>% pull(value)

tibble(xobs, yobs, logit(yobs))
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved current ages `xvals` and observed current ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

#' calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed logit-transformed data. The posterior function can be used to provide estimates of our missing coverage data if inverse-transformed back to a \[0,1\] interval (second plot).

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red")

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals,name="current age x_i") + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```

We can add in the remaining true data (which the model has not seen) to evaluate its performance. The true coverage values are marked as red (seen) and green (not seen) points.

```{r}
#| label: compare_to_data

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(name="current age x_i", breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(age_current)), y=(vax_prov %>% filter(age_current <= last_agecurrent) %>% pull(value))), aes(x=x,y=y), col="green") + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")

#ggsave(paste0("plots/ABpost_l",l,"_ksqexp",".pdf"), height=4.51, width=7.29)
```

## Extensions

To improve this model, the next step would be to experiment with choice of covariance function `k` and lengthscale `l`, and to add in province as a second dimension. It is also worth considering how to treat dose data - currently the model struggles with data points for 2 dose. It may be worth fitting two fairly separate GPs (1+ and 2 dose), or using the 2 dose estimates as minimums for the prior in some way.

# 2D case (current age and province)

The 2D model adds province as a second dimension. In this model, x_i is current age, y_i is province, and vertical axis (z_i) is always estimated vaccine coverage. Note: more information is needed to inform relationship between provinces.

## Covariance and prior

We use the squared exponential covariance function to begin with, with current age lengthscale, `l1`, and *relative* lengthscale of current age to province, `l2`. If the squared exponential is too strong an assumption, we can also increase its variance (i.e. the model uncertainty) with parameter `b`. Relation of provinces currently uses GDP from StatCan (https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610040201) -- this should be updated when a better method is found. Define the prior \~ Normal(0,cov) as before and draw a sample.

Note that we currently assume a mean of 0. We could specify a non-zero mean here in the prior, or equivalently, we could centre the observed data to mean 0, run the Gaussian Process as normal and then rescale the data back at the end. We will do the latter as it simplifies calculation.

```{r}
#| label: prior_2D

#' setup
#' now both x and y are inputs, with output z
first_agecurrent <- 1  # first age for the GP model
last_agecurrent <- 32  #age_pre1970-1  # last age for the GP model 
xvals <- first_agecurrent:last_agecurrent  # a vector of current ages (these x_i's are equally spaced) 

province_relation <- readr::read_csv(here::here("data", "3610040201-noSymbol.csv"), show_col_types = F) %>%
  filter(location %in% unique(vax_clean$location))  # province GDP data from StatCan
yvals <- log(province_relation %>% pull(Dollars))  # a vector of provinces -- currently log(GDP)
names(yvals) <- province_relation %>% pull(location)

npts <- length(xvals)*length(yvals)
l1 <- 1.5  # x lengthscale
l2 <- 3  # relative lengthscale of x values to y values
yvals <- yvals*l2  # scale y values to the x lengthscale

#' covariance matrix in 2D
b <- 1 # optional function scale for ksqexp determining output variance
covmat <- expand_grid(x1=xvals, y1=yvals, x2=xvals, y2=yvals) %>% mutate(k=ksqexp(sqrt((x1-x2)^2 + (y1-y2)^2), l1, b=b)) %>%
  pull(k) %>% matrix(nrow=npts)

#' define prior and take 50 draws
xygrid <- expand_grid(x=xvals, y=yvals)
prior2D <- tibble() 
for(i in 1:50){  # 50 realisations from a 2D gaussian process, with both x, y as independent variables
  prior2D <- bind_rows(prior2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, prior2D=rmnorm(1, rep(0,npts), covmat + 1e-6*diag(npts))))
}

#' plot prior sample
prov_levels <- (province_relation %>% pull(location))  # useful shortcut for defining province factors
prior2D <- prior2D %>% mutate(y_label = names(yvals)[match(y_i, yvals)]) %>% 
  mutate(y_label = factor(y_label, levels=prov_levels))
prior2D %>% ggplot(aes(x=x_i, y=prior2D, group=factor(draw))) + scale_x_continuous(breaks=xvals, name="current age x_i") +
  geom_line(alpha=0.3) + facet_wrap(~ y_label)

# prior2D %>% ggplot(aes(x=x_i, y=invlogit(prior2D + mean(logit(zobs))), group=factor(draw))) +  # transformed centred plot for talks
#   scale_x_continuous(name="current age x_i", limits=c(1,32), breaks=xvals) + ylab("prior") + ylim(c(0,1)) +
#   geom_line(alpha=0.3) + facet_wrap(~ y_label) + theme(axis.text.x = element_text(size = 6))
```

## Observed data

Expose the model to \~50 data points (to see how well it can predict the remainder). Exclude 2-dose data. We will use the logistic transform of z (`logit(zobs)`) which maps from \[0,1\] to \[-inf,inf\]. Our GP uses a prior mean of 0 so we centre the data around 0 after transforming.

To expose the model to the full vaccine coverage dataset, uncomment the three commented lines below.

Extension: A possible model extension is to add 2-dose data as minimums at this point. We are currently considering how to do this...

```{r}
#| label: data_2D

#' observe coverage data for age 4, 9, 13, 19, 20, 21, 25, 30 in all provinces where data exists
xobs <- vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(age_current)   # can optionally add error term here
yobs <- yvals[vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(location)]  # Alberta==25.5 etc as in yvals (log(GDP))  **already scaled by l2
zobs <- vax_clean %>% filter((age_current %in% c(4,9,13,19,20,21,25,30)) & n_doses!="2") %>% pull(value)

#' alternatively uncomment to observe all available data (excluding 2-dose)
xobs <- vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(age_current)
yobs <- yvals[vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(location)]
zobs <- vax_clean %>% filter(age_current<=last_agecurrent) %>% filter(n_doses!="2") %>% pull(value)

#' optional extension: add 2-dose data as minimums?
# vax_obs <- vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2")  # unclear how to incorporate as a minimum

#' apply logistic transform and then centre the data around mean 0
logit_zobs_centred <- logit(zobs) - mean(logit(zobs))

tibble(xobs, yobs, zobs, logit_zobs_centred)
```

## Posterior

As in the 1D case, the posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved independent variables `{xvals, yvals}` and observed independent variables `(xobs, yobs)`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`. In the 1D case, our unobserved was `xvals`, now it is the concatenated vector grid `{xvals, yvals}` of size 320x320 (32 ages in each of 10 provinces).

Plot a sample of posterior curves against the observed data (red points).

```{r}
#| label: posterior_2D

xygrid <- expand_grid(x=xvals, y=yvals)
xyobs <- tibble(x=xobs,y=yobs)

kuo <- generate_2Dksqexp_covmat(xygrid,xyobs,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between unobserved and observed
kou <- generate_2Dksqexp_covmat(xyobs,xygrid,fn=ksqexp,l=l1,b=b)  # 'relationship' pairwise between observed and unobserved
koo <- generate_2Dksqexp_covmat(xyobs,xyobs,fn=ksqexp,l=l1,b=b) + 1e-6*diag(nrow(xyobs))  # protect against non-invertibleness
kuu <- generate_2Dksqexp_covmat(xygrid,xygrid,fn=ksqexp,l=l1,b=b)

#' calculate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit_zobs_centred)  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

```{r}
#| label: plot_posterior_2D

#' draw from the posterior distribution and record in dataframe post2D
post2D <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post2D <- bind_rows(post2D, tibble(draw=i, x_i=xygrid$x, y_i=xygrid$y, post2D=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post2D <- post2D %>% mutate(post2D_constrained=invlogit(post2D + mean(logit(zobs)))) %>%  # add column to return mean and transform back to [0,1]
  mutate(y_label = names(yvals)[match(y_i, yvals)]) %>%  # add column to return province names
  mutate(y_label = factor(y_label, levels=(province_relation %>% pull(location)))) %>%  # order provinces
  select(draw, x_i, y_i, y_label, post2D, post2D_constrained)  # reorder columns

#' plot posterior sample 
yobs_factors <- factor(names(yobs), levels(post2D$y_label))  # convert yobs to factors for use in facet_wrap
post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=logit_zobs_centred, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") +
  facet_wrap(~ y_label)

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) +
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) +
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)
```

As before, we can compare performance by adding back in data points that the model has not seen in green (including for example 2-dose data):

```{r}
#| label: compare2D

post2D %>% ggplot() + geom_point(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.1) + 
  geom_line(aes(x=x_i, y=post2D_constrained, group=factor(draw)), alpha=0.2) + 
  geom_point(data=tibble(x=(vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(age_current)),  # adding in full data from vax_clean
                         y=(vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(value)), 
                         y_label=factor((vax_clean %>% filter(age_current<=last_agecurrent) %>% pull(location)), levels=prov_levels)),
                         aes(x=x,y=y), col="green") + 
  geom_point(data=tibble(x=xobs, y=zobs, y_label=yobs_factors), aes(x=x, y=y), col="red") +
  scale_x_continuous(breaks=xvals, name="current age x_i") + scale_y_continuous(limits=c(0,1)) +
  facet_wrap(~ y_label)

#ggsave(paste0("plots/post2D_l1",l1,"_l2",l2,"_ksqexp","_alldata",".pdf"), height=4.51, width=7.29)
```

Optional: We can also drop any posterior draws that do not satisfy the 2-dose data points as a minimum. Due to the underlying mathematics, this is (*we think*) mathematically equivalent to constraining the prior and proportionally reweighting the remaining valid draws. This turns out to be too strong a constraint so the code snippet is skipped.

```{r}
#| label: set_minimums2D
#| eval: false

# extract 2-dose data from vax_clean, matching column names with post2D
data_2dose <- tibble(x_i=(vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(age_current)),  # adding in full data from vax_clean
                     y_label=factor((vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(location)), levels=prov_levels),
                     value_2dose=(vax_clean %>% filter(age_current<=last_agecurrent & n_doses=="2") %>% pull(value)))
# filter out posterior draws that do not meet the criterion, using all()
post2D_new <- post2D %>% ungroup() %>% left_join(data_2dose, by=c("x_i","y_label")) %>%
  mutate(value_2dose=ifelse(is.na(value_2dose), 0, value_2dose)) %>%  ##UP TO HERE
  mutate(row_valid=(post2D_constrained >= value_2dose)) %>%
  group_by(draw) %>% filter(all(row_valid))
```

# 2D case model extensions

We now wish to improve our basic model by adding information to the prior and experimenting with choice of covariance function and related parameters. First, define helper functions to enable easy running of the 2D model (saved in "R/"):

-   `compute_prior2D()`
-   `plot_prior2D()`
-   `compute_posterior2D()`
-   `plot_posterior2D()`.

The full 2D Gaussian Process model can then be run with `run_GP(vax_dataset)`. The prior plot is currently hidden.

```{r}
#| label: auto2D

#' Run the full 2D Gaussian Process model
#' 
#' First input `vax_dataset` must be specified. All other inputs are optional. Function will return a matrix of posterior draws.
#' Note: x and y are independent variables, with dependent variable z. x is assumed to represent
#' current age, y province, and z vaccine coverage.
#' @param vax_dataset data set of observed vaccine coverage data in our standardised format (tibble)
#' @param first_agecurrent first age for the GP model, i.e. smallest x-variable entry (numeric)
#' @param last_agecurrent last age for the GP model, i.e. largest x-variable entry (numeric)
#' @param yvals numeric vector defining possible range of y-values (numeric)  **do not scale
#' @param k covariance function chosen from the following list: (function) **Functionality not added yet for matern.
#'          - ksqexp,
#'          - kexp,
#'          - kmatern.
#' @param k_name as above, but written as a character string for use in plot titles (character)
#' @param l1 x-lengthscale parameter, for use with covariance functions ksqexp and kexp
#' @param l2 relative lengthscale of x values to y values, for use with covariance functions ksqexp and kexp
#' @param b scale for covariance function determining the output variance
#' @param ndrws specify number of draws to be taken from the prior and posterior distributions
#' @param show_plots if false, plots are hidden
run_GP <- function(vax_dataset, first_agecurrent=1, last_agecurrent=32, yvals=NA, k=ksqexp, k_name=NA, l1=NA, l2=NA, b=1, 
                   ndrws=50, show_plots=T) {
  #' define possible x values
  xvals <- first_agecurrent:last_agecurrent  # a vector of current ages

  #' define possible y values
  province_relation <- readr::read_csv(here::here("data", "3610040201-noSymbol.csv"), show_col_types = F) %>%
    filter(location %in% unique(vax_clean$location))  # province GDP data from StatCan
  prov_levels <- (province_relation %>% pull(location))  # useful shortcut for defining province factors
  if (is.na(yvals)) yvals <- log(province_relation %>% pull(Dollars))  # a vector of provinces -- currently log(GDP)
  names(yvals) <- prov_levels

  #' define covariance and related parameters
  if (is.na(k_name)) k_name <- "ksqexp"  # for plot titles
  if (is.na(l1)) l1 <- 1.5  # x lengthscale
  if (is.na(l2)) l2 <- 3  # relative lengthscale of x values to y values
  yvals <- yvals*l2  # scale y values to the x lengthscale

  #' compute and plot prior
  prior2D <- compute_prior2D(xvals=xvals, yvals=yvals, k=k, l1=l1, ndrws=ndrws, prov_levels=prov_levels, b=b)
  if (show_plots) prior2D %>% plot_prior2D(., k=k_name, k_param1=l1, k_param2=l2) %>% print()

  #' observe data (we may wish to use cNICS here instead of vax_clean)
  xobs <- vax_dataset %>% filter((age_current<=last_agecurrent) & n_doses!="2") %>% pull(age_current)
  yobs <- yvals[vax_dataset %>% filter((age_current<=last_agecurrent) & n_doses!="2") %>% pull(location)]  # this includes scaling by l2
  zobs <- vax_dataset %>% filter((age_current<=last_agecurrent) & n_doses!="2") %>% pull(value)
  logit_zobs_centred <- logit(zobs) - mean(logit(zobs))  # apply logistic transform and centre the data around mean 0
  if (show_plots) tibble(xobs, yobs, zobs, logit_zobs_centred) %>% print()

  #' compute and plot posterior
  post2D <- compute_posterior2D(xvals=xvals, yvals=yvals, xobs=xobs, yobs=yobs, zobs=zobs,
                              k=k, l1=l1, ndrws=ndrws, prov_levels=prov_levels, b=b)
  vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-clean2.csv"), show_col_types = FALSE) %>%  #added for cNICS comparison
    filter(!(pt %in% c("SK","YT","NB")))
  if (show_plots) post2D %>% plot_posterior2D(., xobs=xobs, yobs=yobs, zobs=zobs, k=k_name, k_param1=l1, k_param2=l2, k_param3=b,
                              vax_dataset=vax_clean, last_agecurrent=last_agecurrent, prov_levels=prov_levels) %>% print()
  return(post2D)
}

run_GP(vax_clean)
```

From a quick by-eye model selection, current best GP model parameter values are approximately `l1=1.5, l2=3, b=0.75` for `k=ksqexp`. A more formal model selection (i.e. choosing of parameter values) could also be done, using for example take-one-out cross-validation with log posterior predictive density (lppd) as the 'error' metric (\*\*functionality not yet added).

```{r}
#| label: fit_byeye

# for (l1 in seq(1,3,1)) {  # l1>=2 gives meaningless outputs; l=1.5 (or 1) seems reasonable.
#   for (l2 in seq(1,5,0.5)) {  # l2<=2.5 is too strong a relation; l>=3 seems reasonable for log(GBP) data.
#     for (b in seq(0.5,1,0.25)) {  # b>1 gives too much variance; b=0.75 or 1 seems reasonable at present.
#       run_GP(vax_clean, l1=l1, l2=l2, b=b)
#     }
#   }
# }

run_GP(vax_clean,l1=1.5,l2=3,b=0.75)
```

## Running with cNICS data

Note that the provincial estimates collected in `vax_clean` are from multiple sources with varying data collection methodology. As an alternative, the national cNICS study, although perhaps less trustworthy, does at least have consistent methodology across provinces. We make use of Irena's cleaned, standardised cNICS data and run_GP(). The final plot compares model fit against the full (1+ and 2 dose) provincial dataset `vax_clean`.

Note: Nunavut youngest estimate is described as "treat with caution" in the cNICS online dataset.

```{r}
#| label: run_cNICS

vax_cNICS <- readr::read_csv(here::here("data", "measles_CNICS.csv"), show_col_types = FALSE) %>%
  filter(!(pt %in% c("SK","YT","NB","CA"))) %>%
  mutate(age_current=current_year - year_birth)
run_GP(vax_cNICS, b=0.75, ndrws=15)
```

## Model selection and validation: take-out-one cross validation

We aim to find the optimal choice of model (parameter values `k`, `l1`, `l2`, `b`), and validate with take-one-out cross-validation. An example algorithm is as follows:

-   For a given covariance function k and lengthscale parameters l1, l2, b:

    1.  Pull out one data point.

    2.  Run GP model and compute e.g. sum squared error between posterior mean and the missing data point.

    3.  Repeat for every data point (e.g. 157 model runs for the 1+ provincial data). Sum errors.

-   Select the optimal k/ l1/ l2/ b combination.

Option 1 is to work with the posterior mean at each missing age as described. A better option 2 is to work with a richer goodness-of-fit metric such as the (log) *posterior predictive density* at each missing age instead. This makes use of the full richness of our Gaussian Process model. We hence choose the latter metric.

```{r}
#| label: model_selection

#' ESTIMATE the lppd for one GP model (i.e. for one choice of parameter values of {k, l1, l2, b}).
#' Note: relation of provinces currently uses GDP data.
#' 
#' @param vax_dataset data set of observed vaccine coverage data in our standardised format (tibble)
#' @param last_agecurrent last age for the GP model, i.e. largest x-variable entry (numeric)
#' @param ndrws specify number of draws to be taken from the prior and posterior distributions
#' @param k covariance function chosen from the following list: (function) **Functionality not added yet for matern.
#'          - ksqexp,
#'          - kexp,
#'          - kmatern.
#' @param l1 x-lengthscale parameter, for use with covariance functions ksqexp and/or kexp
#' @param l2 relative lengthscale of x values to y values, for use with covariance functions ksqexp and kexp
#' @param b scale for covariance function determining the output variance
estimate_lppd <- function(vax_dataset, last_agecurrent=30, ndrws=50, k=ksqexp, l1=NA, l2=NA, b=1) {
  # prepare the multiple training datasets (take-one-out)
  vax_dataset <- vax_dataset %>% filter((age_current<=last_agecurrent) & n_doses!="2")  # filter out unused data (2-dose & older ages)
  vax_datasets <- map(1:nrow(vax_dataset), ~ vax_dataset[-.x, ])
  names(vax_datasets) <- paste0("vax_dataset", 1:nrow(vax_dataset))
  
  # run model for each training set and extract 50 draw estimates at the removed (test) point
  lppd_value <- rep(NA,nrow(vax_dataset))
  for (i in 1:nrow(vax_dataset)) {
    xout <- vax_dataset[i, ] %>% pull(age_current)  # age_current of removed data point
    yout <- vax_dataset[i, ] %>% pull(location)  # province of removed data point
    zout <- vax_dataset[i, ] %>% pull(value)  # correct coverage value of removed data point
    ppd <- run_GP(vax_datasets[[i]], show_plots=F, last_agecurrent=last_agecurrent, ndrws=ndrws, k=k, l1=l1, l2=l2, b=b) %>%
      filter(x_i==xout & y_label==yout) %>% pull(post2D_constrained) %>% 
      between(zout - 0.02, zout + 0.02) %>% sum()  # freq of ppd evaluated at true removed value +/-2%
    lppd_value[i] <- ppd  # not taking logs currently; should be log(ppd) eventually
  }
  
  # sum lppd values from each test
  return(sum(lppd_value))
}

example <- estimate_lppd(vax_clean, k=ksqexp, l1=1.5, l2=3, b=1)  # run time: ~65 seconds


#' GP model selection -- identify optimal combination of k, l1, l2, b. Outputs a tibble detailing 
#' each set of model parameters and a goodness-of-fit metric (sum(lppd)) using take-one-out cross validation. 
#' Large lppd is optimal.
#' Note: relation of provinces currently uses GDP data and covariance functions can be ksqexp and/or kexp.
#' 
#' @param vax_dataset data set of observed vaccine coverage data in our standardised format (tibble)
#' @param last_agecurrent last age for the GP model, i.e. largest x-variable entry (numeric)
#' @param ndrws specify number of draws to be taken from the prior and posterior distributions (numeric)
#' @param k_list a named list of covariance functions, e.g. list(ksqexp = ksqexp, kexp = kexp) (named list)
#' @param l1 a vector of x-lengthscale parameters for use with covariance functions ksqexp and/or kexp (vector)
#' @param l2 a vector of relative lengthscale parameters for use with covariance functions ksqexp and/or kexp (vector)
#' @param b a vector of scales for covariance function determining the output variances (vector)
fit_GP <- function(vax_dataset, last_agecurrent=30, ndrws=100, k_list=list(ksqexp=ksqexp), l1=NA, l2=NA, b=1) {
  # first separate k_list into functions and names
  k_tibble <- tibble(k_name = names(k_list), k = unname(k_list))
  
  # compute lppd_sum for all combinations of parameters
  combinations <- tidyr::expand_grid(k_tibble, l1, l2, b)  # functions need to be protected in a list
  combinations <- combinations %>% mutate(lppd_sum = pmap_dbl(list(k, l1, l2, b), ~ estimate_lppd(
    vax_dataset = vax_dataset,
    last_agecurrent = last_agecurrent,
    ndrws = ndrws,
    k = ..1,
    l1 = ..2,
    l2 = ..3,
    b = ..4))) %>% 
    select(-k) %>% mutate(model_no = row_number(), .before = 1)
  
  return(combinations)
}

combinations <- fit_GP(vax_clean, k_list=list(ksqexp=ksqexp), l1=c(1.5,1.6), l2=c(3), b=c(1))
##need to check if this works with kexp
```
