---
title: "Modelling similarly-sparse UK data"
author: "Irena Papst"
format: html
---

```{r}
#| label: setup

library(dplyr)

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates

attach_country <- function(df, country){
    if("pt" %in% names(df)) df <- df |> select(-pt)

    df |>
        mutate(country = !!country) |>
        relocate(country, .before = location)
}

# Canadian data
# - - -
# for Gini model
vax_clean <- readr::read_csv(here::here("data", "measles_vax-coverage-data-cleaned.csv"), show_col_types = FALSE) %>% filter(!(pt %in% c("SK","YT","NB"))) #%>% filter(location %in% unique(pt_lookup()$location))  # filtering out sub-provinces etc estimates
# for low income cleaned data for use with low-income and vaccine hesitancy province relation indicators
vax_cleanLI <- vax_clean %>% filter(!(pt %in% c("NU","YT","NT"))) # "low_income_families" has no data for NU, NT, YT

# England data
# - - -
vax_england <- readr::read_csv("data/ukhsa-chart-download-mmr1-regions.csv") %>%  # data for MMR1 coverage aged 5y from UKHSA (regions of England)
  mutate(age = 5, n_doses="1+") %>% mutate(value=metric_value/100, location=geography, year_report=year) %>%
  mutate(age_current = current_year - year_report + age) %>% select(location, year_report, age, n_doses, value, age_current)

vax_gini <- bind_rows(vax_clean |> attach_country("Canada"), vax_england |> attach_country("England")) |> filter(n_doses == "1+")
vax_LI <- bind_rows(vax_cleanLI |> attach_country("Canada"), vax_england |> attach_country("England"))
```

To strengthen the application of the model to the sparse Canadian data, we want to apply it to a similarly sparse subsample of the very complete UK data.

# Experimental design

The main task here is to emulate the sparsity of the Canadian data to some part of the UK data. The closest comparison is likely to be for England, as it has a similar number of sub-regions as the number of PTs in Canada (exact numbers of PTs depends on the PT-similarity metric as we do not have these metrics for all PTs in our standardized data). The PT-similarity metrics we have for England are: Gini index, percentage of children not in low-income families. (There may be data on vaccine hesitancy, but we have not looked for it yet.)

We will want to run the same experiment (N reps) for each PT similarity metric:

1. Subsample the England data to mimic the CA data sparsity
    - Quantify the CA data as a vector of the number of data points available by PT
    - Sample this vector without replacement 8 times (to match the number of regions in England); call this the "sample size vector".
    - For each region of England, draw V vaccine coverage estimates, where V corresponds to an entry of the sample size vector.
2. Fit the GP model to the subsampled data (code from `gp-model.qmd`)
3. Calculate an accuracy based on left-out data (full set; code from `gp-model.qmd`)
    - Liza mentioned that this falls out of the model fitting process already? I can't remember the details right now...

Then we can summarize accuracy for each PT similarity metric.

# Data exploration and prep

```{r}
#| label: n_obs_gini

n_obs <- vax_gini |> group_by(country) |> select(location) |> distinct() |> count()
```

We start with an analysis of the Gini index model, which has data for `r n_obs |> filter(country == "Canada") |> pull(n)` Canadian PTs and `r n_obs |> filter(country == "England") |> pull(n)` regions of England.

The age ranges in Canadian data do not exactly conincide with those found in the England data:

```{r}
vax_gini |>
    group_by(country, location) |>
    summarize(age_min = min(age_current), age_max = max(age_current))
```

If we filter both down to the largest intersection in age for a fair comparison, and we are left with the following ranges by location: 

```{r}
#| label: filter_data
vax <- vax_gini |>
    group_by(country) |>
    mutate(age_min = min(age_current), age_max = max(age_current)) |>
    ungroup() |>
    mutate(age_min = max(age_min), age_max = min(age_max)) |>
    filter(between(age_current, age_min, age_max))
```

```{r}
vax |>
    group_by(country) |>
    group_split() |>
    purrr::map(\(df){
        df |>
            group_by(location) |>
            summarize(
                age_min = min(age_current), age_max = max(age_current)
            )
    })
```

After filtering, we now only have 6 PTs in the Canadian data, so when testing against the England data, we will only take 6 regions at a time to fit.

# Implementation

We start by getting the sample size of the (filtered) PT data:

```{r}
sample_size <- vax |>
    filter(country == "Canada") |>
    group_by(location) |>
    summarize(n = n())
```

Pull a subset of the England data according to the PT sample sizes:

```{r}
vax_england <- filter(vax, country == "England")

sample_by_location <- tibble(
    location = sample(
        unique(vax_england$location), 
        size = nrow(sample_size), 
        replace = FALSE
    ), 
    n = sample(sample_size$n)
)

vax_input <- vax_england |>
  right_join( # filtering join to subsample regions
    sample_by_location,
    by = "location"
  ) |>
  group_split(location) |>
  purrr::map(\(df) slice_sample(df, n = df$n[1]) |> select(-n)) |>
  purrr::list_rbind()
```

Fit the model to the subsampled data:

```{r}
#| label: fit_england_gini
if(rerun){
    fit <- select_GP(
        vax_input, 
        prov_values="UK-Gini", 
        k_list=list(ksqexp=ksqexp, kexp=kexp),  
        l1=c(1,1.25,1.5,1.75,2,2.25,2.5), 
        l2=c(0.1,0.2,0.5,1,1.5,2,2.5,3.0,3.5,4.0,6.0,8.0,10,25.0,50,75,100,150,200,250,500), 
        b=c(0.5,1,1.5), 
        meas_error=c(0.5,0.1,0.05,0.01,0)
    )

    # temp save since the above takes a while to run
    if(!dir.exists(here::here("results", "tmp"))) dir.create(here::here("results", "tmp"))

    saveRDS(fit, here::here("results", "tmp", "fit_england_gini.rds"))
} else {
    fit <- readRDS(here::here("results", "tmp", "fit_england_gini.rds"))
}

# retreive optimal hyperparameters
fit_opt <- fit |> arrange(desc(lppd_exact)) |> slice(1)
```

To calculate an accuracy score based on the left-out (test) data, it seems reasonable to use the posterior predictive density (PPD) in some way. However, I don't think summing the (L)PPD values will result in an intuitively interpretable accuracy score, because it's not giving us values that are readily interpretable. Point densities are not probabilities. Of course, these values are useful _in comparisson_, _i.e._, if looking for a model that is more or less predictive of left-out data than another one, but it's not straight forward to interpret. 

How should we define the accuracy, then? I discussed this with David Champredon, and we came to the following idea:

1. For each test point, choose an acceptable level of error, _e.g._, within 5% of the true vaccine coverage (the "target interval").
2. Retrieve parameters for the fitted model's (normal) posterior predictive distribution at the test point (distribution mean and variance).
3. Calculate the density's probability mass captured within the target interval (difference between a normal cumulative distribution function evaluated at the edges of the target interval). This tells us how likely it is that the model would have predicted the true value given our chosen error tolerance.

For instance, say true vaccine coverage is 80% for kids aged 7 in London. We choose an error tolerance of 5%, because epidemiologically, these are "similar enough" values. Then our target interval is 77.5%-82.5%. Suppose 70% of the fitted model's probability mass is in that interval; then our model had a 70% chance of predicting the true value with an acceptable amount of error. The more error we tolerate, the higher the accuracy, but this way the accuracy is informed directly by the real-world application, and so it can be interpretted. 

The proceedure above could be repeated for each test point, and then we could compute summary statistics on it: an average accuracy score along with a range or confidence interval.

# Outstanding issues

* Can we talk through retrieving the parameters for the fitted model's posterior predictive distribution at the test points? from the code?
    - `compute_lppd()` has it, look to [this issue](https://github.com/phac-modelling-hub/immunity-landscape/issues/4#issuecomment-3410698100) for the math
    - modularize the computation of the means and vars to be able to extract
    - find indices (row numbers) of test points in input dataset to just pull out means and variants we want