---
title: "Building a first Gaussian Process model"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

## Set Up

Load functions and packages. Read the standardized data and filter to province-level. Fill in age for rows where it is not specified.

As a first pass, we consider a subset of the data: Ontario, reported year 2023, 2-doses, for ages 5-17. Call this mini dataset `vax_ON`.

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(mnormt) # for multivariate normal
library(LaplacesDemon) # for logistic transform

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read data
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(pt_lookup()$location)) |> filter(pt!="CA") # filtering out sub-provinces etc and CA estimate
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970

#' prep data
vax_main  <- vax_raw %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%  # calc age
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>% filter(location == location.pt) %>% select(-location.pt)

vax_ON <- vax_main %>% filter(location=="Ontario") %>% filter(year_report==2023) %>% filter(n_doses %in% c("2"))  # *subset of doses for the toy model
```

IP: This is a small point, but there are definitely questions about whether the ON data for younger ages is super trustworthy. I believe the data come from school administration data, and because of how these vaccines get reported to schools, there has been a bit of an interruption in that from the pandemic, which means that it's like the coverage estimates in ON for younger ages are artificially low. This is only relevant if we start to inform the model with other data that can help predict coverage; you may want to switch to the AB data in the future.

## A note on scales

Modelling fitting is done using Gaussian distributions, whose domain is the entire real line. However, the vaccine coverage data is restricted to $[0,1]$ (Though in practice, they're not on the boundaries of this interval.) Thus, we transform the vaccine coverage estimates using a an inverse logit function to map it from $(0,1)$ to $(-\infty, \infty)$, perform the model fitting on the real line, and then transform the predictions back from $(-\infty, \infty)$ to $(0,1)$.

## Define covariance function / covariance matrix

Use the squared exponential covariance function, `k`, to begin with and assume lengthscale `l=2`. Our dataset has 13 inputs/ independent variables (ages 5-17), so our covariance matrix is a 13x13 matrix defining the 'relationships' between each pair of ages.

IP: What are the implications of using the square exponential covariance function? What kind of data would you expect to meet this assumption? What are the alternatives?

```{r}
#| label: covariance

ksqexp <- function(r,l) {  # this is also defined in ksqexp.R
  exp(-r^2 / (2*l^2))
}

xvals <- (vax_ON %>% mutate(age = pull_first(age)) %>% pull(age))  # a vector of ages (these x_i's are equally spaced)
npts <- length(xvals)
l <- 2
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>%  # evaluation of cov function pairwise for ages x1_i and x2_i (i=1,..,n)
  pull(k) %>% matrix(nrow=npts)
# could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (ages).

IP: Since the mean is zero, is this a model for the *error*? Can you pls clarify what this is a prior of?

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals) + geom_point(alpha=0.3) + geom_line(alpha=0.2)
```

## Introduce observed data

For this mini dataset `vax_ON` (Ontario, year 2023, 2-dose, ages 5-17), we have a complete dataset of 13 data points. First expose the model to 5 data points (to see how well it can predict the remainder). We will use the logistic transform of y (`logit(yobs)`) which maps from \[0,1\] to \[-inf,inf\].

```{r}
#| label: observe_data

# observe coverage data for age 6, 9, 11, 15, 16
xobs <- as.numeric(vax_ON %>% filter(age %in% c("6","9","11","15","16")) %>% pull(age))   # can optionally add an error term here
yobs <- vax_ON %>% filter(age %in% c("6","9","11","15","16")) %>% pull(value)

tibble(xobs, yobs, logit(yobs))
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationships' between unobserved ages `xvals` and observed ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

IP: can you either explain a bit more around the "calculation of individual covariance matrices to determine the pairwise 'relationships' between unobserved and observed ages"? I'm a little confused about how we're getting information about unobserved ages. Is it just from draws of the prior?

IP: Can you pls point me to some reading material and/or your notes where the posterior covariance matrix is derived?

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,fn=ksqexp,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,fn=ksqexp,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,fn=ksqexp,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,fn=ksqexp,l=l)

# calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%(logit(yobs))  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed logit-transformed data. The posterior function can be used to provide estimates of our missing coverage data if inverse-transformed back to a \[0,1\] interval (second plot).

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}
post <- post %>% mutate(post_constrained=invlogit(post))  # add column to transform back to [0,1] interval

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals) + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=logit(yobs)), aes(x=x,y=y), col="red")

post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```
We can add in the remaining true data (which the model has not seen) to evaluate its performance. The true coverage values are marked as red (seen) and green (not seen) points.
```{r}
#| label: compare_to_data
post %>% ggplot() + geom_point(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.3) + scale_x_continuous(name="age x_i", breaks=xvals) + scale_y_continuous(limits=c(0,1)) + geom_line(aes(x=x_i,y=post_constrained,group=factor(draw)), alpha=0.2) + geom_point(data=tibble(x=xvals, y=(vax_ON %>% pull(value))), aes(x=x,y=y), col="green") + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
ggsave(paste0("plots/toy-post_l",l,"_ksqexp",".pdf"), height=4.51, width=7.29)
```

## Extensions

To improve this toy model, the next step would be to experiment with choice of covariance function `k` and lengthscale `l`.

We could then extend to a larger portion of the dataset by adding a second dimension (e.g. year). Geometrically, this would move from a posterior curve as on the above plot, to a posterior *surface* in a 3d plot / cube. We would also want to add a dimension for province, giving a posterior hypersurface in a 4d hypercube... The final model would aim to produce separate hypersurfaces for 1-dose, 1+-dose, and 2-dose. Geometry aside, this just equates to providing an estimate of coverage at each combination of age, year, province, and dose.

<<<<<<< HEAD
Here is an example of adding in year_report:
```{r}
#| label: posterior_surface

vax_ON2 <- vax_main %>% filter(location=="Ontario") %>% filter(n_doses %in% c("2","2+"))  # toy dataset
# now both x and y are inputs, with output z
xvals <- vax_ON2 %>% mutate(age = pull_first(age)) %>% pull(age) %>% unique() %>% sort()  # a vector of ages
yvals <- vax_ON2 %>%  pull(year_report) %>% unique() %>% sort() # a vector of year_report
npts <- length(xvals)
l <- 1

#---
covmat <- expand_grid(x1=xvals, y1=yvals, x2=xvals, y2=yvals) %>% 
  mutate(k=ksqexp(sqrt((x1-x2)^2 + (y1-y2)^2), l)) %>%
  # mutate(k=kexp(sqrt((x1-x2)^2 + (y1-y2)^2), l)) %>%
  pull(k) %>% 
  matrix(nrow=npts^2)

covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>%  # evaluation of cov function pairwise for ages x1_i and x2_i (i=1,..,n)
  pull(k) %>% matrix(nrow=npts)
```

=======
IP: Overall, this is a great first start! I would be hesitant to push the lengthscale too far because I think vaccine coverage by cohort is fairly localized to that cohort. I'd like to think a bit more about the implications of the covariance function `k` and if we can better-inform it with some additional info to help predict for missing ages.
>>>>>>> 21292faec885c9016c1090f564477553a3cad471
