---
title: "Building a first Gaussian Process model"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: false
---

# Set Up
Load functions and packages. Read the standardized data and filter to province-level. Fill in age for rows where it is not specified. 

As a first pass, we consider a subset of the data: Ontario, reported year 2023, 2-doses, for ages 5-18. Call this mini dataset `vax_ON`.
```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(mnormt) # for multivariate normal

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read data
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(pt_lookup()$location)) |> filter(pt!="CA") # filtering out sub-provinces etc
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970

#' prep data
vax_main  <- vax_raw %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%  # calc age
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>% filter(location == location.pt) %>% select(-location.pt) #%>% filter(age!="18+")  # remove adult data (for now)

vax_ON <- vax_main %>% filter(location=="Ontario") %>% filter(year_report==2023) %>% filter(n_doses %in% c("2","1+"))  # *re-add 2+ later*
```

# Define covariance function / covariance matrix
Use the squared exponential covariance to begin with and assume lengthscale `l=1`. Our dataset has 14 parameters (coverage at ages 5-18), so our covariance matrix is a 14x14 matrix defining the 'relationship' between each pair of parameters.
```{r}
ksqexp <- function(r,l) {
  exp(-r^2 / (2*l*2))
}

npts <- 14
l <- 1
yvals <- seq(from=0,to=1,length.out=npts)
covmat <- expand_grid(y1=yvals,y2=yvals) %>% mutate(k=ksqexp(abs(y1-y2),l)) %>%  # pairwise evaluation of cov fn
  pull(k) %>% matrix(nrow=npts)
```
Q: Should I work with _current_ age?

# Define prior distribution (N(0,cov)) and draw a sample
Define a prior ~ Normal(0,cov). It is uniquely defined with mean 0 and variance equal to our covariance matrix.
Take 50 draws from the prior and plot.
```{r}
prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, y=yvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}

```

# Introduce observed data
For this mini dataset `vax_ON` (Ontario, year 2023, 2-dose, ages 5-18), we have a complete dataset of ~14 data points. First expose the model to 4 data points (and see how well it can predict the remainder).
```{r}


```

# Compute posterior distribution
```{r}

```

# Plot posterior against the data
```{r}

```

