---
title: "Building a first Gaussian Process model"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: false
---

## Set Up

Load functions and packages. Read the standardized data and filter to province-level. Fill in age for rows where it is not specified.

As a first pass, we consider a subset of the data: Ontario, reported year 2023, ~2-doses, for ages 5-18. Call this mini dataset `vax_ON`.

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(mnormt) # for multivariate normal

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read data
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(pt_lookup()$location)) |> filter(pt!="CA") # filtering out sub-provinces etc and CA estimate
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970

#' prep data
vax_main  <- vax_raw %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%  # calc age
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>% filter(location == location.pt) %>% select(-location.pt)

vax_ON <- vax_main %>% filter(location=="Ontario") %>% filter(year_report==2023) %>% filter(n_doses %in% c("2","1+"))  # *subset of doses (2/1+) for the toy model
```

## Define covariance function / covariance matrix

Use the squared exponential covariance to begin with and assume lengthscale `l=1`. Our dataset has 14 inputs/ independent variables (ages 5-18), so our covariance matrix is a 14x14 matrix defining the 'relationships' between each pair of ages.

```{r}
#| label: covariance

ksqexp <- function(r,l) {  # this is also defined in ksqexp.R
  exp(-r^2 / (2*l^2))
}

npts <- 14
l <- 1
xvals <- seq(from=5,to=18,length.out=npts)  # a vector of ages *(x_i's are equally spaced and between [0,1] for now)
covmat <- expand_grid(x1=xvals,x2=xvals) %>% mutate(k=ksqexp(abs(x1-x2),l)) %>%  # evaluation of cov function pairwise for ages x1_i and x2_i (i=1,..,n)
  pull(k) %>% matrix(nrow=npts)
# could also use function generate_ksqexp_covmat() here
```

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. x_i is the vector of inputs (ages).

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, x_i=xvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=x_i,y=prior,group=factor(draw))) + scale_x_continuous(breaks=xvals) + geom_point(alpha=0.5) + geom_line(alpha=0.3)
```

## Introduce observed data

For this mini dataset `vax_ON` (Ontario, year 2023, 2-dose, ages 5-18), we have a complete dataset of \~14 data points. First expose the model to 5 _toy_ data points (to see how well it can predict the remainder):

```{r}
#| label: observe_data

# observe coverage data for age 6, 9, 10, 15, 16
xobs <- c(xvals[2],xvals[5],xvals[6],xvals[11],xvals[12])   # can optionally add an error term here
yobs <- rnorm(length(xobs),0,1) # *arbitrary 'height'/ coverage for now
# xobs <- as.numeric(vax_ON %>% filter(age %in% c("6","9","10","15","16")) %>% pull(age))
# yobs <- vax_ON %>% filter(age %in% c("6","9","10","15","16")) %>% pull(value)

tibble(xobs, yobs)
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationship' between unobserved ages `xvals` and observed ages `xobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(xvals,xobs,l=l)  # 'relationship' pairwise between unobserved and observed ages
kou <- generate_ksqexp_covmat(xobs,xvals,l=l)  # 'relationship' pairwise between  observed and unobserved ages
koo <- generate_ksqexp_covmat(xobs,xobs,l=l) + 1e-6*diag(length(xobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(xvals,xvals,l=l)

# calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%yobs  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed data. The posterior function can be used to provide estimates of our missing data.

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, x_i=xvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}

post %>% ggplot() + geom_point(aes(x=x_i,y=post,group=factor(draw)), alpha=0.5) + scale_x_continuous(breaks=xvals) + geom_line(aes(x=x_i,y=post,group=factor(draw)), alpha=0.3) + geom_point(data=tibble(x=xobs, y=yobs), aes(x=x,y=y), col="red")
```
