---
title: "Building a first Gaussian Process model"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

## Set Up

Load functions and packages. Read the standardized data and filter to province-level. Fill in age for rows where it is not specified.

As a first pass, we consider a subset of the data: Ontario, reported year 2023, 2-doses, for ages 5-18. Call this mini dataset `vax_ON`.

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)
library(mnormt) # for multivariate normal

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))  # load functions

#' read data
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(pt_lookup()$location)) |> filter(pt!="CA") # filtering out sub-provinces etc and CA estimate
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)
popsize <- readr::read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970

#' prep data
vax_main  <- vax_raw %>%
  mutate(age = ifelse(is.na(age) & !is.na(year_report) & !is.na(year_birth), year_report - year_birth, age)) %>%  # calc age
  left_join(pt_lookup(), join_by(pt), suffix = c("", ".pt")) %>% filter(location == location.pt) %>% select(-location.pt) #%>% filter(age!="18+")

vax_ON <- vax_main %>% filter(location=="Ontario") %>% filter(year_report==2023) %>% filter(n_doses %in% c("2","1+"))  # *re-add 2+ later
```

## Define covariance function / covariance matrix

Use the squared exponential covariance to begin with and assume lengthscale `l=0.1`. Our dataset has 14 parameters (coverage at ages 5-18), so our covariance matrix is a 14x14 matrix defining the 'relationships' between each pair of parameters.

```{r}
#| label: covariance

ksqexp <- function(r,l) {  # this is also defined in ksqexp.R
  exp(-r^2 / (2*l^2))
}

npts <- 14
l <- 0.1
yvals <- seq(from=0,to=1,length.out=npts)  # *model params equally spaced, and between [0,1] for now
covmat <- expand_grid(y1=yvals,y2=yvals) %>% mutate(k=ksqexp(abs(y1-y2),l)) %>%  # pairwise evaluation of cov function for vectors y1, y2 (with elements/ model parameters y1_i, y2_i)
  pull(k) %>% matrix(nrow=npts)
# could also use function generate_ksqexp_covmat() here
```

Q: Should I work with *current* age?

## Define prior distribution (N(0,cov)) and draw a sample

Define a prior \~ Normal(0,cov). It is uniquely defined with mean 0 and variance equal to our covariance matrix. Take 50 draws from the prior and plot. y_i is the vector of model parameters.

```{r}
#| label: prior

prior <- tibble()
for(i in 1:50) {
  prior <- bind_rows(prior, tibble(draw=i, y_i=yvals, prior=rmnorm(1, rep(0,npts), covmat+1e-6*diag(npts))))  # each rmnorm is a random draw from the (multivariate) prior
}
prior %>% group_by(draw) %>% ggplot(aes(x=y_i,y=prior,group=factor(draw))) + geom_point(alpha=0.5) + geom_line(alpha=0.3)
```

## Introduce observed data

For this mini dataset `vax_ON` (Ontario, year 2023, 2-dose, ages 5-18), we have a complete dataset of \~14 data points. First expose the model to 5 data points (and see how well it can predict the remainder).

```{r}
#| label: observe_data

# observe coverage data for age 6, 9, 10, 15, 16
yobs <- c(yvals[2],yvals[5],yvals[6],yvals[11],yvals[12])  # can optionally add an error term here
# yobs <- vax_ON %>% filter(age %in% c("6","9","10","15","16")) %>% select(age,value)
yobs_T <- rnorm(length(yobs),0,1) # *arbitrary 'height' data points for now
```

## Compute posterior distribution

The posterior is uniquely defined by its posterior mean and posterior covariance matrix. First, calculate individual covariance matrices to determine the pairwise 'relationship' between unobserved `yvals` and observed data `yobs`. Then calculate the posterior covariance matrix as `kuu - kuo%*%solve(koo)%*%kou`.

```{r}
#| label: posterior

kuo <- generate_ksqexp_covmat(yvals,yobs,l=0.1) # 'relationship' between unobserved yvals and observed yobs
kou <- generate_ksqexp_covmat(yobs,yvals,l=0.1) # 'relationship' between  observed yobs and unobserved yvals
koo <- generate_ksqexp_covmat(yobs,yobs,l=0.1) + 1e-6*diag(length(yobs))  # protect against non-invertibleness
kuu <- generate_ksqexp_covmat(yvals,yvals,l=0.1)

# calcuate posterior mean and posterior cov matrix
post_mean <- kuo%*%solve(koo)%*%yobs_T  # conditional mean
post_covmat <- kuu - (kuo%*%solve(koo)%*%kou)  # conditional variance
```

## Plot posterior against the data

Take 50 draws from the posterior and plot against the observed data.

```{r}
#| label: plot_posterior

post <- tibble()
for(i in 1:50) {  # take 50 posterior draws
  post <- bind_rows(post, tibble(draw=i, y_i=yvals, post=rmnorm(1, post_mean, post_covmat + 1e-6*diag(nrow(post_covmat)))))  # each rmnorm is a random draw from the (multivariate) posterior
}

post %>% ggplot() + geom_point(aes(x=y_i,y=post,group=factor(draw)), alpha=0.5) + geom_line(aes(x=y_i,y=post,group=factor(draw)), alpha=0.3) + geom_point(data=tibble(x=yobs, y=yobs_T), aes(x=x,y=y), col="red")
```
