---
title: "LOPO vs LOO: are they equivalent?"
author:
  - name: Liza Hadley
    email: lh667@cam.ac.uk
editor: visual
echo: true
---

Written by Claude. 

## Question

Is `lppd_by_province` from `compute_lppd_lopo` the same as the sum of individual LOO lppd values from `compute_lppd` for the same test points?

**Theoretical expectation: no, for two reasons.**

1.  **Different training sets.** LOO leaves out one point at a time, so when predicting point $i$ in province $p$, it still trains on all *other* ages within $p$. LOPO leaves out all of province $p$, so predictions rely entirely on cross-province covariance — no same-province training data is used.

2.  **Different predictive distributions.** LOO evaluates a univariate Gaussian for each point independently and sums the log-densities. LOPO evaluates a single *joint* multivariate Gaussian over all $n_p$ held-out points in the province, which accounts for residual within-province correlation after conditioning on the rest of the data.

## Setup

```{r}
#| label: setup
library(here)
source(here::here("results", "setup.R"))
source(here::here("results", "results_setup.R"))
```

## Run LOPO

```{r}
#| label: lopo
cfg <- model_configs$Gini
bp  <- best_params$Gini

lopo_result <- compute_lppd_lopo(
  cfg$vax_dataset,
  prov_values = cfg$prov_values,
  k           = get(bp$k_name),
  l1          = bp$l1,
  l2          = bp$l2,
  b           = bp$b,
  meas_error  = bp$meas_error
)

lopo_result$lppd_by_province
```

## Compute per-point LOO lppd values

Replicate the internals of `compute_lppd` to recover individual lppd values (the function currently only returns the total sum).

```{r}
#| label: loo
vax_filt <- cfg$vax_dataset %>%
  filter(age_current >= 5, age_current <= 21, n_doses != "2+")

xobs <- vax_filt %>% pull(age_current)
prov_vals <- extract_province_relation(cfg$prov_values,
                                       vax_dataset = cfg$vax_dataset) * bp$l2
yobs <- prov_vals[vax_filt %>% pull(location)]
zobs <- vax_filt %>% pull(value)
centring_term      <- mean(logit(zobs))
logit_zobs_centred <- logit(zobs) - centring_term

xyobs  <- tibble(x = xobs, y = yobs)
koo    <- generate_2Dksqexp_covmat(xyobs, xyobs, fn = get(bp$k_name),
                                   l = bp$l1, b = bp$b) + 1e-12 * diag(nrow(xyobs))
errmat <- diag(bp$meas_error^2, nrow(koo))
Kinv   <- solve(koo + errmat)
KinvZ  <- Kinv %*% logit_zobs_centred

loo_lppds <- numeric(nrow(vax_filt))
for (i in seq_len(nrow(vax_filt))) {
  ppd_var_i  <- 1 / Kinv[i, i]
  ppd_mean_i <- logit_zobs_centred[i] - KinvZ[i] / Kinv[i, i]
  loo_lppds[i] <- dnorm(logit_zobs_centred[i],
                        mean = ppd_mean_i, sd = sqrt(ppd_var_i), log = TRUE)
}
```

## Compare per province

```{r}
#| label: comparison
comparison <- vax_filt %>%
  mutate(loo_lppd = loo_lppds) %>%
  group_by(location) %>%
  summarise(loo_sum   = sum(loo_lppd),
            n_points  = n(),
            .groups   = "drop") %>%
  mutate(
    lopo      = lopo_result$lppd_by_province[location],
    diff      = lopo - loo_sum,
    mean_loo  = loo_sum / n_points,
    mean_lopo = lopo    / n_points
  ) %>%
  arrange(location)

comparison
```
