---
title: "Quantifying population susceptibility to measles in Canada using vaccine coverage estimates"
editor: visual
author:
  - name: Liza Hadley
    corresponding: true
    email: lh667@cam.ac.uk; liza.hadley@colorado.edu
    affiliations:
      - id: cu
        name: University of Colorado Boulder
        city: Boulder
        state: CO
        country: USA
        postal-code: 80309
      - name: University of Cambridge
        city: Cambridge
        country: UK
        postal-code: CB3 0ES
  - name: Rachael M. Milwid
    corresponding: false
    affiliations:
      - id: phac
        name: Public Health Agency of Canada
        city: Ottawa
        state: ON
        country: Canada
        postal-code:
  - name: Valerie Hongoh
    corresponding: false
    affiliations:
      - ref: phac
  - name: Rania Wasfi
    corresponding: false
    affiliations:
      - ref: phac 
  - name: Stephen M. Kissler
    corresponding: false
    affiliations:
      - ref: cu
    acknowledgements: "SMK is supported by a Mathematical Modeling of Policy Options for Evolving Public Health Challenges award from the U.S. National Science Foundation and Centers for Disease Control and Prevention (DMS- 2436340)."
  - name: Irena Papst
    corresponding: false
    affiliations:
      - ref: phac
keywords:
  - Measles
  - vaccination
  - coverage
  - statistical modelling
  - Canada
abstract: |
  Background & aims of study: The 2024/2025 global measles outbreak underscored ongoing changes in the immunity landscape of some vaccine-preventable diseases (VPDs), which have been occurring steadily over the last two decades. In Canada, there were 3170 reported measles cases between January and early June 2025, far exceeding the average annual case count of <200. This significant outbreak has underscored the need to better understand the present state of population immunity to measles in Canada, an essential input parameter for outbreak response models. Most population immunity in this setting is derived from routine childhood vaccination, but the best available data (vaccine coverage) is often collected cross-sectionally to monitor population adherence to recommended immunization schedules, and not to inform the current state of the population’s susceptibility to measles. This study hence aims to develop statistical methods to adapt available measles vaccine coverage estimates into reliable estimates of present-day immunity in Canada by age and province/territory.
  
  Methods & results: This modelling project centres on a classic problem in both human and veterinary epidemiology - making inferences with incomplete epidemiological data. First, a standardised dataset for vaccine coverage by province/territory was curated from existing publicly-available datasets and reports. Due to significant missing data points by year and age in many of the provinces, this project then uses statistical modelling to approximate current vaccination coverage in missing cohorts, to create a more complete ‘modelled dataset’ of estimated coverage by province/territory, age, and number of doses. This is used to estimate population immunity for current outbreak response models. We compare and contrast two statistical methods to identify strengths and weaknesses of each approach.
  
  Implications: We develop methodology for estimating vaccination coverage in the absence of population immunity data, to support the ongoing measles outbreak modelling response work. While this is primarily a measles epidemiology project, we build the associated code/tools such that the methodology can be applied to other vaccine-preventable diseases for use in future outbreak settings.
key-points:
  - ...All Vaccine articles need a 'highlights'. 3-5 key points, 85 characters each.
  - ...hightlight 2
  - ...
funding: "Funding text to be stated here."
conflict-of-interest: "SMK has received consulting fees from Moderna Therapeutics. All other authors declare XXX. Edit as needed."
bibliography: []
number-sections: true
---

## Introduction

*Vaccine* articles have a word limit of 5000, with up to 10 figures+tables and up to 50 references. The abstract needs to be under 300 words - ours is currently 326. The journal also allows/ encourages a graphical abstract.

Introductory paragraph

Background on Measles/ Measles in Canada

Paragraph on the research problem - limited serological data, vaccine data only by province & very heterogeneous. Flag susceptibility vs vaccination. Flag the need for modelling (preliminary linear stuff didn't work/ datasets all have different methodology!).

Paragraph outlining what this article adds ("Using XXX, this article estimates ...")

## Material and methods

### Data

#### Provincial data sources

Explain data sources and standardisation. Will we make the standardised dataset available on GitHub?

#### National data sources

Flag nuances in the data, perhaps as a table: 1/2 vs 1+/2+ dose, differing vaccination schedules between provinces (second dose administered at 18mnths vs 4-6y), level of reporting, methodology (whether self-reported or from GP data if known?).

Discuss ageing of estimates.

### Model 1: Gaussian Process model

Hyperparameters used in the Gaussian Process model are shown in @tbl-gp-modelparameters. These were chosen from hyperparameter optimisation using leave-one-out cross-validation (see @sec-fitting for details).

```{r}
#| label: tbl-gp-modelparameters
#| tbl-cap: "Hyperparameters used in the GP model."
#| tbl-number: 1
#| echo: FALSE

GPmodelparameters_table <- data.frame(
  Hyperparameter = c("Covariance function", "Age lengthscale", "Relative province lengthscale", "Weight"),
  Notation = c("k", "l1", "l2", "b"),
  Value = c("Squared Exponential", 2.0, 0.5, 2.0)  # example values
)

knitr::kable(GPmodelparameters_table)
```

### Model 2: GLM with XXX

### Model fitting, selection, and validation procedures {#sec-fitting}

For the (non-parametric) Gaussian process model, there is no model fitting stage. Model selection and validation were performed using leave-one-out cross-validation, with log posterior predictive density (lppd) as the error metric at test points. Model selection and validation were first performed on the provincial dataset (@tbl-gp-modelselectionresults). The combination of plausible hyperparameter values with largest absolute lppd were used in analysis (@tbl-gp-modelparameters). Model selection was also repeated with the national cNICS dataset (@tbl-gp-modelselectionresults2) and different indicators for relation of provinces ().

```{r}
#| label: tbl-gp-modelselectionresults
#| tbl-cap: "text"
#| tbl-number: 2
#| echo: FALSE

GPcombinations_GDP <- readr::read_csv(here::here("results", "GPcombinations_GDP.csv"), show_col_types = F)
knitr::kable(GPcombinations_GDP)  # move to supplementary?
```

```{r}
#| label: tbl-gp-modelselectionresults2
#| tbl-cap: "text"
#| tbl-number: 3
#| echo: FALSE

GPcombinations_GDP_cNICS <- readr::read_csv(here::here("results", "GPcombinations_GDP_cNICS.csv"), show_col_types = F)
knitr::kable(GPcombinations_GDP_cNICS)  # move to supplementary?
```

```{r}
#| label: tbl-gp-modelselectionresults3
#| tbl-cap: "text"
#| tbl-number: 3
#| echo: FALSE

GPcombinations_lowincome <- readr::read_csv(here::here("results", "GPcombinations_lowincome.csv"), show_col_types = F)
knitr::kable(GPcombinations_lowincome)  # move to supplementary?
```

### Model comparison

## Results

### Vaccine coverage data

Plot/s showing the curated data.

```{r}

```

### Vaccine coverage estimates

Optimal hyperparameters for the Gaussian Process model are noted in Table X.

Example results plots.

![GP model results, using provincial 1+ dose vaccine coverage data. GDP data is used for relation of provinces. Plots are overlaid with national cNICS data in green (unobserved by the model).](plots/bestGPmodel_GDP.pdf)

![GP model results, using national cNICS vaccine coverage data. GDP data is used for relation of provinces. Plots are overlaid with provincial coverage data (unobserved by the model), 1+ and 2-dose, in green.](plots/bestGPmodel_GDP_cNICS.pdf)

![GP model results, using provincial 1+ dose vaccine coverage data. Health inequalities data is used for relation of provinces (percentage of children in low-income families). Plots are overlaid with national cNICS data in green (unobserved by the model).](plots/bestGPmodel_lowincome.pdf)

### Model comparison

### Additional sensitivity analysis and uncertainty quantification

## Discussion

Paragraphs to discuss findings

Paragraph/s to compare to other literature

Miscellaneous discussion points - list here as we think of them: Immunity vs Vaccination (they are not the same); flag that we haven't tested the model on other countries/pathogens;

Limitations paragraph: note older ages not explicitly considered/note the assumptions on older ages; some discussion of relation of provinces; ageing of estimates.

Conclusions paragraph
