---
title: "measles_prelim-vax-coverage-data-processing"
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| message: false

library(dplyr)
library(tidyr)
library(ggplot2); theme_set(theme_bw())
library(zoo)
library(stringr)
library(readr)
library(forcats)

# Source functions
invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))

# Define functions

#' Add full sequence of age groups, adding new limits
#' 
#' @param df a data frame where ages are stored in the `age_current` column
#' @param new_ages new ages to add before completing the sequence
#' @return a data frame
expand_age <- function(df, new_ages){
  df |>
  # attach new ages
  group_modify(\(x, ...) add_row(x, age_current = new_ages)) |>
  # fill in all missing ages from expanded range
  complete(age_current = full_seq(age_current, period = 1))
}

#' Parse age groups into one observation per single-year age
#'
#' @param df data frame
#' @param col column to transform, as a symbol
parse_age_groups <- function(df, col, last_age = 21){
  df |>
    rowwise() |> 
    mutate(
      {{col}} := list(pull_first({{col}}):pull_last({{col}}, last_age))
    ) |>
    unnest_longer({{col}})
}
```

We've collected a large dataset of vaccine coverage data (see `measles_vax-coverage-data.qmd`). Here we carry out data processing tasks as outlined in the methods of our ms.

## Read data

```{r}
#| label: read_data

# read in data and filter to province-level
PT_lookup <- readr::read_csv(here::here("data", "pt_lookup.csv"), show_col_types = FALSE)
popsize <- read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(PT_lookup$nm_en))

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
```

## Remove low Nunavut point

```{r}
vax_raw <- vax_raw |> filter(!(pt == "NU" & year_report == 2021))
```

## Parse ages and doses

```{r}
#| label: parse_ages_and_doses

vax_parsed <- vax_raw |>
  # parse aNICS "18+" as 18
  mutate(age = if_else(source == "aNICS", "18", age)) |>
  # parse age groups into single-year ages
  mutate(age = if_else(is.na(age), as.character(year_report-year_birth), age)) |> # fill in age at report where not present
  parse_age_groups(age) |>
  # round non-integer ages (QC)
  mutate(age = round(age))
```

## Age the estimates to current year

```{r}
#| label: age_estimates

vax_aged <- vax_parsed |> 
  mutate(age_current = age + (current_year - year_report))
```

## Collapse exactly 1 and exactly 2 dose types to 1+

For data with exactly 1 dose and exactly 2 dose, Ontario's is the only data in which the sum of the coverage estimates across dose types (0, 1, 2) is 1, *i.e.* the estimates are consistent and therefore valid. (See `measles_prelim-vax-coverage-modelling.qmd` take 2 approach for full details).

Thus, for ON we will sum 1 and 2 dose coverage as reported to get a true 1+ estimate.

NB: Note that the Ontario data came from two sources, one with 0, 1, and 2 doses, and another with estimates for 2 doses only. For consistency, we only use the multiple doses data.

```{r}
#| label: take2_filter_ON

vax_ON <- vax_aged %>%
  filter(
    pt == "ON",
    source == "PT",
    n_doses %in% c("1","2+") # keep 1 and 2+ doses (compatible estimates)
  ) %>%
  # sum the uptake for each year to give estimates for 1+
  group_by(pt, location, year_report, year_birth, age, age_current, source) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  mutate(n_doses = "1+")  # (*)

# replace only the ON PT data with the version above
vax_new <- vax_aged %>%
  filter(!(pt == "ON" & source == "PT")) %>%
  bind_rows(vax_ON)
```

And then relabel all estimates of 1/1+ to "1+" (because after dealing with Ontario, all 1 and 1+ data does truly represent "at least one dose" now). Also relabel all 2/2+ to "2+" as these data all now represent "two or more doses".

```{r}
#| label: correct_dose_labels

vax_corrected_labs <- vax_new %>% 
  mutate(n_doses = if_else(n_doses == "1", "1+", n_doses)) %>% 
  mutate(n_doses = if_else(n_doses == "2", "2+", n_doses))
```

## De-duplicate

Between PT-reported data and estimates from national surveys, prioritize PT-reported data per age and dose number. If duplicates still exist (within PT-reported data), take the most recent estimate.

```{r}
#| label: take2_dedupe

vax_dedupe <- vax_corrected_labs %>%
  # prioritize PT-reported data over national surveys
  mutate(source_priority = if_else(source == "PT", 1L, 2L)) %>%
  group_by(pt, location, age_current, n_doses) %>%
  filter(source_priority == min(source_priority)) %>%
  # if still duplicates within PT, keep the most recent estimate
  filter(year_report == max(year_report)) %>%
  ungroup() %>%
  select(-source_priority)
```

## Fill in year_birth, where missing

(Unsure if this is still used, but keeping in as not to break anything...)

```{r}
#| label: year_birth_and_aNICS

vax_clean <- vax_dedupe %>% 
  mutate(year_birth = ifelse(is.na(year_birth) & !is.na(year_report) & !is.na(age), year_report - age, year_birth))
```

## Filter age range

```{r}
#| label: filter_age_range

first_agecurrent <- 5
last_agecurrent <- 21
vax_clean <- vax_clean %>% filter(
    between(age_current, first_agecurrent, last_agecurrent)
)
```

## Plot and export

Here's what the data looks like:

```{r}
#| label: plot

vax_clean |> 
  ggplot(aes(x = age_current, y = value)) +
  geom_point(aes(shape = n_doses, colour = n_doses)) + 
  facet_wrap(~location) +
  xlim(1, 21)
```

We can now export `vax_clean` for use in `gp_model.qmd`:

```{r}
#| label: export_vax_clean

# model input data (1+ doses)
vax_input <- vax_clean |> dplyr::filter(n_doses != "2+") 
readr::write_csv(vax_input, here::here("data", "measles_vax-coverage-data-cleaned.csv"))
# 2+ doses for plot overlay
vax_clean |> dplyr::filter(n_doses == "2+") %>% readr::write_csv(here::here("data", "measles_vax-coverage-data-cleaned_2plus.csv"))
# model input data without adults
vax_input |> filter(source != "aNICS") %>% readr::write_csv(here::here("data", "measles_vax-coverage-data-cleaned_noadults.csv"))
```