---
title: "measles_prelim-vax-coverage-data-processing"
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| message: false

library(dplyr)
library(tidyr)
library(ggplot2); theme_set(theme_bw())
library(zoo)
library(stringr)
library(readr)
library(forcats)

# Source functions
invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))

# Define functions

#' Add full sequence of age groups, adding new limits
#' 
#' @param df a data frame where ages are stored in the `age_current` column
#' @param new_ages new ages to add before completing the sequence
#' @return a data frame
expand_age <- function(df, new_ages){
  df |>
  # attach new ages
  group_modify(\(x, ...) add_row(x, age_current = new_ages)) |>
  # fill in all missing ages from expanded range
  complete(age_current = full_seq(age_current, period = 1))
}

#' Parse age groups into one observation per single-year age
#'
#' @param df data frame
#' @param col column to transform, as a symbol
parse_age_groups <- function(df, col, last_age = 99){
  df |>
    rowwise() |> 
    mutate(
      {{col}} := list(pull_first({{col}}):pull_last({{col}}, last_age))
    ) |>
    unnest_longer({{col}})
}
```

We've collected a large dataset of vaccine coverage data (see `measles_vax-coverage-data.qmd`). Here we carry out preliminary data processing tasks, largely following the 'take 2' approach outlined in `measles_prelim-vax-coverage-data.qmd` but with some changes outlined below (\*).

## Read data

```{r}
#| label: read_data

# read in data and filter to province-level
PT_lookup <- readr::read_csv(here::here("data", "pt_lookup.csv"), show_col_types = FALSE)
popsize <- read_csv(here::here("data", "popsize.csv"), show_col_types = FALSE)
vax_raw <- readr::read_csv(here::here("data", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(PT_lookup$nm_en))
vax_schedule <- readr::read_csv(here::here("data", "measles_raw", "vax_schedule.csv"), show_col_types = FALSE)

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates
age_pre1970 <- current_year - 1969 # min age of those born strictly before 1970
```

## Split off 18+ estimates

```{r}
#| label: spilt_data

vax_split <- vax_raw |>
  mutate(aNICS = (age == "18+" & !is.na(age)))

vax_raw <- vax_split |> filter(!aNICS) |> select(-aNICS)
vax_aNICS <- vax_split |> 
  filter(aNICS) |> 
  select(-aNICS) |>
  # prepare for attaching later
  # age estimates
  mutate(age_current = pull_first(age) + (current_year - year_report)) |> transmute(
    pt, n_doses, 
    age_current, value,
  ) |>
  # complete age group by pt and n_doses and repeat only obs for all current ages (don't bother estimating for those born 1970 or later---we handle their vax coverage separately)
  group_by(pt, n_doses) |> 
  expand_age(new_ages = c(age_pre1970-1)) |>
  mutate(
    source = "aNICS",
    value = na.approx(value, rule = 2) 
  )

vax_aNICS_full <- vax_split |> filter(aNICS) |> select(-aNICS) |> group_by(pt, location, year_report) |>
  mutate(age_current = pull_first(age) + (current_year - year_report)) |> 
  group_by(pt, location, year_report, age, n_doses) |> expand_age(new_ages = c(age_pre1970-1)) |>
  mutate(value = na.approx(value, rule = 2)) |> mutate(year_birth = current_year - age_current) |>
  mutate(age = year_report - year_birth) |> transmute(pt, location, year_report, year_birth, age, n_doses, value, age_current)
```

## Parse ages and doses

```{r}
#| label: parse_ages_and_doses

vax_parsed <- vax_raw |>
  # parse age groups into single-year ages
  mutate(age = if_else(is.na(age), as.character(year_report-year_birth), age)) |> # fill in age at report where not present
  parse_age_groups(age)
```

## Round ages for data from non-integer ages and keep (e.g. QC) (\*)

```{r}
#| label: non_integer_ages

vax_parsed2 <- vax_parsed |> mutate(age = round(age))
```

## Age the estimates to current year

```{r}
#| label: age_estimates

vax_aged <- vax_parsed2 |> 
  mutate(age_current = age + (current_year - year_report)) |>
  # filter out ages over 99 (ABM max)
  filter(age_current < 100)
```

## Set up "coverage" functions for infants and those born before 1970

```{r}
#| label: attach_immunity_fns

#' Attach immunity assumption for those born before 1970
#' 
#' @param df data frame matching format expected for input into ABM
#' @param age_pre1970 minimum age of those born in 1970 and before
attach_immunity_pre1970 <- function(df, age_pre1970){
  # add coverage for those born in 1970 and before
  bind_rows(
    df, 
    tibble(
      day = 1,
      prov = PT_lookup$al_code,
      age_min = age_pre1970,
      age_max = 99,
      zone = "",
      vstatus = "None",
      percent = 1,
      label = "1-dose-pre-1970"
    )
  )
}

#' Attach maternal immunity assumption for infants
#' 
#' @param df data frame matching format expected for input into ABM
attach_immunity_maternal <- function(df){
  # add coverage infants under 1 year old
  bind_rows(
    df,
    tibble(
      day = 1,
      prov = PT_lookup$al_code,
      age_min = 0,
      age_max = 0,
      zone = "",
      vstatus = "None",
      percent = 0.25,
      label = "maternal-immunity"
      )
    )
}

#' Adjust vaccine coverage estimate for zone X by assuming an unvaccinated subgroup in the PT
#' 
#' @param df dataframe of vaccine coverage (`value`) by population centre (`zone`) and province (`pt`) 
#' @param popsize dataframe of population sizes (read from `data/popsize.csv`)
#' @param pt PTs to adjust
#' @param prop_unvaccinated Proportion of the PT population that is unvaccinated
adjust_X_by_unvaccinated_subgroup <- function(df, popsize, pt, prop_unvaccinated){
  adj <- popsize |>
    filter(pt %in% {{pt}}) |>
    mutate(
      prop_adj = (prop_pop_rural-prop_unvaccinated)/prop_pop_rural
    ) |>
    select(pt, prop_adj)
    
  
  df |>
    left_join(adj, by = join_by(pt)) |>
    mutate(value = if_else(zone == "X" & !is.na(prop_adj), value*prop_adj, value)) |>
    select(-prop_adj)
}
```

## Collapse exactly 1 and exactly 2 dose types to 1+

For data with exactly 1 dose and exactly 2 dose, Ontario's is the only data in which the sum of the coverage estimates across dose types (0, 1, 2) is 1, *i.e.* the estimates are consistent and therefore valid. (See `measles_prelim-vax-coverage-modelling.qmd` take 2 approach for full details).

Thus, for ON we will sum 1 and 2 dose coverage as reported to get a true 1+ estimate.

NB: Note that the Ontario data came from two sources, one with 0, 1, and 2 doses, and another with estimates for 2 doses only. For consistency, we only use the multiple doses data.

```{r}
#| label: take2_filter_ON

vax_ON <- vax_aged %>%
  filter(
    pt == "ON",
    n_doses %in% c("1","2+") # keep 1 and 2+ doses (compatible estimates)
  ) %>%
  #sum the uptake for each year to give estimates for 1+
  group_by(pt, location, year_report, year_birth, age, age_current) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  mutate(n_doses = "1+")  # (*)

vax_new <- vax_aged %>%
  filter(pt != "ON") %>%
  bind_rows(vax_ON)
```

And then relabel all estimates of 1/1+ to "1+" (because after dealing with Ontario, all 1 and 1+ data does truly represent "at least one dose" now). (\*)

```{r}
#| label: correct_1_and_1+_labels

vax_corrected_labs <- vax_new %>% mutate(n_doses = if_else(n_doses == "1", "1+", n_doses))
```

And relabel all 2/2+ to "2+" as these data all now represent "two or more doses".

```{r}
#| label: correct_2_and_2+_labels

vax_corrected_labs2 <- vax_corrected_labs %>% mutate(n_doses = if_else(n_doses == "2", "2+", n_doses))
```

## De-duplicate

Take the most recent estimates by current age *and dose*, where there are duplicates. (\*)

```{r}
#| label: take2_dedupe

vax_dedupe <- vax_corrected_labs2 %>%
  # choose the most recent estimate (by current age and dose type) using year reported
  group_by(pt, location, age_current, n_doses) %>%
  filter(year_report == max(year_report)) 
```

Some provinces have two estimates for the same year, for different dosages. We do not need to remove these - this could be useful information to pull from and/or for validation (\*). Note the GP model currently filters out all 2/2+ estimates anyway.

## Fill in year_birth, reattach aNICS estimates and prioritise provincial data if there are duplicates

Fill in year_birth where it is missing, reattach adults estimates, and prioritise provincial data over aNICS where there are duplicates (because this only happens in Alberta current age 21+, where the provincial data is fairly recent (reporting age 17) and expected to have better sampling? than aNICS). (\*)

```{r}
#| label: year_birth_and_aNICS

vax_clean_noadults <- vax_dedupe %>% ungroup() %>% 
  mutate(year_birth = ifelse(is.na(year_birth) & !is.na(year_report) & !is.na(age), year_report - age, year_birth))

vax_clean <- vax_clean_noadults %>% mutate(priority = 1L) %>%
  bind_rows(vax_aNICS_full %>% filter(age == 18) %>% mutate(priority = 2L)) %>%  # attach one adult estimate from aNICS for each province (age 18)
  group_by(pt, location, age_current, n_doses) %>% filter(priority == min(priority)) %>% # prioirtise provincial data when there are duplicates between provincial data and aNICS
  ungroup() %>% select(-priority)
```

## Plot and export

Here's what the data looks like:

```{r}
#| label: plot

vax_clean |> #filter(n_doses == "1+") |>
  ggplot(aes(x = age_current, y = value)) +
  geom_point(aes(shape = n_doses)) + 
  facet_wrap(~location) +
  xlim(1, 30) +
  ylim(0, 1)
```

The changes made in this file compared to the preliminary modelling file (all marked with an \*) have recovered 6 more data points in BC, 5 more data points in AB, and 8 more data points in QC. These all previously got removed when taking the most recent estimates regardless of dose type (as in done in `measles_prelim-vax-coverage-modelling.qmd`). We can now export vax_clean for use in `gp_model.qmd`:

```{r}
#| label: export_vax_clean

vax_clean %>% readr::write_csv(here::here("data", "measles_vax-coverage-data-cleaned.csv"))
vax_clean_noadults %>% readr::write_csv(here::here("data", "measles_vax-coverage-data-cleaned_noadults.csv"))
```
