---
title: "A closer look at CNICS/ANICS"
author: "Irena Papst"
editor: visual
---

```{r}
#| label: setup

library(readr)
library(dplyr)
library(tidyr)
library(skimr)

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))
```

I chatted with Ashleigh Tuite yesterday about our project and she pointed out that a major challenge with our present data set is that the data is not all coming from the same methodology... the survey products (CNICS/ANICS) will at least have consistent methodology between PTs for a given survey, and likely consistency from survey to survey.

I'd like to take a closer look at what's really publicly available for CNICS/ANICS in terms of downloadable data. It may not be the case that we want to switch to using it as a basis for our project, but perhaps it could be useful for setting good priors when fitting models to the vaccine coverage data?

## Looking for the available data

### ANICS

Valerie pointed me to a [methodological report](https://epe.bac-lac.gc.ca/100/200/301/pwgsc-tpsgc/por-ef/public_health_agency_canada/2023/140-22-e/index.html) for ANICS 2023 that includes a csv download:

```{r}
data_dir <- file.path("data", "CNICS-ANICS_raw")
# here::here(data_dir, "2023_ANICS_POR140-22-Detailed-Tables-all-tables-e.csv")
```

This `.csv` isn't structured in the most machine-readible way, but we can make it work if needed. Searching through the doc for measles, I see the following relevant tables:

- Table 7a. Routine vaccine coverage by vaccine, overall, by sex, by age group, among adults with chronic health conditions, and among health care workers, Canada, 2023"

- Table 7b. Routine vaccine coverage by vaccine, by province/territory, Canada, 2023

- Table 10. Respondents' intention to get vaccinated¹, overall, by sex, by age group, among people with chronic health conditions, and among health care workers, Canada, 2023

  - ¹Among non-immunized individuals and those who responded "Don't know" or "Prefer not to answer" when asked about their vaccination status

Unfortunately, there is no view of the responses with vaccine coverage by age group **and** PT. However, the age groups are more granular than 18+ estimates reported in top-level summaries...

### CNICS

Looking searching the data tab on the StatCan website, I was able to find exactly [one table](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310087001) with a search for "Childhood National Immunization Coverage Survey" (and its metadata). The data are in:

```{r}

filename <- here::here(data_dir, "13100870.csv")
```

(The metadata has a similar filename.)

Some key info from the metadata file:

- The Childhood National Immunization Coverage Survey (CNICS) has been conducted nationally every two years since 2011 and by province/territory every four years since 2013.

- All participants were asked to provide information on their children's immunization through the child's vaccination record and were also asked to complete and return by mail a form that provides consent for Statistics Canada to contact their child's health care provider to obtain their child's vaccine history.

  - Starting in 2019, respondents from Prince Edward Island and Manitoba (the only jurisdictions with current registry data sharing agreements) were asked for consent to use their child immunization data if present in provincial registries.

Thankfully, this dataset is somewhat machine readible! It looks like this dataset might contain all the CNICS data (since 2021)? Let's filter down to just measles...

```{r}
#| results: "asis"

df <- read_csv(filename) |> janitor::clean_names() |> filter(antigen_or_vaccine=="Measles")
names(df)

df  |> skim() 
```

Lots of stuff here, not all of which is needed... let's keep the key columns that we need for our standardized format and drop what we don't need

```{r}
#| results: "asis"

df_filt <- df |> 
  select(ref_date, geo, target_population, characteristics, value) |>
  drop_na(value)

df_filt |> skim() 
```

For now, we will keep just the value reported for percentage vaccinated (some kind of central tendency), though it's worth noting there's uncertainty bounds available for each estimate...

```{r}
unique(df_filt$target_population)

df_std <- df_filt |> 
  # select only mean estimates
  filter(characteristics=="Percentage vaccinated") |>
  # finish reformating into standard format
  rename(year_report = ref_date, location = geo) |>
  mutate(
    age = as.numeric(stringr::str_extract(target_population, "[[:digit:]]+")),
    year_birth = year_report - age,
    n_doses = "1+",
    value = value/100 # as a proportion
  ) |>
  full_join(pt_lookup(), by = join_by(location)) |>
  select(pt, location, year_report, year_birth, age, n_doses, value)

DT::datatable(df_std)
```

```{r}
# estimates as is
plot_coverage(df_std)

# naive ageing of estimates
plot_coverage(df_std |> mutate(age_current = 2025 - year_birth), x = "age_current")
```

Did we filter down sufficiently? Are there any unexpected duplicates?

```{r}
df_std |>
  group_by(pt, location, year_report, year_birth, n_doses) |>
  count() |>
  filter(n > 1)
```

Nope, looks good! Let's save this tidied dataset for further use:

```{r}
#| eval: false

write_csv(df_std, here::here("data", "measles_CNICS.csv"))
```
