---
title: "Investigating potential bias between survey and PT-reported data"
format: html
---

Here we briefly investigate whether there may be evidence of bias in the vaccine coverage estimates stemming from survey data (cNICS, aNICS) compared to the PT-reported data.

```{r}
#| label: setup

library(dplyr)
library(ggplot2); theme_set(theme_bw())

this_theme <- theme_bw(base_size = 11) +
  theme(
    strip.text.x     = element_text(size = 8),
    axis.title       = element_text(size = 12),
    axis.text        = element_text(size = 10),
    legend.title     = element_text(size = 8),
    legend.text      = element_text(size = 8),
    plot.title       = element_text(size = 16),
    panel.grid.minor = element_blank()
  )
theme_set(this_theme)

invisible(lapply(list.files(here::here("R"), full.names = TRUE), source))
```

The following data processing was copied over from `2_prepare-data.qmd` (everything up to de-duplication):

```{r}
#| label: read_data

# read in data and filter to province-level
PT_lookup <- readr::read_csv(here::here("data", "pt-relation", "pt_lookup.csv"), show_col_types = FALSE)
vax_raw <- readr::read_csv(here::here("data", "coverage", "generated", "measles_vax-coverage-data.csv"), show_col_types = FALSE) |> filter(location %in% unique(PT_lookup$nm_en))

current_year <- as.integer(format(Sys.Date(), "%Y")) # for ageing estimates

# Remove low Nunavut point
vax_raw <- vax_raw |> filter(!(pt == "NU" & year_report == 2021))

## Parse ages and doses
vax_parsed <- vax_raw |>
  # parse aNICS "18+" as 18
  mutate(age = if_else(source == "aNICS", "18", age)) |>
  # parse age groups into single-year ages
  mutate(age = if_else(is.na(age), as.character(year_report-year_birth), age)) |> # fill in age at report where not present
  parse_age_groups(age) |>
  # round non-integer ages (QC)
  mutate(age = round(age))

## Age the estimates to current year
vax_aged <- vax_parsed |> 
  mutate(age_current = age + (current_year - year_report))

## Collapse exactly 1 and exactly 2 dose types to 1+

# For data with exactly 1 dose and exactly 2 dose, Ontario's is the only data in which the sum of the coverage estimates across dose types (0, 1, 2) is 1, *i.e.* the estimates are consistent and therefore valid. (See `measles_prelim-vax-coverage-modelling.qmd` take 2 approach for full details).

# Thus, for ON we will sum 1 and 2 dose coverage as reported to get a true 1+ estimate.

# NB: Note that the Ontario data came from two sources, one with 0, 1, and 2 doses, and another with estimates for 2 doses only. For consistency, we only use the multiple doses data.

vax_ON <- vax_aged %>%
  filter(
    pt == "ON",
    source == "PT",
    n_doses %in% c("1","2+") # keep 1 and 2+ doses (compatible estimates)
  ) %>%
  # sum the uptake for each year to give estimates for 1+
  group_by(pt, location, year_report, year_birth, age, age_current, source) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  mutate(n_doses = "1+")  # (*)

# replace only the ON PT data with the version above
vax_new <- vax_aged %>%
  filter(!(pt == "ON" & source == "PT")) %>%
  bind_rows(vax_ON)

# And then relabel all estimates of 1/1+ to "1+" (because after dealing with Ontario, all 1 and 1+ data does truly represent "at least one dose" now). Also relabel all 2/2+ to "2+" as these data all now represent "two or more doses".

vax_corrected_labs <- vax_new %>% 
  mutate(n_doses = if_else(n_doses == "1", "1+", n_doses)) %>% 
  mutate(n_doses = if_else(n_doses == "2", "2+", n_doses))
```

Find pts and ages where coverage for "1+" doses exists in both PT-reported and survey (cNICS or aNICS) data: 

```{r}
#| label: vax_matched
vax_matched <- vax_corrected_labs |>
    filter(n_doses == "1+", age_current <= 21) |>
    # get rid of older duplicates for AB
    filter(
        !(pt == "AB" & source == "PT" & age_current >= 20 & age == 13)
    ) |>
    # compress cNICS and aNICS sources to "survey"
    mutate(source = if_else(source %in% c("cNICS", "aNICS"), "survey", source)) |>
    # find matched measures
    select(pt, age_current, source, value) |>
    tidyr::pivot_wider(id_cols = c(pt, age_current), names_from = source, values_from = value, names_prefix = "value_") |>
    tidyr::drop_na()
```

Calculate differences and estimate mean bias with a confidence interval:

```{r}
bias <- vax_matched |> 
    mutate(diff = value_survey - value_PT, mean = (value_survey + value_PT)/2)

bias_mean <- mean(bias$diff)
bias_se <- sd(bias$diff) / sqrt(nrow(bias))
bias_mean_ci <- bias_mean + c(-1, 1) * qt(0.975, df = nrow(bias) - 1) * bias_se

glue::glue("Mean bias (survey - PT, coverage as a percentage): {round(bias_mean*100, 1)}%\n95% CI: ({round(bias_mean_ci[1]*100, 1)}%, {round(bias_mean_ci[2]*100, 1)}%)")
```

Make Bland-Altman plot:

```{r}
ggplot(bias, aes(x = mean, y = diff, colour = diff)) +
    geom_point(size = 4) +
    geom_hline(yintercept = bias_mean, linetype = "dashed") +
    geom_hline(yintercept = bias_mean_ci, linetype = "dotted") +
    annotate("text", x = max(bias$mean), y = bias_mean,
             label = "Mean", hjust = 1, vjust = -0.5, size = 3) +
    annotate("text", x = max(bias$mean), y = bias_mean_ci[1],
             label = "95% CI", hjust = 1, vjust = 1.5, size = 3) +
    annotate("text", x = max(bias$mean), y = bias_mean_ci[2],
             label = "95% CI", hjust = 1, vjust = -0.5, size = 3) + 
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) + 
    scale_colour_gradient2(mid = "grey70") +
    labs(
        x = "Mean coverage (%) between survey and PT sources", 
        y = "Absolute difference in coverage (%) \n(survey - PT)",
        colour = stringr::str_wrap("Absolute difference in coverage (%), survey - PT", width = 15),
        title = stringr::str_wrap("Bland-Altman plot of coverage estimates demonstrates low evidence of bias in survey responses", width = 55)
    ) 
```